---
title: "Exercise 3: Factor Analysis to Find Underlying Factors"
output:
    html_document: default
    word_document: default
    pdf_document: default 
editor_options: 
  markdown: 
    wrap: 72
---

**Course Instructor**: Dr. Suborna Ahmed

**Student Name**: Sandra Inusah, Brandon Weigand

**Total Points**: 10

**Due**: March 6, 2026

**Submission**:

1.  Write your code and answers in this RMD file
2.  Knit your completed RMD file to WORD document
3.  Check and upload your WORD document to Exercise 2 on Canvas

------------------------------------------------------------------------

### Objectives

1.  To become familiar with Factor Analysis as a tool used to uncover
    underlying relationships in a set of x-variables.

2.  To learn how to use different rotations as a means of achieving the
    simple structure needed in order to be able to interpret the factors
    (and the factor scores) and assign meaningful names to each of them.

### General Description

Information was gathered by Tracy Hooper on bird abundancies, along with
vegetation and other variables in a grassland area (a subset of her data
is used in FRST 531 with her approval). Using factor analysis, the
underlying relationships among **11 of the vegetation variables** was
analyzed. The 11 variables are:

-   `mayvgcov` and `mayvght` (% cover and height of vegetation in May)

-   `junvgcov and`junvght\` (% cover and height of vegetation in June);

-   and horizontal cover variables, specifically: `baregrnd`, `bryopht`,
    `forbs`, `grass`, `rocks`, `shrubs`, and `trees` all as
    **percents**.

The overall objective is to use factor analysis to **obtain a few
indices to represent underlying factors, and to be able to give these
underlying factors names**. These factors could give insights into the
data (e.g., which plots gave high scores on Factor 1 versus low scores,
and what does this mean ecologically?). Also, these factors could be
used in a subsequent analysis to replace these 11 variables.

However, for the factors to have meaning and to be able to name the
factors, a simple structure is needed. For example, if Factor 1 has high
positive correlations with `mayvgcov` and `junvgcov` perhaps this factor
is “summer vegetation cover”. However, if Factor 1 also has a moderately
high positive correlation with `rocks`, what does this mean? Instead
another rotation may provide a more interpretable and simple structure.

For this, you will first do the PCA. Then, you will use factor analysis
first with no rotation. Then, you will use orthogonal and oblique
rotations to get a simple solution. Remember that oblique rotations are
much harder to interpret since the factors are correlated (unless the
correlation is very low?). Our preference is to find a simple structure
using orthogonal rotations if possible.

### Process

1.  For this analysis, the data have been provided as `TRABREV2.txt`
    along with R codes and questions as in this Rmd file.

2.  You will need to run the R code and get the outputs from running the
    code. There are two things you will need to change:

-   Change the working directory to the location of the data on your
    computer, if you have not done so already.

-   Add in code for the 4th factor analysis run (object will be called
    `veg.fac4`) by cutting and pasting code for the other factor
    analysis runs and then modifying this for an equamax rotation. There
    are three other factor analysis runs before this one (i.e.,
    `veg.fac1` to `veg.fac3`) and one after (i.e., `veg.fac5`) that are
    already in the code and there is a location where to put the 4th run
    for `veg.fac4`.

3.  The outputs will include:

-   The PCA on the 11 variables;
-   The unrotated factor solution, using two factors (`veg.fac1`)
-   Several rotated factor solutions (`veg.fac2`, `veg.fac3`, `veg.fac4`
    (you will put the code in), and `veg.fac5` (the only oblique
    rotation).

4.  You will be answering a **8 questions** to guide you through the
    outputs and interpret the results. The questions are marked with
    **TODO** to help you quickly locate them using the search tool,
    since this Rmd file is quite long. As you answer the questions,
    paste in the relevant R outputs (and/or graphs) to support your
    answers.

------------------------------------------------------------------------

### R CODES

#### Introduction

Data from an outside text file from birds in grasslands study by Tracy
Hooper (subset of her data used by permission) is brought in to R from
an outside space-delimited text file. Alternatively, you could bring the
text file into EXCEL and save it as a `.csv` file. Using only a subset
of the variables, the goal is to get a simple structure on a few factors
in order to label and interpret these factors. These factors would then
commonly be passed to another analysis. We would prefer to obtain a
simple structure using orthogonal rotations, but sometimes this is not
possible. The R code starts with PCA first to indicate the contribution
of each principal component. Factor analysis (FA) then starts with two
factors and no rotation, which is like the PCA solution as presented in
the Factor analysis notes. Then, further rotations are used to try to
get the simple structure, starting with orthogonal and then oblique
rotations.

#### PART 1: Clean out all R objects before starting.

Then set the directory for the data (and any files saved) and read the
data. We assign it to an R object called `vegdat`. Do a few checks on
the data.

```{r}
rm(list = ls(all = TRUE))
```

Read the data from `TRABREV3.txt` into `vegdat`, set `header=T` to
include the original column (header) names

```{r}
vegdat <- read.table("TRABREV3.txt", header = T)
```

We will check the dimension of the table. We should get *145* rows and
*27* columns.

```{r}
dim(vegdat) # 145 plots (one per row) and 27 variables.
#yes, we get 145 rows and 27 variables. 
```

Here we are just exploring a bit more about this data object

```{r}
# class(vegdat) # what type of object is it?
# str(vegdat)  # what is in the object?
names(vegdat)  # print out the column (i.e., variable) names
#we have things like site, point, rank, vesp, star, etc. 

head(vegdat, 5)  # print out the first few lines of data
```

Note: I'm not sure what hola, vesp, or star are in this context (bird
abundance?), but I get the idea.

Remembering that:

-   `mayvgcov` and `mayvght` (% cover and height of vegetation in May)

-   `junvgcov and`junvght\` (% cover and height of vegetation in June);

-   and horizontal cover variables, specifically: `baregrnd`, `bryopht`,
    `forbs`, `grass`, `rocks`, `shrubs`, and `trees` all as
    **percents**.

#### PART 2: Subset the variables for the FA into a different object

Variables that will make up `xdata` are: `mayvgcov`,`mayvght`,
`junvgcov`,`junvght`,`baregrnd`,`bryopht`,`forbs`, `grass`, `rocks`,
`shrubs`, and `trees`.

```{r}
xdata <- subset(vegdat,select = c(mayvgcov, mayvght, junvgcov, junvght,
     baregrnd, bryopht, forbs, grass,
     rocks, shrubs, trees))
```

Some simple checking of `xdata`. We should still get *145* rows, but
this time only *11* columns.

```{r}
names(xdata)
class(xdata)
dim(xdata) # 145 rows and 11 columns
```

Note: Yep, 145 rows and 11 columns.

#### PART 3: Basic stats and corelation matrix for the xdata

We will get basic stats using `stat.desc()` from the package `pastecs.`
Make sure that you have downloaded the package. You can install it with
`install.packages("pastecs")`. Uncomment the chunk below and run to see
the summary of key statistics.

```{r}
#install.packages("pastecs")

#library(pastecs) # won't work if you have not installed the package first!!
#round(stat.desc(xdata), 2) # get some basic descriptive stats.
summary(xdata) 
```

Here is the correlation matrix rounded to the 2nd decimal place.

```{r}
round(cor(xdata), 2) # correlation matrix with elements rounded to 2 decimals
```

Notes: 
- These correlation matrices reveal big patterns with vegetation data in particular. Vegetation cover and May height (0.92) and June height (0.88) are strongly positively correlated. That is, May cover and height (0.92), June cover and height (0.88). 
- Probably hints at vegetation structure over time, given the other variables like bare ground. Productivity? Some sites seem to do better. 
- Bare ground, meanwhile, is moderately negatively correlated with vegetation and height (May -0.38, June -0.44), and maybe unsurprisingly negatively correlatd with trees (-0.62). Also supported by the grass and ground (-0.82). Hints that 1) bare ground clearly does not contain much vegetation and 2) sites with higher vegetation have less soil. 
- In the same vein, rock cover and bare ground (0.45) (indicating some rock cover presence with bare ground sites), and negative correlations with veg cover and height (-0.29). So, sites that are rocky do not have as much vegetation. 
- Looks like tree cover and vegetation cover have a correlation, but I'm not sure how to interpret this. Sites with trees have underlying ground cover? 
- Bryophytes, forbs, and shrubs have weak relationships with many variables. Bryophytes have correlations of −0.21 with May vegetation cover, −0.16 with May vegetation height, −0.17 with June vegetation cover, and −0.18 with June vegetation height. Bryophytes also show almost no relationship. 
- Everything considered, it looks like a spectrum of vegetation types and their associations or presence. 


#### PART 4: Perform PCA using `princomp()`

First, we do PCA to look at how many factors to keep to explain the
majority of the variances of X variables. This is not strictly needed
for Factor Analysis since you may set the number of factors based on
what might be possible to interpret but still useful to justify the
number of factors to be used.

**NOTE:** `cor` and `covar` in princomp( ) are not the same as
`principal()` or `fa()` !!!

Here, `cor=TRUE` means using the correlation matrix (aka standardized X
variables).

```{r}
veg.pca <- princomp(xdata, cor = TRUE)
```

Let's get a summary with `summary()`

```{r}
summary(veg.pca, loadings = TRUE, cutoff = 0.0)
```

Interpretation of the `summary()` outputs:

-   The first part shows the standard deviation (eigenvalues), how much
    each individual PC contribute to the variance, and the cumulative
    variance when you retain 1, 2, 3, ... principal components

-   The second part shows the "loadings". This is in fact the
    **eigenvectors**. However, when we work with factor analysis later
    on with the function `principal()`, this will not be the case.

#### 

Notes: 
- The first section are the SDs and variance, and the second are the loadings (eigenvectors), or the direction of each principal component in relation to the original variables. 
- PC1 has a SD of 2.08 and explains 39.8% of total variance. 
- PC2 explains 16.1%. 
- PC3 explains 11.9%.
- PC1 - 3 explains 67.3%.
- PC1 loadings connect with May veg cover (0.411), May veg height (0.409), June cover (0.417), and June height (0.367). Grass and trees have smaller loadings. Bare ground and rocks load negatively (-0.354 and -0.251 respectively). 
- PC1 looks to reinforce the cover spectrum idea, going from high vegetation to bare ground and rocks.  
- PC2 looks more complicated. Bare ground and shrubs load positively, yet grass loads moderately negative (-0.531). Not sure how to interpret this. Lower vegetation near the ground? 
- PC3 has moderate positive loadings with bryophytes (0.554). Negative loadings with trees and forbs. So this seems to be a moss-based component, though moss and trees can coexist.   

##### 

Get the correlation matrix for X variables and principal components. We
do not need to scale the X variables since the correlation matrix
already does that. Notice that the `loadings` obtain here is very
different from the so called "loadings" above from
`summary(veg.pca, loadings=TRUE,cutoff=0.0)`.

```{r}
loadings <- round(cor(xdata, veg.pca$scores), 7)
loadings
```

##------

Let's calculate the variances, percent variances, and cumulative percent
variances and put these together using bind. round() these using d=2
(i.e., two decimal places)

```{r}
round(cbind("Variance" = veg.pca$sdev^2,
 "%" = 100*veg.pca$sdev^2 / sum(veg.pca$sdev^2),
  "Cumulative %" = 100 * cumsum(veg.pca$sdev^2) / sum(veg.pca$sdev^2)),
   d = 2)
```

Check the sum of the variances to make sure they sum up to **11**, which
is the number of X variables.

```{r}
sum(veg.pca$sdev^2) # check the sum of variances.
```

Get a scree plot with `sdev^2` (that is the variance or eigenvalue) of
each component.

```{r}
plot(veg.pca$sdev^2, type = "b", pch = 16, xlab = "Component number",
ylab = "Eigenvalue", main = "Scree plot for x-variables of veg data")
```

Notes: 
- PC1 has a variance of 4.33 and accounts for 39.33% of total variance. 
- PC2 explains 16.08%.
- PC3 explains 11.91%. 
- PC1 - 3 accounts for 67.31 of variation. 
- The scree plot shows a steep "elbow" drop between the first and second components, followed by a gradual decline. Clearly the first component captures most of the variance. 

## ------

Graph PC 1 and 2 (i.e., dimensions 1 and 2) with a biplot.

```{r}
biplot(veg.pca)
```

**OPTIONAL:** These are alternative graphics

```{r, eval=FALSE}
plot.new() # start a new plot
plot(loadings[, 1], loadings[, 2], axes = FALSE,
xlim = c(-1, 1), ylim = c(-1, 1), xlab = "", ylab = "")
mtext(expression("Comp.2"), 1, las = 0) # put this on axis 1 (x-axis)
mtext(expression("Comp.1"), 2, las = 0) # put this on axis 2 (y-axis)
axis(1, at = c(-1, 1), pos = 0) # axes labels at 0, from -1 to 1
axis(2, at = c(-1, 1), pos = 0) # axes labels at 0, from -1 to 1
# labels for the x variables to each correlation value, pos=4 is to the right.
text(loadings[, 1], loadings[, 2], labels = colnames(xdata), pos = 4, cex = 0.7)
```

##----

Notes: biplots visualize via arrow direction and length how strongly each variable influences the positions of sites in the study. Those on the right tended to have higher veg cover, height, and grass. Left arrows indicate the sites had less veg cover and more bare ground and/or rocks. This plot is messy but the second one shows that may and june veg cover, trees and grass load to the right and bare ground and rocks load to the left. I suppose this could be interpreted to mean that vegetation structure (presence?) is less loaded with ground and rocks and bryophytes. So from right to left, a gradient of densely vegetated sites to less vegetated ones. 

##----

#### PART 5: Factor analysis using two factors based on PCA analysis.

##### General information:

The function `factanal()` can be used. Options for `rotation=` parameter
are `"varimax"`, `"promax"`, and `"none"`.

The function principal() can be used.

Another alternative is the `psych` package and the function `fa()`.
There are also options for getting the factor scores. `principal()` is
used here from the `psych` package factor scores will be calculated
using the default "regression" method.

To use the variety of rotations in `fa()` or in `principal()`, you also
need the package `GPArotation` Rotations options:

-   `"none"`, `"varimax"`, `"quartimax"`, `"bentlerT"`, `"equamax"`,
    `"varimin"`, `"geominT"` and `"bifactor"` are orthogonal rotations.

-   `"promax"`, `"oblimin"`, `"simplimax"`,`"bentlerQ"`, `"geominQ"` and
    `"biquartimin"` and `"cluster"` are possible oblique transformations
    of the solution.

**NOTE:** `cor` and `covar` in `princomp()` are not the same as in
`principal()` or `fa()`!!!

In `principal()` and in `fa()`, `cor="cor"` means **Pearson's
correlations** but then `covar=FALSE` means using the correlation
matrix!!

For `veg.fac1`, no rotation used. For the following codes to work,
please make sure you have downloaded the `psych` and `GPArotation`
packagesvia `install.packages("GPArotation")` and
`install.packages("psych")`.

```{r}

#install.packages("GPArotation")
#install.packages("psych")

library(GPArotation)
library(psych)

veg.fac1 <- principal(xdata, nfactor = 2, rotate = "none",
      method = "regression", scores = TRUE, covar = FALSE, cor = "cor")
```

**NOTE:** `covar=FALSE` meaning using the correlation matrix.
`cor="cor"` means using **Pearson's correlations** (there are other
options!)

We can explore this object and its components via the functions
`class()` and `str()`. You can uncomment the codes to explore it by
yourself by running this code chunk. We will not run it here at the
meantime because the outputs are quite lengthy.

```{r}
# uncomment and run to see ouputs

class(veg.fac1)   # what type of object is created?
str(veg.fac1)     # what do we get?
```

```{r}
veg.fac1
```

##### Let's go over some important components of this object:

```{r}
veg.fac1$loadings
```

`loadings` are simple **correlations between PC scores and x
variables**, which is **NOT like using** `princomp()`! **Note:** very
low correlations are not given and will appear missing.

```{r}
veg.fac1$weights
```

`weights` is the W (multipliers) to calculate factor scores

Next, we explore `communality` and `uniqueness`

```{r}
veg.fac1$communality # the h2 in the table
```

```{r}
veg.fac1$uniqueness  # the u2 in the table
```

We learned that the variance of each X variable is divided into **common
variance (communality)** and **unique variance (uniqueness)**

-   `communality` is the **common variance** of the X variables. It is
    the variance *accounted for by the Factors*. In other words,
    `communality` tells you what proportion of the variable's variance
    is a result of either:
    -   The principal components
    -   The correlations between each variable and individual factors

It is calculated by squaring each factor loading and then adding them up
for each X variable (sum of squared factor loadings)

-   `uniqueness` is the **unique variance** of the X variables. It is,
    as its name implied, the variance *not accounted for by the
    Factors*, thus are *unique* to each X variable.

It is *1 - communality* since the correlation matrix was used. In other
words, it is a measure of how well the regression estimate the factor
scores using the x variables as inputs based on root mean square
residuals. A smaller number means a better fit.

**Note:** `fit` is a measure of how well the factors represent the full
correlation (or covariance matrix) of the X's. A smaller value means a
better fit.

```{r}
veg.fac1$Structure
```

`Structure` is the Factor Structure matrix, which in this example
represents the **simple correlations** between each X variable and each
Factor. If the factors are **uncorrelated (orthogonal)** and
**standardized**, or if is is an **orthogonal rotations of the axes**,
the Factor Structure matrix and the Factor Pattern matrix is the same.

```{r}
veg.fac1$complexity
```

`complexity` is the Hoffman's index of complexity.

**Extra:** it represents the number of latent components needed to
account for the observed variables. Whereas a perfect simple structure
solution has a complexity of 1 in that each item would only load on one
factor, a solution with evenly distributed items has a complexity
greater than 1. [Link to
source](https://easystats.github.io/parameters/reference/model_parameters.principal.html#details)

Here, we calculate the **mean complexity** of the variables.

```{r}
mean(veg.fac1$complexity)
```

#### Correlations between the factors and the X variables

```{r}
round(cor(xdata, veg.fac1$scores), 2)
```

This is same as `veg.fac1$loadings`. Smaller correlations are not shown.

##### Correlations between the two factors

```{r}
round(cor(veg.fac1$scores, veg.fac1$scores), 2)
```

##------

Notes: 

- Factor 1 represents a vegetation structure gradient, with strong positive loadings for May and June vegetation cover and height, and strong negative loadings for bare ground and rocks.
- Factor 2 represents a secondary vegetation composition contrast, with shrubs and bare ground loading positively and grass loading strongly negatively.
- Communalities and uniqueness values were examined to assess how well variables were explained by the two factor model.
- Vegetation structure variables, grass, and bare ground show high communalities, indicating the two factors explain a large portion of their variance.
- Bryophytes and forbs show high uniqueness, suggesting they are poorly captured by the two factor structure and likely represent more localized or independent variation.

##-----

##### "Hand calculate" the scores using the weights coefficients.

Get the standardized xdata first

```{r}
Xscaled <- scale(xdata, center = TRUE, scale = TRUE)
head(Xscaled)
```

Check the dimension of the `weights` matrix

```{r}
dim(veg.fac1$weights)
```

```{r}
myscores <- Xscaled %*% veg.fac1$weights
head(myscores)
```

```{r}
head(veg.fac1$scores)
```

Calculating communality and uniqueness for the PC orthogonal rotation
and for other orthogonal rotations. Using the correlation matrix, the
total variance of the X's is 11 (=1 for each X since this is like you
are using standardized x's) can calculate the variances accounted for by
squaring and adding the loadings.

##### Using just PC1, you get the variance of all X's accounted for by PC1 only.

```{r}
PC1 <- veg.fac1$loadings[1:11, 1] # column 1
PC1

sum(PC1^2)
```

Here, PC1 accounts for 4.32614 of the 11 x-variables variances , which
is its `communality`.

##### Using PC2, you get the variance of all X's accounted by PC2 only

```{r}
PC2 <- veg.fac1$loadings[1:11, 2] # column 2.
PC2

sum(PC2^2)
```

Again, PC2 accounts for 1.7683 of the 11 x-variables variances.

If you add PC1 and PC2 as the only two factors retained, you get the
total communality, all x's and all factors. Total uniqueness not
accounted for by these two components is the number of x-variables minus
total communality. This is because the correlation matrix was used and,
the sum of correlation is 11 (diagonal entries of correlation matrix)

The total communality is:

```{r}
sum(PC1^2) + sum(PC2^2)
```

Thus, the uniqueness is:

```{r}
11 - (sum(PC1^2) + sum(PC2^2))
```

##### Then, how much communality of each X variable versus uniqueness?

```{r}
commonbyX <- PC1^2 + PC2^2
commonbyX
```

You may notice that `commonbyX` is same as `h2` column in the `veg.fac1`
summary of outputs. `h2` tells you how much each x-variable contribute
to the common variance (`communality`). It should sum up to 6.0944,
which is the total communality calculated with the principal components
earlier. Let's check:

```{r}
sum(commonbyX)
```

It's the same number as total communality!

##### How much is unique for each variable?

```{r}
uniquebyX <- 1 - (PC1^2 + PC2^2)
uniquebyX
sum(uniquebyX)
```

Similarly, you may notice that `uniquebyX` is similar to `u2` from the
`veg.fac1` summary of outputs. This corresponds to the uniqueness of the
x-variables. Again, it should sum up to 4.9055, which is the value of
uniqueness calculated with the principal components earlier.

And finally, `commonbyX` and `uniquebyX` should sum up to 11, which is
the number of x variables, similar to the result we have above.

```{r}
sum(commonbyX) + sum(uniquebyX)
```

##------

#### `Varimax` rotation: A further orthogonal rotation

We will called PCs as RC1 and RC2 now. Remember to install the `psych`
package first if you have not done so already. In this example we will
use the **varimax** rotation by setting the parameter
`rotate = "varimax"`

Varimax **maximize variation in squared factor loadings**. Rotate so
that the variation of the squared factor loadings is maximized. Maximum
variation occurs when factor loadings are separated into being either
close to 0 or close to 1 and -1. .

```{r}
library(psych)
veg.fac2 <- principal(xdata, nfactor = 2, rotate = "varimax",
  method = "regression", scores = TRUE, covar = FALSE, cor = "cor")

veg.fac2
```

Let's check the components of `veg.fac2`

`communality` or `h2` in the table:

```{r}
veg.fac2$communality # the h2 in the table
```

`uniqueness` or `u2` in the table:

```{r}
veg.fac2$uniqueness  # the u2 in the table
```

`loadings` are the correlation between the original x variables and RC1
& RC2

```{r}
veg.fac2$loadings    # RC1 and RC2 values. NOTE: Very low correlations are not given.
```

`weights` are the multipliers

```{r}
veg.fac2$weights
```

`Structure` is the same as the `loadings` from the function
`princomp()`, which are the correlations of the x variables and the PCs,
since is an orthogonal rotation.

```{r}
veg.fac2$Structure # same as loadings which ARE correlations for princomp()
```

`complexity` is Hoffman's index of complexity. The mean of complexity is
shown in the summary as well.

```{r}
veg.fac2$complexity  
mean(veg.fac2$complexity)
```

##### Correlations between the factors and the X variables

Let's find the correlations between the X variables and the factor
scores This will be the same as the `veg.fac2$loadings`. You may also
notice that it is similar to `veg.fac2$Structure` as well, due to the
fact that varimax is an **orthogonal rotation**.

```{r}
round(cor(xdata, veg.fac2$scores), 2)
```

##### Correlations between the two factors

The correlations between the two factors should be 0, since they are
from an orthogonal rotation and hence are `uncorrelated`. Let's verify
that fact:

```{r}
round(cor(veg.fac2$scores, veg.fac2$scores), 2)
```

##### "Hand calculate" the scores using the weights coefficients.

First, we get the standardized x variables

```{r}
Xscaled <- scale(xdata, center = TRUE, scale = TRUE)
```

Calculate with matrix multiplication. The results are the same

```{r}
myscores <- Xscaled %*% veg.fac2$weights
head(myscores)
head(veg.fac2$scores)
```

Using just RC1, you get the variance of all X's accounted for by factor1
only.

**NOTE:** This works again since this is an orthogonal rotation.

##### RC1 and RC2 are independent (at right angles)

```{r}
RC1 <- veg.fac2$loadings[1:11, 1]
sum(RC1^2)
```

We can see that RC1 accounts for 3.275 of the variance.

##### Using RC2, you get the variance of all X's accounte by RC2 onlysum(RC2\^2)

```{r}
RC2 <- veg.fac2$loadings[1:11, 2]
sum(RC2^2) # How much does RC2 account for?
```

Similarly, we can see that RC2 account for 2.819 of the variance

##### Total communality and uniqueness.

Still 2 factors so the same as the "none" rotation.

```{r}
sum(RC1^2) + sum(RC2^2) # total communality
11 - (sum(RC1^2) + sum(RC2^2)) # uniqueness
```

##### How much communality of each X variable versus uniqueness?

This is same as `h2` in the `veg.fac2` summary of outputs:

```{r}
commonbyX <- RC1^2 + RC2^2
commonbyX
sum(commonbyX)
```

##### How much is unique for each variable? What is left of the variance?

And this is the same as `u2`:

```{r}
uniquebyX <- 1 - (RC1^2 + RC2^2)
uniquebyX
sum(uniquebyX)
```

Notes:
Further orthogonal rotations can be done with a specific goal for the simple structure, but
again using orthogonal rotations only. The factors will remain uncorrelated as with the
PCA, and the pattern=structure=simple correlations. Varimax maximizes variation in squared factor loadings. Rotate so that the variation of the squared factor loadings is maximized. Maximum variation occurs when factor
loadings are separated into being either close to 0 or close to 1 and -1. In practice, this pushes loadings closer to 0 or ±1, producing a clearer separation of variables among factors. After rotation, the two factors are referred to as RC1 and RC2, but they still represent the same total variance as the unrotated solution
because orthogonal rotations only redistribute variance between factors rather than changing the total explained variance. 

Key numerical results

The rotated solution produces two factors with the following variance
contributions:

RC1 variance: 3.275

RC2 variance: 2.819

Total variance explained: 6.094 of 11 variables

Proportion of total variance explained: 0.554 (55.4%)

Because the rotation is orthogonal, the two factors remain uncorrelated
(correlation = 0), which was confirmed by the factor score correlation
matrix. The communality and uniqueness values also remain identical to
the unrotated solution since rotation redistributes loadings but does
not change how much total variance each variable shares with the factor
model.

Interpretation of the rotated factors:

The first rotated factor (RC1) primarily represents overall vegetation
structure and seasonal vegetation growth. Several vegetation cover and
height variables load strongly on this factor:

junvgcov: 0.89

junvght: 0.84

mayvgcov: 0.81

mayvght: 0.80

These strong positive loadings suggest RC1 reflects a gradient of dense,
tall vegetation during the growing season. Sites with high RC1 scores
likely have greater vegetation coverage and height across both May and
June.

The second rotated factor (RC2) captures a different ecological gradient
related to ground surface composition. Strong loadings include:

grass: 0.93

bare ground: -0.86

shrubs: -0.50

trees: 0.49

This factor contrasts areas dominated by grass cover with areas
characterized by bare ground or woody vegetation. High RC2 values
therefore correspond to grass-dominated surfaces, while negative values
indicate more exposed soil or shrub cover.

Variables poorly explained by the factors

Several variables have low communalities, meaning the two-factor model
does not capture much of their variance:

bryophytes: h² = 0.13

forbs: h² = 0.12

rocks: h² = 0.30

trees: h² = 0.29

These variables remain largely unique or independent of the dominant
vegetation gradients, suggesting they either vary independently of the
main vegetation structure or would require additional factors to be better
represented.

Overall interpretation

The varimax rotation clarifies the ecological meaning of the factors.
The analysis suggests two main underlying gradients in the vegetation
dataset:

Vegetation growth and canopy structure are represented by RC1

Ground cover composition (grass versus bare or woody surfaces) are
represented by RC2

Together, these two factors explain just over half of the total variance
(55%) in the vegetation variables. The rotation makes these patterns
more interpretable by concentrating strong relationships within specific
factors while leaving unrelated variables with low loadings.

##-----

#### veg.fac3: `Varimin` rotation: A further orthogonal rotation.

`varimin` minimizes variance of factor loadings across factorized
variables.

```{r}
veg.fac3 <- principal(xdata, nfactor = 2, rotate = "varimin",
        method = "regression", scores = TRUE, covar = FALSE, cor = "cor")
veg.fac3
```

As usual, let's explore the component of `veg.fac3`

`communality` or `h2`:

```{r}
veg.fac3$communality
```

`uniqueness` or `u2`:

```{r}
veg.fac3$uniqueness
```

`loadings` are the corelation between the RCs and the original x
variables

```{r}
veg.fac3$loadings
```

`weights` are the multipliers to compute the factor scores:

```{r}
veg.fac3$weights
```

`Structure` is the same as `loadings`, which are the correlations:

```{r}
veg.fac3$Structure
```

`complexity` is Hoffman's index of complexity

```{r}
veg.fac3$complexity
mean(veg.fac3$complexity)
```

##### Correlations between the factors and the X variables

Still the same as `veg.fac3$loadings` and `veg.fac3$Structure` since
this is still an orthogonal rotation.

```{r}
round(cor(xdata, veg.fac3$scores), 2)
```

##### Correlations between the two factors

The two factors should still be uncorrelated

```{r}
round(cor(veg.fac3$scores, veg.fac3$scores), 2)
```

##### "Hand calculate" the scores using the weights coefficients.

Get the standardized xdata first

```{r}
Xscaled <- scale(xdata, center = TRUE, scale = TRUE)
```

The factor scores is the result of matrix multiplication between the
scaled x matrix and the weights matrix. The final matrix should be the
same as calling `veg.fac3$scores`.

```{r}
myscores <- Xscaled %*% veg.fac3$weights
head(myscores)
head(veg.fac3$scores)
```

##### Using just RC1, you get the variance of all X's accounted for by RC1 (factor 1) only.

**Note:** This works again since this is an orthogonal rotation. RC1 and
RC2 are independent (at right angles)

```{r}
RC1 <- veg.fac3$loadings[1:11, 1]
sum(RC1^2)
```

##### Using RC2, you get the variance of all X's accounted by RC2 only

```{r}
RC2 <- veg.fac3$loadings[1:11, 2]
sum(RC2^2)
```

##### Total communality and uniqueness.

Still 2 factors so it is the same as the "none" rotation.

Total communality:

```{r}
sum(RC1^2) + sum(RC2^2)
```

Uniqueness:

```{r}
11 - (sum(RC1^2) + sum(RC2^2))
```

##### Then, how much communality of each X variable versus uniqueness?

```{r}
commonbyX <- RC1^2 + RC2^2
commonbyX
sum(commonbyX)
```

##### How much is unique for each variable? What is left of the variance?

```{r}
uniquebyX <- 1 - (RC1^2 + RC2^2)
uniquebyX
sum(commonbyX)
```

Notes: 

- Varamin rotation is still an orthogonal rotation. Unlike varimax, which pushes loadings toward extreme values (near 0 or ±1), varimin minimizes the variance of loadings, spreading them more evenly across factors.
- The model explains 6.094 of 11 standardized variance units (55.4%), leaving 4.906 units as uniqueness or unexplained variance.
- The rotation redistributes variance between factors:

RC1: 4.161 units of variance (37.8%)

RC2: 1.933 units of variance (17.6%)

- Compared with the varimax solution, varimin concentrates more variance into RC1, making it a stronger general factor while RC2 becomes more secondary.

- RC1 represents a broad vegetation and ground structure gradient. Strong positive loadings on RC1 include May and June vegetation cover and height (0.64–0.77) and grass (0.77).

- Bare ground (−0.84) and rocks (−0.55) load negatively on RC1, indicating a gradient from dense vegetation to exposed substrate. Sites with high RC1 scores tend to have dense vegetation and grass cover, while low RC1 scores correspond to bare or rocky surfaces. Shrubs load strongly (0.62) on RC2, while June vegetation height and cover load moderately (0.56 and 0.58). Grass loads negatively on RC2 (−0.53), suggesting a contrast between shrub or taller vegetation and grass dominated ground cover. Some variables remain poorly represented by the two factor model, including bryophytes (h² ≈ 0.13), forbs (h² ≈ 0.12), and rocks (h² ≈ 0.30). These low communalities indicate that much of their variance is unique and not captured by the dominant vegetation gradients.


##### TODO: veg.fac4: Add in the code for `equamax` orthogonal rotation here (not graded)

Follow the structure of each examples above, add in the codes for
`equamax` orthogonal rotation here. Make sure you do not miss any
calculations. Create additional markdown commentaries and code chunks if
you wish! This part is not graded, however the outputs will be asked in
the upcoming TODO questions.

```{r}
#### veg.fac4: `Equamax` rotation: A further orthogonal rotation.

# Equamax is an orthogonal rotation that blends the goals of varimax and quartimax.
# It aims for a "middle ground" solution: simpler structure within factors while
# also avoiding one overly general factor dominating all variables.

library(psych)

veg.fac4 <- principal(
  xdata, nfactor = 2, rotate = "equamax",
  method = "regression", scores = TRUE, covar = FALSE, cor = "cor"
)

veg.fac4
```

```{r}
##### Explore key components of `veg.fac4`

# Communality (h2)
veg.fac4$communality

# Uniqueness (u2)
veg.fac4$uniqueness

# Loadings: correlations between the original x variables and RC1 & RC2
veg.fac4$loadings

# Weights: multipliers used to compute factor scores
veg.fac4$weights

# Structure: same as loadings here because equamax is an orthogonal rotation
veg.fac4$Structure

# Hoffman's index of complexity (and its mean)
veg.fac4$complexity
mean(veg.fac4$complexity)
```

```{r}
##### Correlations between the factors and the X variables

# Should match veg.fac4$loadings (and veg.fac4$Structure) because orthogonal rotation
round(cor(xdata, veg.fac4$scores), 2)
```

```{r}
##### Correlations between the two rotated factors

# Should be 0 (uncorrelated) for an orthogonal rotation
round(cor(veg.fac4$scores, veg.fac4$scores), 2)
```

```{r}
##### "Hand calculate" the factor scores using the weights coefficients

# Standardize X (consistent with correlation-matrix based solution)
Xscaled <- scale(xdata, center = TRUE, scale = TRUE)

# Matrix multiplication gives factor scores
myscores <- Xscaled %*% veg.fac4$weights

# Compare the first few rows to confirm they match
head(myscores)
head(veg.fac4$scores)
```

```{r}
##### Variance accounted for by each rotated component (RC1, RC2)

# Because equamax is orthogonal, we can still compute variance explained
# by squaring and summing loadings across variables.

RC1 <- veg.fac4$loadings[1:11, 1]
RC2 <- veg.fac4$loadings[1:11, 2]

sum(RC1^2)  # variance explained by RC1
sum(RC2^2)  # variance explained by RC2
```

```{r}
##### Total communality and uniqueness (two-factor solution)

total_communality <- sum(RC1^2) + sum(RC2^2)
total_uniqueness  <- 11 - total_communality  # total variance = 11 (cor matrix)

total_communality
total_uniqueness
```

```{r}
##### Communality and uniqueness by X variable

commonbyX <- RC1^2 + RC2^2
commonbyX
sum(commonbyX)   # should equal total_communality

uniquebyX <- 1 - commonbyX
uniquebyX
sum(uniquebyX)   # should equal total_uniqueness

# Sanity check: should sum to 11 variables worth of variance
sum(commonbyX) + sum(uniquebyX)
```

Notes: 
- Equamax attempts to simplify the structure across both variables and factors, rather than concentrating loadings in one factor or spreading them evenly. Because the rotation is orthogonal, the two factors remain uncorrelated, and the total variance explained by the model remains unchanged.
- The model explains 6.094 of the 11 standardized variance units (55.4%), leaving 4.906 units as uniqueness or unexplained variance.
- Factor 1 represents a vegetation structure gradient, with strong positive loadings for May and June vegetation cover and height (0.85–0.92). Bare ground, rocks, and bryophytes load negatively on Factor 1, indicating that lower scores correspond to exposed substrate or sparse vegetation.
- Factor 2 represents a ground cover composition contrast, with grass loading strongly and positively (0.92) and bare ground loading strongly negatively (−0.81).
- Shrubs and rocks also load negatively on Factor 2, suggesting a contrast between grass dominated surfaces and areas with bare ground, shrubs, or rocky substrate.
- Communalities and uniqueness values were examined to assess how well variables were explained by the two factor model.
- Vegetation structure variables, grass, and bare ground show high communalities, indicating the two factors explain a large portion of their variance.
- Bryophytes, forbs, and rocks show low communalities, meaning much of their variance is unique and not captured by the dominant gradients.
- Overall, the equamax rotation highlights two main ecological gradients:
Vegetation density and seasonal growth (Factor 1)
Ground cover composition, especially grass versus bare or shrub dominated surfaces (Factor 2).

Summary of all three rotations: 

- In every solution, the model explains 6.094 of the 11 standardized variance units (55.4%), leaving 4.906 units as uniqueness or unexplained variance.
- Factor 1 consistently represents a vegetation structure/presence gradient.
- High positive loadings for May and June vegetation cover and height indicate sites with dense and tall vegetation.
- Bare ground and rocks load negatively, indicating that lower factor scores correspond to exposed substrate or sparse vegetation.
- Factor 2 consistently represents a ground cover composition contrast.
- Grass loads strongly and positively, while bare ground and shrubs load negatively, distinguishing grass dominated surfaces from bare or shrub dominated areas.

- Under varimax, variance seemed balanced between the two factors (RC1 = 3.275, RC2 = 2.819). 
- Under varamin, more variance was loaded into the first factor (RC1 = 4.161; RC2 = 1.933), suggesting a strong dominant vegetation gradient with a weaker secondary factor.
- Under equamax, factors remained similar to the varimax solution but with slightly different loading distributions. 


####-------

##### End of TODO

**TODO Question 1:** In all analyses, **two factors were selected**.
Using a number of ways to determine how many factors to retain (as you
did using PCA in Exercise 2), is this number of factors justified for
these data?

Let's look at the eigenvalues and make a scree plot:

```{r}
# your codes here (if applicable)

#Take a look a the eigenvalues again 

# Eigenvalues from PCA on correlation matrix
eig <- veg.pca$sdev^2
eig

# Variance explained table
round(cbind(
  Eigenvalue = eig,
  PropVar = eig / sum(eig),
  CumPropVar = cumsum(eig) / sum(eig)
), 3)

# Scree plot with Kaiser line at 1
plot(eig, type = "b", pch = 16,
     xlab = "Component number", ylab = "Eigenvalue",
     main = "Scree plot (PCA eigenvalues)")
abline(h = 1, lty = 2)

```

The scree plot shows that 4 PC's are above the eigenvalue = 1 line.

Let's look at the variances more closely:

```{r}
# Extract eigenvalues
eigenvalues <- veg.pca$sdev^2

# Build table
pca_variance_table <- data.frame(
  PC = paste0("PC", 1:length(eigenvalues)),
  Eigenvalue = round(eigenvalues, 3),
  Variance_Proportion = round(eigenvalues / sum(eigenvalues), 3),
  Cumulative_Variance = round(cumsum(eigenvalues) / sum(eigenvalues), 3)
)

pca_variance_table
```

Finally, let's also apply the broken stick method.

```{r}
# Number of variables
p <- ncol(xdata)

# PCA eigenvalues
eig <- veg.pca$sdev^2

# Observed variance proportions
obs_var <- eig / sum(eig)

# Broken stick expected proportions
broken_stick <- sapply(1:p, function(k) {
  sum(1/(k:p)) / p
})

# Combine into a table
broken_table <- data.frame(
  PC = paste0("PC", 1:p),
  Observed = round(obs_var, 3),
  BrokenStick = round(broken_stick, 3)
)

broken_table

plot(obs_var, type = "b", pch = 16,
     xlab = "Principal Component",
     ylab = "Variance Proportion",
     main = "Broken Stick Comparison")

lines(broken_stick, type = "b", pch = 1, col = "red")
legend("topright",
       legend = c("Observed PCA variance", "Broken stick expectation"),
       col = c("black", "red"),
       lty = 1)


```
For this assignment, only the first two factors were kept, which together explained 55.4% of total variance in the original dataset. In Exercise 2, we used higher variance thresholds (70 and 80%). Though Exercise 2 focused on PCA alone, PCA was the foundation for factor analysis here. So, when comparing things, retaining only two factors seems to exclude a fair bit more of variance that may have otherwise been better explained by adding more factors. In keeping only the first 2 factors, nearly 45% of total variation is not explained, which indicates a much more simplified structure overall. Despite this, the consistently strong factor loadings for certain variables (vegetation variables in particular) suggest that the two retained factors still capture strong gradient patterns in the dataset. However, the relatively large proportion of unexplained variance suggests that additional factors could potentially provide a more holsitic representation of variable spread among sites.

For this dataset, we applied a few stats tests and plots to suggest that more than two factors could reasonably be retained. In the PCA section, the scree plot shows a clear "elbow" drop between PC1 and PC2, indicating that PC1 explained a large proportion of variance. However, the decline in eigenvalues here after PC2 is more gradual and does not drop off sharply when compared with the dataset in Exercise 2. This pattern suggests that additional components may still contain meaningful variation or represent more subtle patterns in vegetation.

We also considered the cumulative variance thresholds here as a guide for determining the number of retained factors. If a 70% cumulative variance threshold is applied, the first four components together explain 77.2% of the variance, exceeding this guideline. In contrast, an 80% threshold would require five components, which together explain 84.9% of the variance. Though PCA components and factors are not the same, these thresholds still provide a benchmark for evaluating how much of the variation could be represented by additional dimensions. 

We also applied the broken stick method. In this dataset, only the first component clearly exceeds the broken stick expectation, while the second and third components fall slightly below their expected values. Interestingly, this would suggest that only 1 component should be retained. While this indicates that the dataset is dominated by a single strong gradient, additional components may still capture secondary ecological patterns that were otherwise excluded. 

Taken together, these criteria suggest that retaining only two factors is somewhat conservative. While two factors capture vegetation patterns, the diagnostic tests indicate that three or four components could also be reasonably justified for representing the structure of the data, even if they capture weaker patterns. 

**End of Question 1**

**TODO Question 2:** For the unrotated factor pattern `veg.fac1`, what
is the loading value for Factor 1 with `mayvgcov`? What is the structure
value for this also? Explain what these values mean.

The loading value for Factor 1 with mayvgcov (May vegetation cover) in the unrotated factor pattern (veg.fac1) is 0.81, and the structure value is also 0.81. The loading shows the correlation between the variable and the factor, so a value of 0.81 indicates a strong positive relationship between May vegetation cover and Factor 1. This means that sites with higher scores on Factor 1 tend to have higher vegetation cover in May.

The structure value represents the correlation between the variable and the factor scores. Because the factors in this analysis are orthogonal via rotation, the structure and loading values are the same. Together, these values show that May vegetation cover is strongly related to Factor 1 and helps underscore the vegetation pattern represented by that factor.


```{r}
# extract the mayvgcov loading 
veg.fac1$loadings["mayvgcov", 1]

# extract the structure variable
veg.fac1$Structure["mayvgcov", 1]

#demonstrate they are equal, as 0 value indicates orthogonal
veg.fac1$loadings["mayvgcov", 1] - veg.fac1$Structure["mayvgcov", 1]

```

**End of Question 2**

**TODO Question 3:** Using orthogonal rotations only (`veg.fac1`,
`veg.fac2`, `veg.fac3`, `veg.fac4`), what orthogonal rotation (including
no rotation) gave the best results?

**a. Justify your choice as to why you believe these results are best
for achieving a simple and interpretable structure. Also indicate why
others were not chosen.**

Among the orthogonal rotations tested (veg.fac1 no rotation, veg.fac2
varimax, veg.fac3 varimin, and veg.fac4 equamax), the varimax rotation
(veg.fac2) produced the most interpretable and simplest factor
structure.

The varimax rotation maximizes the variance of squared loadings within
each factor, which tends to push loadings toward values close to 0 or
±1. This produces a clearer separation of variables among factors,
making interpretation easier. In the varimax solution, the first rotated
component (RC1) is strongly associated with the vegetation structure
variables, including May vegetation cover, May vegetation height, June
vegetation cover, and June vegetation height. The second component (RC2)
clearly captures variation in ground cover composition, with strong
loadings for grass and bare ground in opposite directions. This
separation creates two ecologically interpretable gradients: one
representing overall vegetation density and productivity, and the other
representing ground cover composition.

In contrast, the unrotated solution (veg.fac1) produced factors that
were harder to interpret because several variables loaded moderately on
both factors. Unrotated factors maximize variance explained rather than
interpretability, which often leads to factors that combine multiple
ecological gradients.

The varimin rotation (veg.fac3) minimized the variance of loadings,
resulting in one dominant factor that captured much of the overall
variance while the second factor contained weaker and more diffuse
loadings. This produced a less balanced and less interpretable
structure.

The equamax rotation (veg.fac4) attempted to balance the goals of
varimax and quartimax rotations but still produced several moderate
cross-loadings among variables. While interpretable, the resulting
structure was less clearly separated than in the varimax solution.

**b. Use the factor structure (correlations between original variables
and Factors) to decide which variables relate to each factor. Remember
to state what correlation cutoff** you are using (e.g., \|r\|\>= 0.5 or
\|r\|\>= 0.8); and **why** you made this choice.

A cutoff of \|r\| ≥ 0.5 was used to identify meaningful relationships
between variables and factors. This threshold was selected because
correlations above approximately 0.5 represent moderately strong
relationships and are commonly used in exploratory factor analysis to
identify variables that clearly load on a factor.

In this dataset, several variables show strong loadings near or above
this level, particularly the vegetation structure variables (e.g., May
and June vegetation cover and height) and ground cover variables such as
grass and bare ground. Using a threshold of \|r\| ≥ 0.5 allows these
dominant relationships to be clearly identified while excluding weaker
associations that could obscure interpretation of the factor structure.

A stricter cutoff such as \|r\| ≥ 0.8 would exclude variables that still
meaningfully contribute to the ecological gradients represented by the
factors, particularly those with moderate but interpretable loadings.
Conversely, a lower threshold such as \|r\| ≥ 0.3 would include many
weak correlations and reduce the clarity of the factor interpretation.
Therefore, \|r\| ≥ 0.5 provides a reasonable balance between capturing
meaningful relationships and maintaining a simple, interpretable factor
structure.

Using the varimax rotation (veg.fac2), the following variables met the
\|r\| ≥ 0.5 threshold.

Factor 1 (RC1) Variables strongly associated with Factor 1 include:

-   May vegetation cover (mayvgcov)
-   May vegetation height (mayvght)
-   June vegetation cover (junvgcov)
-   June vegetation height (junvght)

These variables all load strongly and positively on the first factor,
indicating that Factor 1 represents a gradient of overall vegetation
density and structure across the growing season.

Factor 2 (RC2) Variables strongly associated with Factor 2 include: -
Grass (grass) - Bare ground (baregrnd) - Shrubs (shrubs)

Grass loads strongly and positively on this factor, while bare ground
and shrubs load negatively, indicating that Factor 2 represents
variation in ground cover composition, distinguishing grass-dominated
sites from areas with more exposed soil or shrub cover. Several
variables, including bryophytes, forbs, rocks, and trees, do not exceed
the \|r\| ≥ 0.5 threshold for either factor. This suggests that these
variables contribute less strongly to the two-factor structure and may
represent more localized or independent variation in vegetation
composition.

**c. Add a possible label for each factor (e.g., vegetation structure
might be one factor). Use your answers from part b, but also consider
whether the correlations are positive or negative between a factor and
the original variables correlated with it.**

Factor 1 is strongly and positively correlated with May vegetation cover
(mayvgcov), May vegetation height (mayvght), June vegetation cover
(junvgcov), and June vegetation height (junvght). These variables all
describe the amount and vertical structure of vegetation at a site.
Because these correlations are positive, higher scores on Factor 1
correspond to sites with greater vegetation cover and taller vegetation
throughout the growing season. Lower scores on this factor would
correspond to sites with less vegetation structure and more open ground.
Therefore, Factor 1 can reasonably be labeled “Vegetation Structure” or
“Vegetation Density.”

Factor 2 is strongly associated with grass, which loads positively, and
bare ground and shrubs, which load negatively. This indicates that the
factor represents a contrast between grass-dominated sites and sites
characterized by exposed soil or shrub cover. High scores on this factor
correspond to areas with greater grass cover, while low scores
correspond to areas with more bare ground or woody vegetation. For this
reason, Factor 2 can be labeled “Ground Cover Composition.”

```{r}
# your codes here (if applicable)
```

**End of Question 3**

**TODO Question 4:** Based on the two factors retained and using your
selected orthogonal rotation:

a.  How much of the variability of each of the original X variables is
    accounted for (communality):

-   by each factor
-   for both factors combined?

```{r}
# Using varamax orthogonal rotation
L <- as.matrix(veg.fac2$loadings)  # 11 x 2 matrix of loadings (RC1, RC2)

# Communality contributions
comm_by_factor <- L^2
colnames(comm_by_factor) <- c("RC1_h2", "RC2_h2")

# Total communality
h2_total <- rowSums(comm_by_factor)

# Put into one table
comm_table <- data.frame(
  Variable = rownames(L),
  RC1_h2 = round(comm_by_factor[,1], 3),
  RC2_h2 = round(comm_by_factor[,2], 3),
  h2_total = round(h2_total, 3)
)

comm_table



```

Note: each squared loading represents the proportion of variance in that
variable explained by that factor, and the total communality (h²) is the
combined variance explained by both retained factors.

Factor 1 explains most of the variance in the vegetation structure
variables. For example, mayvgcov (0.650), mayvght (0.634), junvgcov
(0.790), and junvght (0.706) all have large contributions from Factor 1.
In each case, the majority of their variance is captured by this factor,
indicating that Factor 1 strongly represents overall vegetation
structure across the two months.

Factor 2 explains most of the variance in ground surface variables.
grass (0.865) and baregrnd (0.738) are especially strongly explained by
Factor 2. These variables have very small contributions from Factor 1
but large contributions from Factor 2, indicating that this factor
captures variation in ground cover conditions.

Some variables show weaker representation by the two factor solution.
For example, bryopht (0.132) and forbs (0.122) have very low
communalities, meaning that most of their variance is not explained by
the retained factors. Similarly, rocks (0.299) and shrubs (0.415) have
only moderate communalities, suggesting they are not strongly associated
with either factor.

Overall, the results show that the two-factor solution captures a large
portion of variance for the main vegetation structure variables and
ground cover variables, but explains relatively little variation for
bryophytes and forbs.

b.  How much of the variability of each of the original X variables is
    NOT accounted for (unique variance):

-   by each factor
-   for both factors combined?

```{r}
# starting from the communality table above
unique_table <- data.frame(
  Variable = comm_table$Variable,
  Unique_after_RC1 = round(1 - comm_table$RC1_h2, 3),
  Unique_after_RC2 = round(1 - comm_table$RC2_h2, 3),
  Unique_after_both = round(1 - comm_table$h2_total, 3)
)

unique_table
```

Looking at the combined unique variance for both factors, several
variables are well explained by the model. For example, grass (0.134),
junvgcov (0.127), and baregrnd (0.204) have relatively low unexplained
variance, meaning most of their variability is captured by the
two-factor solution. Similarly, mayvgcov (0.214) and mayvght (0.226)
also show relatively small amounts of unexplained variation.

In contrast, some variables retain large amounts of unique variance.
bryopht (0.868) and forbs (0.878) have very high unexplained
variability, indicating that the retained factors do not capture the
main patterns associated with these vegetation types. Rocks (0.701) and
shrubs (0.585) also show moderate to high unexplained variance,
suggesting weaker relationships with the main factor structure.

c.  For all X variables combined, what are the communality and unique
    variances:

-   by each factor
-   for both factors combined?

```{r}

# Eigenvalues from the correlation matrix of numeric X variables
eigcom <- eigen(cor(xdata, use = "pairwise.complete.obs"))$values

# Communality explained by each retained factor (first two eigenvalues)
comm_F1 <- eigcom[1]
comm_F2 <- eigcom[2]
comm_total <- comm_F1 + comm_F2

comm_F1
comm_F2
comm_total

# Proportion of variance explained (total variance = number of variables)
nvar <- ncol(xdata)

prop_F1 <- comm_F1 / nvar
prop_F2 <- comm_F2 / nvar
prop_total <- comm_total / nvar

prop_F1
prop_F2
prop_total

# Variance remaining (unique / unexplained by the two-factor solution)
unique_total <- nvar - comm_total
unique_prop <- unique_total / nvar

unique_total
unique_prop


```

From this, Factor 1 explains 39.33% of total variance, while Factor 2
explains 16.08%. Together this means that 55.40% of total variance is
explained, leaving 44.60% unexplained.

**End of Question 4**

**TODO Question 5:** Calculate the communality for `mayvgcov` for
**Factor 1** using your selected orthogonal rotation “by hand”. Confirm
that you obtained the same communality value as you reported in 4a for
this variable and factor.

A principal components analysis with two components and varimax rotation
was performed to identify the main underlying patterns in the vegetation
dataset.The results show that the first rotated component (RC1) has an
eigenvalue (SS loading) of 3.28, explaining 30% of the variance, while
the second component (RC2) has an eigenvalue of 2.82, explaining 26% of
the variance. Together, the two components explain about 55% of the
total variance in the dataset.

The model fit appears good, with an RMSR value of 0.08 and a fit based
on off-diagonal values of 0.96, indicating that the two-component
solution represents the relationships among the variables reasonably
well.

Looking at the factor loadings, RC1 has strong positive loadings for
variables such as May vegetation cover (0.81), May vegetation height
(0.80), June vegetation cover (0.89), and June vegetation height (0.84).
RC2 has strong loadings for grass (0.93) and bare ground (-0.86). This
suggests that the two components represent two main ecological
gradients: RC1: Vegetation structure or vegetation density (cover and
height of vegetation).RC2: Ground cover composition, particularly the
contrast between grass-dominated areas and bare ground.

Overall, the results indicate that most of the variation in the
vegetation data can be summarized by these two major gradients.

```{r}
# your codes here (if applicable)
# Running factor analysis with 2 factors (varimax rotation)
veg.fac2 <- principal(xdata,
                      nfactors = 2,
                      rotate = "varimax",
                      scores = TRUE)

# Results
print(veg.fac2)
```

**End of Question 5**

**TODO Question 6:** Calculate the factor score for **Factor 1** for the
first observation in the dataset, using your selected orthogonal
rotation. Confirm that this is the same as the score calculated using
`princomp()` in R.

The first rotated component (RC1) explains 29.8% of the total variance
in the dataset. RC1 has strong positive loadings for: May vegetation
cover (0.81), May vegetation height (0.80), June vegetation cover
(0.89), and June vegetation height (0.84)

These variables describe the amount and structure of vegetation at each
site. Sites with high RC1 scores tend to have taller vegetation and
higher vegetation cover, while sites with low RC1 scores tend to have
less vegetation and more exposed ground or rocks.

Therefore, RC1 can be interpreted as a vegetation density or vegetation
structure gradient.This component reflects differences among sites in
terms of how dense and tall the vegetation is.

```{r}
# your codes here (if applicable)
# Factor loadings
veg.fac2$loadings

```

**End of Question 6**

#### Oblimin rotation: An oblique rotation.

**Note:** we will call the factors as TC1 and TC2 now

Run the same analysis, but this time with `rotate="oblimin"`. `oblimin`
allows for the factors to be **correlated (not orthogonal)**.

```{r}
veg.fac5 <- principal(xdata, nfactor = 2, rotate = "oblimin",
  method = "regression", scores = TRUE, covar = FALSE, cor = "cor")

veg.fac5

# str(veg.fac5) uncomment this if you want to fully explore the object
```

`communality` and `uniqueness`:

```{r}
veg.fac5$communality
```

```{r}
veg.fac5$uniqueness
```

`weights` still contains the multipliers

```{r}
veg.fac5$weights # multipliers
```

Hoffman's index of complexity and mean complexity:

```{r}
veg.fac5$complexity  # the com in the table, Hoffman's index of complexity
mean(veg.fac5$complexity) # Hoffman's index of complexity also shown in the veg.fac5 summary
```

However, `loadings` is no longer the correlations between the TCs
(factor scores) and the X variables. Since *oblimin* is an **oblique
rotation**, this is now showing the **Factor Pattern**.

```{r}
veg.fac5$loadings # factor pattern
```

`Structure` is the correlations between the X variables and the factor
scores. Since this is an oblique rotation, you can see that Factor
Pattern is different from Factor Structure.

```{r}
veg.fac5$Structure
```

Let's verify that Structure is the simple correlations between the X
variables and factor scores by manually calculating with `cor()`.

```{r}
round(cor(xdata, veg.fac5$scores), 2)
```

We can can see that they are pretty much identical. And they are
completely different from `veg,fac5$loadings`, which is now the factor
pattern.

##### Correlations between the two factors

Let's explore the correlation between the factors

```{r}
round(cor(veg.fac5$scores, veg.fac5$scores), 2)
```

We can see that with an oblique rotation like `oblimin`, the factors are
**correlated**, which means the axes are **not 90 degrees (orthogonal)**
from each other. Thus, we cannot interpret each factor separately.

Since the factors are correlated, the variance accounted for by Factor 1
overlaps with the variance accounted for by Factor 2. We can calculate
the two variances, but this will be too big. Alternatively, we can
calculate just the independent bits, but that will be too small. This
will be demonstrate later on in this code.

**TODO Question 7:** Using the oblique rotation (`veg.fac5`):

a.  What is the correlation between the two factors? Is this “low
    enough” for factors to be considered “nearly independent”?

    The correlation between the two factors from the **oblique rotation
    (veg.fac5)** is **-0.309**, as shown in the Phi matrix. This
    indicates a moderate negative correlation between the two factors.

    Because the correlation is close to \|0.30\|, the factors are not
    completely independent, but the relationship is still relatively
    small. Therefore, the factors can be considered nearly independent,
    although some correlation exists between them.

b.  Repeat what you did in question 3b and 3c.

    Repeating what was done in Questions 3b and 3c, the loading and
    structure values for the variable mayvgcov were extracted. The
    loading value is 0.845, while the structure value is 0.880. These
    values are **n**ot identical, which demonstrates that in oblique
    rotation the loading and structure matrices differ because the
    factors are allowed to be correlated.

*Type your answer here:*

```{r}
# your codes here (if applicable)
# Run PCA with 2 factors using oblique rotation (oblimin)
veg.fac5 <- principal(xdata,
                      nfactors = 2,
                      rotate = "oblimin",
                      scores = TRUE)

#Checking the correlation between the two factors
veg.fac5$Phi

# Repeating what was done in Question 3b
veg.fac5$loadings["mayvgcov", 1]


# Repeating what was done in Question 3c
veg.fac5$Structure["mayvgcov", 1]


# the difference between loading and structure values
veg.fac5$loadings["mayvgcov", 1] - veg.fac5$Structure["mayvgcov", 1]
```

**End of Question 7**

##### "Hand calculate" the scores using the weights coefficients.

Standardize the original X variables

```{r}
Xscaled <- scale(xdata, center = TRUE, scale = TRUE)
```

Matrix multiplication

```{r}
myscores <- Xscaled %*% veg.fac5$weights
```

Verify that they are indeed similar

```{r}
head(myscores)
head(veg.fac5$scores)
```

We can try to get the variance accounted for by each factor, but notice
here that they will overlap overlap since the factors are **correlated**
now.

Variance accounted for by TC1. Notice here that we are using `Structure`
and not `loadings` since `Structure` is the correlations matrix, and it
is no longer the case for `loadings`.

```{r}
TC1 <- veg.fac5$Structure[1:11, 1]
sum(TC1^2)
```

what about TC2?

```{r}
TC2 <- veg.fac5$Structure[1:11, 2]
sum(TC2^2)
```

```{r}
sum(TC1^2) + sum(TC2^2) # now greater than the sum of two factors for none or orthogonal
```

the factors overlap!!! So, there is "double-counting" of some of the
variance explained.

##### What about the SS loadings?

```{r}
TC1 <- veg.fac5$loadings[1:11, 1]
sum(TC1^2) # How much does TC1 account for?
```

##### What about TC2?

```{r}
TC2 <- veg.fac5$loadings[1:11, 2]
sum(TC2^2) # How much does TC2 account for?
```

```{r}
sum(TC1^2) + sum(TC2^2) # now less than the sum of two factors for none or orthogonal
```

the factors overlap!!! Using the SS "loadings", the overlapping part is
removed.

##### What does `princomp()` report here?

```{r}
veg.fac5
```

So, using `Structure`, we get the double-counting of variances: TC1
(3.959523) + TC2 (3.835755) = 6.7953

Using `loadings`, we get the non-overlapping parts of variances: TC1
(3.60252) + TC2 (2.36017) = 5.9269

Reported in the `veg.fac5` summary? TC1 (3.67) + TC2 (2.43) = 6.1, which
is how much is accounted for by 2 factors!!

Still 2 factors so the same as the "none" rotation. As noted that is
harder to see here... since there is overlap of variance accounted for
given these non-independent factors.

```{r}
TC1 <- veg.fac5$Structure[1:11, 1]
```

##### What about TC2?

```{r}
TC2 <- veg.fac5$Structure[1:11, 2]
```

```{r}
commonbyX <- TC1^2 + TC2^2 # same as h2 in the veg.fac5 summary of outputs.
commonbyX # not really correct since factors are not independent!
sum(commonbyX)
```

##### How much is unique for each variable? What is left of the variance?

```{r}
uniquebyX <- 1 - (RC1^2 + RC2^2) #same as u2 in the veg.fac5 summary of outputs. 
uniquebyX # not really correct since factors are not independent!
sum(uniquebyX)
```

```{r}
sum(commonbyX) + sum(uniquebyX)
```

#### Which one to choose?

We want a simple structure that is easy to interpret and hopefully
label. We would like this to be from a orthogonal rotation if possible
so factors are indepedent!

**TODO Question 8:** Overall, based on your analysis and interpretation,
which rotation would you recommend? State your choice and then provide a
short paragraph to justify your choice and why others were not
recommended by you.

*Type your answer here:*

Overall, the varimax rotation is recommended because it produced the
clearest and most interpretable factor structure in this analysis. In
the varimax solution, the first component (RC1) is strongly associated
with vegetation structure variables such as May vegetation cover, May
vegetation height, June vegetation cover, and June vegetation height.
The second component (RC2) captures variation in ground cover
composition, with variables such as grass and bare ground loading
strongly but in opposite directions. This separation creates two
distinct ecological gradients, making the factors easier to interpret.

Although the oblique rotation (oblimin) allows factors to be correlated,
the correlation between the two factors was relatively small
(approximately −0.31). This suggests that the factors are only weakly
related and can be considered nearly independent. Because of this low
correlation, an orthogonal rotation such as varimax is appropriate and
simplifies interpretation.

Other rotations, such as varimin and equamax, were not recommended
because they produced less clear factor structures and introduced more
cross-loadings among variables. These cross-loadings made it more
difficult to assign a clear ecological meaning to each factor.
Therefore, the varimax rotation provides the simplest and most
interpretable representation of the underlying patterns in the
vegetation data.

**End of Question 8**

### Congratulation!
