---
title: "PondAnalysis, PCA_MDA_Cluster"
author: "Brandon Weigand"
date: "2026-04-03"
output:
  html_document: default
  word_document: default
---

To-do: 

Limitations for each section (Sandra & Brandon's respective stats sections)
Risk & ranking section (Brandon)
Conclusion (Sandra)
Edits and cuts (Sandra & Brandon)
Discussion (Recap of everything, limits)
Next steps section (Brandon)
Introduction and references (Brandon)


Presentation:

Sandra: 
- Cluster 
- Introduction (reasoning and research questions)
- PCA score regression 
- Discussion & Summary & Management Implications
- Pond risk section, dryout ponds section & final rankings (starting at "Pond-Specific Deficit Interpretation (Dryout Ponds)" to final interpretations of the risk ranking)

Brandon:
- variables and data cleaning (introduction)
- site selection 
- PCA
- MDA
- Pond risk axes intro and explanation, refill ponds section 
- Final conclusion







#######

/ INTRODUCTION
Research Question & Motivation

#######

Urban ponds are increasingly recognized as ecologically valuable yet hydrologically vulnerable components of city landscapes. They provide habitat for aquatic and semi-aquatic species, support urban biodiversity, and contribute to localized thermal and hydrologic regulation. However, many urban ponds exhibit pronounced seasonal fluctuations in water level, with some systems experiencing substantial drawdown or complete dry out during summer months. Under intensifying urban heat and shifting precipitation regimes, understanding the factors that structure these drying dynamics has become an important priority for both ecological research and urban management.

In Vancouver and surrounding urban areas, seasonal pond drying has become increasingly visible in recent years. Several ponds have exhibited repeated dry out events since at least 2023, with some sites drying in consecutive summers and prompting community and municipal responses. This recurrence suggests that dry out is not simply the result of short-term variability in precipitation, but rather reflects persistent environmental conditions that predispose certain ponds to water loss. At the same time, many ponds remain wet throughout the summer, either due to favorable environmental conditions or because they receive supplemental inflows and are exempt from water restrictions. This creates a key challenge: distinguishing ponds that are environmentally resilient from those whose persistence is maintained through management intervention.

From a hydrologic perspective, pond water levels are governed by the balance between multiple inflow and outflow pathways, including precipitation, surface runoff, groundwater exchange, evaporation, and evapotranspiration. Traditional water budget approaches attempt to quantify these components individually, but in urban systems, these pathways are often difficult to measure directly and vary substantially across space and time. Moreover, when the objective is to understand ecological outcomes such as hydroperiod and habitat persistence, it may be more informative to focus on the net hydrologic response of ponds rather than partitioning individual fluxes.

This study adopts that perspective by using net seasonal water loss as the primary response variable. Repeated measurements of pond depth throughout the field season capture both short-term gains associated with hydrologic inputs and longer-term declines driven by cumulative losses. These trajectories represent an integrated expression of all underlying hydrologic processes, allowing ponds to be compared based on realized drying behavior.

Rather than directly estimating individual fluxes, this study focuses on how variation in environmental context shapes differences in drying dynamics. Variables describing canopy cover, realized shading, thermal conditions, vegetation change, and basin depth were selected to represent distinct processes influencing energy exposure, water storage, and biological feedbacks. These variables do not directly quantify inflows or outflows, but instead capture the structural and microclimatic conditions under which water loss occurs.

Importantly, the study includes ponds representing three distinct hydrologic and management contexts: recurrent dryout ponds, stable ponds without continuous inflows, and refill ponds that receive supplemental water inputs. These categories are not treated as fixed or independent classes, but rather as outcomes that emerge from underlying environmental conditions and management practices. This distinction is critical, as ponds that remain wet under managed conditions may nonetheless share environmental characteristics with drying systems.

Within this framework, the study asks: how do environmental gradients structure variation among urban ponds, and how do these gradients relate to drying vulnerability under differing management conditions? Specifically, it addresses four questions: (1) what dominant environmental gradients characterize pond systems, (2) how these gradients relate to observed rates of seasonal water loss, (3) how ponds with differing hydrologic behaviors are distributed along these gradients, and (4) whether refill ponds occupy similar environmental space as dryout ponds, indicating potential vulnerability masked by supplemental inflows.

By linking multivariate environmental gradients to realized drying behavior, this approach shifts the focus from isolating individual hydrologic mechanisms to identifying the conditions under which drying occurs. In doing so, it provides a practical framework for identifying ponds that are potentially vulnerable to drying, even when that vulnerability is not immediately expressed under current management regimes.

*Research Questions:*
How do environmental gradients structure variation among urban ponds, and how do these gradients relate to drying vulnerability under differing management conditions?

Which ponds exhibit environmental characteristics associated with high drying vulnerability, and to what extent does management alter the expression of that vulnerability?

Sub-questions:

- What are the dominant environmental gradients that characterize variation among urban ponds, and how do these environmental gradients relate to observed rates of seasonal water loss?
- How are ponds with different hydrologic behaviors (dryout, stable, refill) distributed along these environmental gradients?
- Do refill ponds occupy similar positions in environmental space as dryout ponds, suggesting underlying vulnerability masked by supplemental inflows?
- Can environmental gradients be used to identify ponds that are likely vulnerable to drying under reduced or absent management intervention?


###########

#DATA DESCRIPTION

###########


#######

SITE SELECTION

#######

Ponds were selected to represent a range of observed hydrologic behaviors, including recurrent dryout systems, stable ponds without continuous inflows, and ponds maintained through supplemental water inputs. These categories were not treated as discrete or independent classes, but rather as positions along a continuum of drying vulnerability shaped by environmental conditions and management. Dryout ponds represent systems in which vulnerability is consistently expressed, while stable ponds persist without intervention and provide a critical contrast for identifying environmental conditions associated with resilience. Refill ponds, in contrast, allow for the evaluation of whether persistence reflects inherent environmental buffering or the influence of continued hydrologic support.

Sites were drawn from an initial pool of approximately 50 urban water bodies and intentionally selected to maximize variation across both environmental conditions and hydrologic outcomes. Classification of ponds was informed through consultation with the Vancouver Board of Parks and Recreation, ensuring consistency with observed management practices and documented drying histories.

Site selection was guided by the objective of capturing variation along key environmental gradients hypothesized to influence drying dynamics. In particular, ponds were chosen to span a range of surface areas and canopy cover conditions, the latter representing the most consistently available landscape variable across all candidate sites. This ensured representation of both highly exposed and heavily shaded systems, as well as variation in basin size and associated storage capacity.

Rather than employing random sampling, sites were purposefully selected to capture variability within and across hydrologic conditions, enabling the identification of continuous environmental gradients and their relationship to drying behavior. This gradient-focused approach is consistent with the study’s objective of linking multivariate environmental context to patterns of water loss, rather than estimating population-level averages.

To reduce spatial bias and avoid pseudo-replication, ponds were also selected to achieve broad geographic representation across the study area, while limiting overrepresentation of clustered sites. Where multiple ponds occurred within the same park or managed landscape (e.g., Queen Elizabeth Park or McCleery Golf Course), a single representative pond was selected. This ensured that each site contributed independent environmental information, allowing observed patterns to reflect broader gradients rather than localized site-specific conditions.


########

VARIABLES

#######

The study integrates field-based measurements, in-situ sensor data, and spatial datasets to examine patterns associated with excelerated rates of urban pond water loss.

Water level data were collected through repeated field visits, where pond depth was measured using meter sticks at fixed sampling locations (minimum of six visits per pond, roughly 3 weeks between visits).

Vegetation cover was estimated using a grid-based visual survey approach during each site visit, where 2 observers independently assessed percentage cover within predefined 20x20 meter grid cells and averaged their estimates. Water quality variables (pH and phosphate) were measured in the field using high-sensitivity test strips and vial-based assays.

Microclimate data (air temperature, water temperature, and humidity) were recorded using iButton loggers (MAXIM Inc.) deployed at each pond. These sensors recorded measurements at 20-minute intervals throughout the study period. Loggers were installed at centroid locations within each pond.

For shade data, ArcGIS Pro was used to model within-pond shading dynamics using a LiDAR-derived workflow. Digital elevation (DEM), surface (DSM), and canopy height (CHM) models were generated from high-resolution LiDAR data and used to represent terrain and vegetation structure. Solar position (azimuth and altitude) was calculated at hourly intervals on June 21, July 21, and August 21 between 07:00 and 18:00, and hillshade rasters were produced for each time step to simulate incident solar radiation. These outputs were reclassified into binary shade and sun exposure layers and aggregated to derive spatially explicit metrics of shade intensity, coverage, and distribution. Based on these outputs, areas within each pond were categorized into high shade, intermediate, and high exposure conditions to capture daily shade variability.

Spatial predictor variables were derived from geospatial datasets. Impervious surface and canopy cover were calculated within 50 m and 100 m buffers using a classified City of Vancouver land cover raster (4.8 m resolution) and processed in R using the landscapemetrics package.

All raw and cleaned data are available at: https://docs.google.com/spreadsheets/d/1INgRR1rHrhaoGNrlkq3hGot1tx46EbdfggESUael5dQ/edit?gid=0#gid=0.


########

DATA PREPARATION

########

```{r}
metrics_wide_2 <- read.csv("metrics_wide_final.csv")
colnames(metrics_wide_2)
```

########

# PCA Variable Selection

#######

*Multi-Criteria Variable Selection*

Variable selection was guided by a combination of statistical, ecological, and practical considerations. The relatively small sample size (n = 15 ponds) imposed a strict limit on the number of variables that could be included in the PCA without overfitting, necessitating a reduction to a small set of representative predictors. This constraint was particularly important given the large initial pool of 72 variables derived from extensive field measurements and spatial analyses, each reflecting hypothesized drivers of pond drying such as shading, thermal dynamics, vegetation, and hydrological context.

To address this, variables were selected to represent distinct ecological processes while minimizing redundancy. Many candidate variables were strongly correlated, reflecting the inherently covarying nature of urban pond systems, where shading, temperature, and vegetation dynamics are tightly linked. Correlation screening and iterative testing were therefore used to identify and remove collinear variables, retaining only those that captured unique aspects of system behaviour. Although PCA is well-suited to handling correlated predictors, this pre-selection step ensured that resulting components remained interpretable and not dominated by clusters of redundant variables.

Additional guidance was drawn from complementary analyses, including clustering and exploratory classification approaches, which helped clarify dominant gradients and reinforce the importance of certain variables over others. Finally, variable selection was informed by the applied context of the study, developed in collaboration with the Vancouver Board of Parks and Recreation, with an emphasis on identifying drivers that are both ecologically meaningful and relevant for urban pond management. Together, these considerations ensured that the final variable set reflects a balance between statistical robustness, ecological realism, and practical applicability.

Because the following are response variables, they will not be considered in the PCA:
- daily_loss_est
- total_depth_loss_cm
- total_volume_lost_m3
- mean_daily_volume_lost_m3
- z_loss_rate
- z_total_loss

*Final set:* 
canopy_50
av_hrs_air_july_mintomax
july_msf
veg_percent_change
depth_1

Process	Variables:
- Structural shading:	canopy_50
- Thermal dynamics:	av_hrs_air_july_mintomax
- Realized shade:	july_msf
- Vegetation feedback:	veg_percent_change
- Basin constraint:	depth_1

canopy → structure
thermal timing → energy dynamics
shade → realized conditions
vegetation → feedback
depth → constraint

*Justification of Final Predictor Variables*

To construct a parsimonious and interpretable PCA given the sample size of 15, variables were selected to represent distinct ecological mechanisms influencing pond drying, while minimizing redundancy among predictors. Each retained variable captures a unique component of the system, supported by correlation screening and ecological relevance.

*canopy_50* was included as a measure of structural shading potential surrounding each pond. Canopy cover strongly influences incoming solar radiation, mediates microclimate, and indirectly affects both water temperature and evaporation rates. Although correlated with shade metrics, canopy represents the physical capacity for shading rather than realized conditions, justifying its inclusion as a structural variable.

Reflects the percentage of canopy cover present within a 50 meter buffer drawn around each pond in ArcGIS. The R package “landscapemetrics” was used for this. Pixels within these buffers are about 4.8 square meter resolution (which is why the picture here looks blocky). The landscapemetrics package uses a landcover map from the City of Vancouver (2022) to identify landscape composition within the buffers. Other landscape classes (like grass, water, and sand) were also computed, but are not considered here. 

*av_hrs_air_july_mintomax* captures daily thermal dynamics during peak summer conditions, specifically the duration of warming from minimum to maximum air temperature. Unlike absolute temperature metrics, this variable reflects the rate and persistence of energy accumulation within the system. It was prioritized over water temperature variables to avoid incorporating downstream responses that are already influenced by shading and pond-specific conditions.

Represents thermal magnitude and thermal timing. Average number of hours required for air temperature to increase from the daily minimum to the daily maximum during each month for each pond. Units: hours. Represents the average daily warming duration of near-pond air conditions per month. Higher values mean air temperature typically took longer to warm from its daily low to its daily high. = mean(hours_min_to_max for cleaned July daily air temperature records, per pond).

This variable was calculated in two steps. First, for each pond and date, only cleaned air temperature observations (value_clean) were retained. The first occurrence of the daily minimum air temperature was identified, and then the first occurrence of the daily maximum after that minimum was located. The elapsed time between those two points was calculated in hours as hours_min_to_max. Second, these daily values were averaged across all June days for each pond. This means the metric is based on daily min to max timing, not on a single monthly temperature curve. This produces a monthly summary of daily warming duration, rather than a raw monthly temperature average. Because the code uses the first occurrence of the daily minimum and the first occurrence of the daily maximum after that, this metric is intentionally conservative and ordered in time. It avoids nonsensical negative durations and ensures that warming is measured in the expected daily direction. Because it relies on cleaned logger observations, it is less affected by anomalous spikes or abrupt artifacts. A pond could have moderate average air temperatures, but a very long warming period, which would suggest a slower daily heating trajectory. These values represent the average number of hours it took, on each day, for air temperature to go from its first daily minimum to its first daily maximum after that point

This variable does not represent:
- average daily temperature range
- average monthly temperature
- time from midnight to the daily maximum

*july_msf* (July mean shade fraction) was selected as the representative shade metric following redundancy analysis, which revealed near-perfect collinearity among alternative shade indices. This variable directly quantifies the proportion of time pond surfaces remain shaded during peak summer, representing realized shading conditions that influence thermal buffering and evaporative demand.

Average fraction of time that each pond pixel is shaded during each month. Units: proportion (0–1). Represents the temporal persistence of shading at the pixel level. Higher values indicate that locations within the pond remain shaded for a greater proportion of the day. = mean(pixel-level proportion of shaded time across all time steps, aggregated to pond, per month). This metric is calculated by determining, for each pixel, the proportion of modeled time steps during which it is shaded. These pixel-level values are then averaged across the pond surface. Monthly values reflect how consistently shading is maintained over time, rather than how much area is shaded at any single moment.

*veg_percent_change* was included to represent seasonal vegetation dynamics within each pond. Vegetation plays a dual role by contributing to shading and driving evapotranspiration, thereby influencing water loss through both physical and biological pathways. The use of percent change, rather than static cover, captures temporal dynamics in vegetation growth that align with cumulative drying patterns.

Relative change in vegetation cover (%) over the season, expressed as a percentage of the initial cover:
 (final − initial) / initial × 100.

*depth_1* was retained as a measure of basin storage capacity using the depth (cm) measured at the start of the season, representing a key physical constraint on water loss. Deeper ponds can buffer against short-term evaporative losses and exhibit slower rates of drawdown, making depth an essential variable for distinguishing between rapid and gradual drying systems. The meter stick measurements (in cm) at each sampling interval.

######

*Justification for Prioritizing July-Based Metrics*

Variables derived from July conditions were prioritized to capture peak seasonal dynamics in pond drying processes. July represents the period of maximum solar radiation, longest photoperiod, and highest atmospheric demand, resulting in the strongest coupling between thermal conditions, evaporative loss, and biological activity. During this time, differences among ponds are most pronounced, as structural shading, vegetation growth, and basin characteristics interact under sustained thermal stress. Earlier-season variables (e.g., June) may reflect transitional or lagged conditions, while late-season variables (e.g., August) increasingly incorporate cumulative effects of drying, including reduced water volume, vegetation expansion, and episodic rainfall inputs. By focusing on July, the analysis targets the period when environmental gradients are most directly linked to underlying mechanisms rather than downstream outcomes, allowing for clearer interpretation of drivers influencing variation in drying rates across ponds.

This is particularly relevant in urban pond systems, where shallow water bodies respond rapidly to atmospheric forcing, amplifying mid-summer differences in thermal exposure and water loss.

This frames July as peak forcing, not just a midpoint, and separates drivers (July) from outcomes (August).

*Synthesis*

Together, these variables capture complementary gradients of structural shading (canopy_50), realized shading (july_msf), thermal dynamics (av_hrs_air_july_mintomax), biological feedbacks (veg_percent_change), and physical storage capacity (depth_1). This combination ensures that the PCA reflects interacting ecological processes rather than redundant representations of a single mechanism, providing a robust framework for interpreting variation in pond drying dynamics.

These chosen variables reflect:
- one variable per mechanism
- removal of redundant shade metrics
- removal of weak variables
- removal of mixed-scale predictors

Correlation check:
```{r}
cor_matrix5 <- cor(
  metrics_wide_2[, c(
    "canopy_50",
    "av_hrs_air_july_mintomax",
    "july_msf",
    "veg_percent_change",
    "depth_1"
  )],
  use = "complete.obs"
)

round(cor_matrix5, 2)
```
The selected predictor set shows low to moderate correlations overall, with no pairwise relationships exceeding r = 0.6. This indicates that multicollinearity is not a major concern and supports the inclusion of all five variables in the PCA. The strongest relationship occurs between canopy cover and July MSF (r = 0.57), suggesting a shared underlying shading signal. However, this correlation is not strong enough to justify removing one of the variables, as they represent conceptually distinct processes: canopy cover reflects structural shading potential, while MSF captures realized shading under seasonal solar conditions.

The remaining variables exhibit relatively weak correlations, reinforcing their roles as independent drivers. Vegetation percent change is moderately negatively associated with canopy, shading, and thermal variables, suggesting that ponds with greater vegetation change tend to be more exposed and environmentally dynamic. Depth shows minimal correlation with other predictors, indicating that it captures a distinct aspect of pond capacity rather than microclimate or structure. Similarly, the temperature metric (average hours from minimum to maximum air temperature in July) appears largely independent, representing a separate dimension of thermal dynamics. Overall, the variable set is well-balanced, capturing multiple ecological mechanisms while maintaining sufficient independence for robust multivariate analysis.


####################################

#METHODS

####################################

To address the research questions, a combination of multivariate and comparative statistical approaches was used, including Principal Component Analysis (PCA), hierarchical cluster analysis, Multiple Discriminant Analysis (MDA), and a deficit-based vulnerability assessment. All environmental predictor variables were standardized prior to analysis to ensure comparability across units and scales.

*Principal Component Analysis (PCA)*

PCA was used to reduce dimensionality and identify the dominant environmental gradients structuring variation among ponds. The analysis included only predictor variables describing environmental conditions, such as canopy cover, shade metrics, thermal dynamics, vegetation change, and basin characteristics. Response variables related to water loss were excluded from the PCA to avoid circularity.

PCA was conducted across all ponds simultaneously, with each pond treated as a single observation. Resulting principal components represent orthogonal gradients derived from covariation among the original variables. Component loadings were used to interpret the environmental meaning of each axis, while pond scores along these axes indicate their position within multivariate environmental space .

To evaluate how environmental gradients relate to drying dynamics, pond scores on the primary PCA axes were compared to water loss metrics, including standardized drying rate and total seasonal loss, using correlation and regression analyses. This approach links multivariate environmental structure to observed drying behavior without directly modeling individual hydrologic fluxes.

*Cluster Analysis*

Hierarchical cluster analysis was used to assess whether ponds form distinct environmental groupings or vary along continuous gradients. Clustering was performed on the same standardized environmental variables used in the PCA, using a variance-minimizing linkage method.

Results were visualized using a dendrogram to evaluate similarity among ponds and identify potential group structure. Cluster membership was compared with both PCA positioning and observed drying behavior to determine whether ponds with similar environmental conditions and water loss patterns form discrete groups or are distributed along continuous gradients.

Cluster analysis was interpreted as complementary to PCA: while PCA identifies dominant axes of variation, clustering evaluates whether those gradients correspond to discrete environmental types or gradual transitions across ponds.

*Multiple Discriminant Analysis (MDA)*

Multiple Discriminant Analysis was used as a supplementary approach to assess the extent to which predefined hydrologic classes (dryout, stable, refill) separate in multivariate environmental space. Using PCA scores as input variables, MDA identifies linear combinations of axes that maximize separation among these groups.

This analysis was used diagnostically rather than as a primary classification tool, given the small sample size and the expectation that pond behavior reflects continuous processes rather than discrete categories. The strength and direction of group separation were evaluated to determine whether observed hydrologic classes align with underlying environmental gradients or overlap substantially.

*Deficit-Based Vulnerability Analysis*

To further interpret drying vulnerability, a deficit-based framework was applied using PCA results. The primary environmental gradient (PC1) was interpreted as an exposure–buffering axis, with lower scores indicating greater environmental exposure and reduced shading.

A vulnerability index was derived by transforming PC1 scores into an exposure-based deficit metric, where higher values represent greater deviation from environmentally buffered conditions. This metric was used to:

- compare refill ponds to the range of observed dryout conditions
- assess whether managed systems occupy environmentally vulnerable positions
- estimate relative drying susceptibility independent of observed hydrologic inputs

Predicted drying rates based on PCA-derived gradients were also compared across ponds to evaluate whether refill systems exhibit environmental characteristics associated with elevated drying risk.

*Integration*
Together, these analytical approaches provide a complementary framework for linking environmental structure to drying dynamics. Principal Component Analysis (PCA) identifies the dominant environmental gradients that structure variation among ponds, establishing the multivariate context in which drying occurs. Regression of water loss metrics against PCA scores then evaluates how these gradients relate to observed drying behavior. Cluster analysis complements this by assessing whether ponds form discrete environmental groupings or instead vary continuously along these gradients. Multiple Discriminant Analysis (MDA) further examines how predefined hydrologic classes are positioned within this multivariate space, providing a diagnostic assessment of whether observed pond behaviors align with or overlap across environmental gradients. Finally, the deficit-based vulnerability analysis builds directly on PCA results to quantify relative exposure and identify ponds that exhibit environmental characteristics associated with high drying susceptibility, regardless of current management status. Taken together, this framework allows for the identification of both expressed and potentially masked vulnerability, addressing how environmental gradients structure pond variation and how management may modify the realization of drying outcomes.




#################

# PCA

################

Data preparation
```{r}
library(dplyr)

# Select final variables
pca_vars <- metrics_wide_2 %>%
  dplyr::select(
    canopy_50,
    av_hrs_air_july_mintomax,
    july_msf,
    veg_percent_change,
    depth_1
  )

# Remove missing values
pca_vars_clean <- na.omit(pca_vars)
```

PCA Test (Scaled)
```{r}
pca_res <- prcomp(pca_vars_clean, scale. = TRUE)
```

Scree Plot
```{r}
# Variance explained
var_explained <- pca_res$sdev^2 / sum(pca_res$sdev^2)

plot(var_explained,
     type = "b",
     pch = 19,
     xlab = "Principal Component",
     ylab = "Proportion of Variance Explained",
     main = "Scree Plot")

abline(h = 0.2, col = "red", lty = 2)  # optional reference line
```

Summary
```{r}
summary(pca_res)
```

Loadings
```{r}
loadings <- pca_res$rotation
loadings
```

Basic Biplot
```{r}
library(ggplot2)

biplot(pca_res, scale = 0)
```

```{r}
library(ggplot2)
library(ggrepel)

# Scores
pca_scores <- as.data.frame(pca_res$x)

# Loadings
loadings_df <- as.data.frame(pca_res$rotation)
loadings_df$var <- rownames(loadings_df)

# Scale arrows
arrow_scale <- 3

ggplot(pca_scores, aes(x = PC1, y = PC2)) +
  
  geom_point(size = 3, color = "steelblue") +
  
  # Arrows
  geom_segment(data = loadings_df,
               aes(x = 0, y = 0,
                   xend = PC1 * arrow_scale,
                   yend = PC2 * arrow_scale),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "black") +
  
  geom_text_repel(data = loadings_df,
                  aes(x = PC1 * arrow_scale,
                      y = PC2 * arrow_scale,
                      label = var),
                  size = 4) +
  
  theme_minimal() +
  
  labs(title = "PCA Biplot (Environmental Variables)",
       x = paste0("PC1 (", round(summary(pca_res)$importance[2,1]*100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_res)$importance[2,2]*100, 1), "%)"))
```

With Pond Labels
```{r}
# Match metadata back to cleaned rows
pca_scores$pond <- metrics_wide_2$pond[complete.cases(pca_vars)]
pca_scores$drought_class <- metrics_wide_2$drought_class[complete.cases(pca_vars)]
```

Final Interpretive Biplot
```{r}
ggplot(pca_scores, aes(x = PC1, y = PC2, color = drought_class)) +
  
  geom_point(size = 4) +
  
  geom_text_repel(aes(label = pond),
                  size = 4,
                  max.overlaps = 20) +
  
  # Arrows
  geom_segment(data = loadings_df,
               aes(x = 0, y = 0,
                   xend = PC1 * arrow_scale,
                   yend = PC2 * arrow_scale),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "black") +
  
  geom_text_repel(data = loadings_df,
                  aes(x = PC1 * arrow_scale,
                      y = PC2 * arrow_scale,
                      label = var),
                  size = 4,
                      color = "black") +
  
  theme_minimal() +
  
  labs(title = "PCA Biplot with Drought Classification",
       x = paste0("PC1 (", round(summary(pca_res)$importance[2,1]*100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_res)$importance[2,2]*100, 1), "%)"))
```


# Interpretation 

PC1 controls: structural and realized shade buffering versus exposure
PC2 controls: basin capacity and vegetation feedbacks versus thermal timing
PC3 controls: thermal timing and depth relative to realized shade

Alternatively,

PC1 controls: how strongly ponds are protected from solar forcing
PC2 controls: how storage capacity and ecological change interact with warming dynamics
PC3 controls: how thermal timing and pond capacity are balanced against functional shade

#Interpretation:

*PCA Variance Structure and Component Retention*

The PCA revealed a multi-dimensional structure underlying variation among ponds, with no single dominant axis overwhelming the others. PC1 explained 37.4% of the variance, followed by PC2 (25.5%) and PC3 (24.7%), with the first two components together accounting for 62.9% of total variation and the first three explaining 87.6%. This indicates that pond conditions are structured by one primary environmental gradient alongside at least two meaningful secondary gradients. Based on this distribution, PC1 and PC2 were retained for primary interpretation and visualization, while PC3 was also examined due to its substantial contribution to overall variance.

*PC1: Structural and Realized Shading Versus Exposure and Vegetation Dynamics*

PC1 represents a dominant gradient of environmental buffering versus exposure. It is characterized by strong positive loadings for canopy cover (0.646) and July mean shade fraction (0.546), and a strong negative loading for vegetation percent change (-0.486). Depth contributes weakly in the positive direction, while thermal timing plays a minimal role.

This axis separates ponds with greater structural and realized shading from those that are more exposed and exhibit stronger seasonal vegetation change. Importantly, vegetation change does not behave as a buffering mechanism along this axis. Instead, it aligns with exposed conditions, suggesting that increased vegetation dynamics may reflect ecological responses to shallow, thermally stressed, or drying environments rather than resistance to them.

Mechanistically, PC1 captures a combined shading gradient that integrates both canopy structure and realized radiation conditions. Ponds with high PC1 scores are more buffered from solar input, while those with low scores are more exposed and dynamically changing. This axis likely represents the primary environmental gradient influencing pond vulnerability and is a strong candidate for subsequent deficit analysis.

*PC2: Basin Capacity and Vegetation Feedbacks Versus Thermal Timing*

PC2 captures a secondary gradient contrasting basin capacity and ecological dynamics with thermal behavior. It is defined by strong positive loadings for depth (0.656) and vegetation percent change (0.546), and a negative loading for July air warming duration (-0.429). Contributions from canopy and shade are present but comparatively weak.

This axis separates deeper ponds with greater seasonal vegetation change from ponds associated with longer daily warming trajectories in air temperature. Unlike PC1, this gradient does not primarily reflect shading conditions, but instead highlights the interaction between physical storage capacity, biological change, and atmospheric thermal dynamics.

Ecologically, this suggests that pond depth and vegetation dynamics form a distinct dimension of system behavior, potentially influencing resilience, internal processes, or water retention independently of shading. This axis is particularly important for distinguishing ponds that may not differ strongly in shade but vary in their capacity to buffer or respond to thermal forcing.

*PC3: Thermal Dynamics and Depth Versus Realized Shade*

PC3 represents an additional, independent gradient linking thermal timing and basin capacity against realized shading. It loads strongly and positively on July air warming duration (0.699) and depth (0.514), and negatively on July mean shade fraction (-0.481), with minimal contribution from vegetation and only weak influence from canopy.

This axis reveals that realized shading is not fully explained by canopy structure alone. While canopy and MSF align on PC1, they diverge here, indicating that functional shading conditions vary independently depending on pond configuration, solar exposure, and microclimate.

PC3 therefore captures finer-scale differences in how ponds experience and respond to thermal forcing, particularly under midsummer conditions. If linked to drying rates in subsequent analyses, this axis may highlight the importance of thermal timing and functional shade interactions beyond simple structural shading metrics.

*Minor Components (PC4 and PC5)*

PC4 (7.7%) and PC5 (4.7%) explain relatively small proportions of variance and do not represent dominant ecological gradients. While they contain mathematically valid structure, their contributions are secondary and are not central to the interpretation of pond dynamics. These components were therefore not emphasized in further analysis.

*Integrated Interpretation of Environmental Gradients*

Together, the PCA results support the conceptual framework used in variable selection, demonstrating that pond variability is structured by multiple interacting ecological processes rather than a single dominant driver.

PC1 captures a primary gradient of structural and realized shading versus exposure and ecological change, aligning closely with hypothesized mechanisms of solar buffering and environmental stress. PC2 introduces a distinct dimension related to basin capacity and vegetation dynamics, independent of shading effects. PC3 further refines this structure by separating thermal timing and depth from realized shading, highlighting the complexity of energy dynamics within urban pond systems.

Importantly, canopy cover and realized shade, while moderately correlated, do not behave identically across components, supporting their inclusion as separate variables. Vegetation change emerges as a dynamic and context-dependent factor, aligning with exposure on PC1 but with capacity on PC2. Depth and thermal timing both contribute independent and meaningful variation, reinforcing the importance of considering multiple interacting gradients when assessing pond behavior.

*Summary of PCA Results*

The PCA identified three key environmental gradients structuring variation among ponds: (1) a dominant shading and exposure gradient integrating canopy structure and realized shade (PC1), (2) a secondary gradient contrasting basin capacity and vegetation dynamics with thermal timing (PC2), and (3) a tertiary gradient linking thermal dynamics and depth against realized shading (PC3). These results indicate that pond drying dynamics are shaped by interacting effects of structural shading, functional shading, thermal behavior, vegetation feedbacks, and basin constraint. This multivariate structure provides a robust foundation for subsequent analyses linking environmental gradients to variation in pond drying rates.




##########

#PCA Score Regressions 

#########

*Justification for Modeling Water Loss Rate (z_loss_rate)*

To link environmental gradients identified in the PCA to an ecologically meaningful outcome, regression analyses were conducted using standardized water loss rate (z_loss_rate) as the response variable. This metric captures relative differences in drying intensity across ponds, allowing comparison independent of pond size and absolute depth change.

Recent water management policies in Vancouver have reduced or eliminated artificial refilling of many urban ponds, effectively shifting these systems toward more natural hydrological regimes. Under regional climatic conditions, characterized by low summer precipitation and sustained evaporative demand, seasonal drying is expected in many of these systems. While intermittent or ephemeral conditions are not inherently detrimental and can support distinct ecological communities, the rate at which ponds lose water remains a key determinant of habitat persistence and ecological function.

Rapid drying can compress hydroperiods beyond thresholds required for aquatic insects, amphibians, and other taxa, particularly in urban environments where elevated temperatures and altered microclimates may further accelerate water loss. In contrast, slower rates of decline may allow ponds to function as seasonal refugia, even in the absence of artificial inputs.

By modeling z_loss_rate as a function of PCA-derived environmental gradients, this analysis evaluates how combinations of structural shading, realized shading, thermal dynamics, vegetation change, and basin capacity influence drying behaviour. This approach explicitly links multivariate environmental structure to variation in drying rates, allowing identification of the key gradients associated with accelerated water loss.

Rather than aiming to prevent drying entirely, this framework focuses on understanding and managing hydroperiod dynamics under urban conditions, with particular emphasis on moderating the rate of decline. This distinction is critical, as it aligns management strategies with the ecological realities of urban pond systems operating under reduced water inputs.*Thus, managing hydroperiod length and rate of decline under urban stress underlines the purpose of this study.* 

Extract PCA scores and align with response
```{r}
# PCA scores
pca_scores <- as.data.frame(pca_res$x)

# Match rows used in PCA
complete_idx <- complete.cases(
  metrics_wide_2[, c(
    "canopy_50",
    "av_hrs_air_july_mintomax",
    "july_msf",
    "veg_percent_change",
    "depth_1"
  )]
)

# Build modeling dataframe
model_df <- pca_scores
model_df$z_loss_rate <- metrics_wide_2$z_loss_rate[complete_idx]
model_df$pond <- metrics_wide_2$pond[complete_idx]
```

Using only PC1 and PC2 score loadings, to prevent possible overfitting with n=15
```{r}
model_1 <- lm(z_loss_rate ~ PC1 + PC2, data = model_df)
summary(model_1)
```

Using PC1, PC2, PC3 scores
```{r}
model_2 <- lm(z_loss_rate ~ PC1 + PC2 + PC3, data = model_df)
summary(model_2)
```

Compare
```{r}
AIC(model_1, model_2)
```

Diagnostics
```{r}
par(mfrow = c(2,2))
plot(model_1)
```

Visualize
```{r}
library(ggplot2)

ggplot(model_df, aes(PC1, z_loss_rate)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal()

ggplot(model_df, aes(PC2, z_loss_rate)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal()
```

# Interpretation:

*Regression of Drying Rate on PCA-Derived Environmental Gradients*

To evaluate how multivariate environmental gradients influence pond drying behaviour, linear models were fit using standardized water loss rate (z_loss_rate) as the response variable and PCA scores as predictors. Two models were compared: a reduced model including the first two principal components (PC1 and PC2), and an expanded model including PC3. Model performance was assessed using coefficient significance, explained variance, and AIC.

*Model Results and Statistical Patterns*

Across both models, PC1 emerged as the only statistically significant predictor of drying rate. In the PC1 + PC2 model, PC1 showed a significant negative relationship with z_loss_rate (β = -0.415, p = 0.026), while PC2 was not significant (p = 0.693). This model explained 35.5% of the variance in drying rate (adjusted R² = 0.248), with a marginal overall model significance (p = 0.072).

Including PC3 increased the total explained variance (R² = 0.437; adjusted R² = 0.283), but neither PC2 nor PC3 were statistically significant predictors, and the overall model remained marginal (p = 0.087). The coefficient for PC3 was negative (β = -0.248) but not significant (p = 0.232). Model comparison using AIC showed negligible improvement with the inclusion of PC3 (ΔAIC ≈ 0.04), indicating that the added complexity does not meaningfully improve model performance.

Diagnostic plots suggest that model assumptions are reasonably met, with no strong evidence of nonlinearity or extreme heteroscedasticity. However, the residuals versus leverage plot indicates the presence of at least one moderately influential observation, which may affect coefficient stability given the small sample size.

*Mechanistic Interpretation of PCA Axes in Relation to Drying Rate*

The negative relationship between PC1 and z_loss_rate indicates that ponds with higher PC1 scores experience slower drying rates, while those with lower PC1 scores dry more rapidly. Based on PCA loadings, PC1 represents a gradient of structural and realized shading opposed to exposure and vegetation dynamics. This suggests that ponds with greater canopy cover and higher midsummer shade persistence are more buffered against evaporative loss, while more exposed ponds with stronger vegetation change tend to experience faster water loss.

This result provides strong support for the hypothesis that solar exposure and shading are primary drivers of drying rate in urban ponds. Importantly, this effect reflects the combined influence of both structural canopy and realized shading conditions, rather than either variable in isolation. Vegetation change, which loads negatively on PC1, appears to be associated with the exposed end of this gradient, suggesting that increased vegetation dynamics may co-occur with or respond to conditions that promote drying, rather than mitigating it.

In contrast, PC2 does not significantly explain variation in drying rate, despite representing a meaningful environmental gradient involving depth and vegetation change opposed to thermal timing. This indicates that while basin capacity and ecological dynamics structure differences among ponds, they are not primary determinants of drying rate at the scale captured by this dataset.

Similarly, PC3 does not significantly contribute to explaining drying rate, despite capturing variation in thermal timing and its relationship with depth and realized shade. Although this axis reflects important differences in how ponds experience midsummer energy input, these differences do not appear to translate directly into variation in overall drying rate, at least within the current sample.

*Integrated Interpretation and Implications*

Taken together, these results indicate that pond drying rates are primarily structured along a single dominant environmental gradient, corresponding to the balance between solar exposure and shading. While additional gradients related to basin capacity, vegetation dynamics, and thermal timing exist and contribute to overall environmental variation, they do not independently explain variation in drying rate once the primary shading gradient is accounted for.

This finding reinforces the interpretation of PC1 as the key mechanistic axis linking environmental structure to drying behaviour. It also validates the use of PCA as a dimensionality reduction tool, as it isolates the dominant driver of drying within a multivariate system characterized by interacting processes.

From a management perspective, this suggests that interventions targeting shade enhancement or reduction of solar exposure may be the most effective strategies for moderating drying rates in urban ponds. In contrast, modifications to depth or vegetation alone may have more limited or context-dependent effects on drying dynamics, unless they also influence the shading regime.

At the same time, the moderate explanatory power of the models (R² ≈ 0.36–0.44) and the lack of significance for secondary components highlight that drying is not controlled by a single factor alone. Instead, it reflects the combined influence of multiple interacting processes, with shading acting as the dominant but not exclusive driver.


############

# MDA

###########

*Use of Discriminant Analysis as a Supplementary Approach*

To further evaluate whether environmental gradients identified through PCA correspond to observed differences in pond drying outcomes, a multiple discriminant analysis (MDA) was conducted using drought class as the grouping variable. Whereas PCA preserves continuous variation in environmental conditions, MDA provides a constrained framework that maximizes separation among predefined groups, allowing assessment of whether these groups occupy distinct regions of environmental space.

Given the small sample size and the expectation that pond drying reflects a continuum rather than discrete categories, MDA was used strictly as a supplementary analytical tool. The objective was not to define or validate discrete drought classes, but to examine whether the dominant environmental gradients associated with drying behaviour also contribute to partial separation among these classes.

This distinction is important, as any observed separation is interpreted as evidence of underlying environmental structure rather than as confirmation of discrete ecological states. As such, MDA results are used to support and contextualize findings from PCA and regression analyses, rather than to establish independent conclusions.

Performing the MDA using PCA scores
```{r}
library(MASS)

# Use same model_df
model_df$drought_class <- as.factor(metrics_wide_2$drought_class[complete_idx])

mda_model <- lda(drought_class ~ PC1 + PC2, data = model_df)
```

MDA model
```{r}
mda_model
```

Predict + confusion matrix
```{r}
pred <- predict(mda_model)

table(Predicted = pred$class, Actual = model_df$drought_class)
```

Plot discriminant space
```{r}
lda_scores <- as.data.frame(pred$x)
lda_scores$class <- model_df$drought_class
lda_scores$pond <- model_df$pond

library(ggplot2)
library(ggrepel)

ggplot(lda_scores, aes(LD1, LD2, color = class)) +
  geom_point(size = 4) +
  geom_text_repel(aes(label = pond), size = 4) +
  theme_minimal() +
  labs(title = "MDA (LDA) of Drought Classes")
```

# Interpretation 

Multiple discriminant analysis (MDA) was used to evaluate whether environmental gradients identified through PCA correspond to differences among drought classes. Using PC1 and PC2 as predictors, the analysis revealed partial but not complete separation among classes, with most of the discriminatory power concentrated along a single axis (LD1). Classification accuracy was moderate, with several misclassifications, reinforcing that drought response represents a continuum rather than discrete groupings.

*Key Results*
Group Means (in PCA space)
- Dryout: PC1 = -0.025, PC2 = -0.643
- Refill: PC1 = -0.216, PC2 = -0.086
- Stable: PC1 = 0.242, PC2 = 0.729

This shows a clear shift:

- Stable ponds → higher PC1 and PC2
- Dryout ponds → lower PC2 and near-neutral PC1
- Refill ponds → intermediate

*Discriminant Functions*

LD1 (97.2% of separation):
- PC2 loading = 0.945 (dominant)
- PC1 loading = 0.149 (minor)

LD2 (2.8% of separation):
- PC1 loading = 0.669
- PC2 loading = -0.155

Interpretation:
- LD1 is essentially a PC2-driven axis
- LD2 is a weak PC1 refinement

*Classification Performance*

Confusion matrix:
- Correct classifications: 9 / 15 (~60%)
- Misclassifications: 6 / 15 (~40%)

Patterns:
- Dryout: 4 correct, 3 misclassified
- Refill: 2 correct, 3 misclassified
- Stable: 3 correct, 3 misclassified

No class is cleanly separated.

*Interpretation of Discriminant Axes*

The first discriminant axis (LD1) captures the overwhelming majority of separation among drought classes (97.2%) and is strongly dominated by PC2. Given that PC2 represents a gradient of basin capacity and vegetation feedbacks opposed to thermal timing, this indicates that differences among drought classes are most strongly associated with depth, vegetation dynamics, and aspects of thermal structure, rather than with shading. Group means reinforce this pattern, with stable ponds exhibiting high PC2 values and dryout ponds characterized by lower PC2 values. This suggests that ponds with greater storage capacity and more pronounced vegetation dynamics tend to align with stable conditions, while those with lower capacity and differing thermal behavior are more likely to be associated with drying. Notably, this pattern contrasts with the regression results, where PC2 did not significantly explain variation in drying rate.

The second discriminant axis (LD2) explains only a small proportion of variation (2.8%) and is primarily associated with PC1, which represents the gradient of structural and realized shading versus exposure. Because this axis contributes minimally to overall class separation, it should be interpreted cautiously. At most, it suggests a minor secondary role for shading in refining differences among drought classes, but not as a dominant mechanism driving group separation.

*Integration with PCA and Regression Results*

Taken together, the PCA, regression, and MDA results provide a consistent but nuanced picture of how environmental gradients influence pond drying dynamics. Both PCA and MDA confirm that ponds are structured along meaningful environmental gradients and that these gradients are related to drying outcomes. However, the two approaches highlight different aspects of this relationship.

Regression results indicate that PC1, representing shading and exposure, is the primary driver of continuous variation in drying rate, with more shaded ponds drying more slowly. In contrast, MDA shows that separation among categorical drought classes is driven primarily by PC2, which reflects basin capacity, vegetation dynamics, and thermal structure. This distinction suggests that the factors controlling how fast ponds dry are not identical to those that differentiate broader drying outcomes.

This apparent mismatch is not contradictory, but instead reveals a layered system. Shading-related processes captured by PC1 appear to regulate the rate of water loss, while capacity and system-level characteristics represented by PC2 influence how ponds are ultimately categorized as dryout, refill, or stable. In this sense, PC1 governs drying intensity, whereas PC2 contributes to broader system state tendencies.

*Continuum of Drying Responses*

The moderate classification accuracy (approximately 60%) and visible overlap among classes in discriminant space further indicate that drought classes do not form distinct environmental clusters. Instead, ponds occupy overlapping regions along continuous environmental gradients, where similar conditions can lead to different outcomes. This suggests that drying behavior is influenced by interacting factors, potential thresholds, or additional variables not fully captured in the current analysis.

These findings support the use of PCA and regression as primary analytical tools, as they better capture the continuous and multivariate nature of pond drying processes, while MDA serves to illustrate how these gradients relate to categorical outcomes without imposing artificial boundaries.

*Final Interpretation*

Overall, the MDA results support the presence of structured environmental gradients but do not indicate strong separation among drought classes. Most of the discriminant power is driven by PC2, reflecting differences in basin capacity, vegetation dynamics, and thermal behavior, while the primary shading gradient represented by PC1 contributes less to categorical separation.

When considered alongside regression results, this suggests that shading-related processes are the dominant drivers of drying rate, whereas capacity and system-level characteristics influence broader differences in drought outcomes without directly controlling drying intensity. Together, these results reinforce that pond drying dynamics are best understood as a continuous, multivariate process shaped by interacting environmental gradients, rather than as a set of discrete ecological states.


######## 

CLUSTER ANALYSIS

######## 

*Clustering Ponds Based on Environmental Conditions/Gradient*

To further investigate the environmental gradients influencing pond
drying, hierarchical cluster analysis using Ward’s method was conducted
using the selected predictor variables. This analysis builds on the
variable selection process, which identified a reduced set of key
environmental factors representing shading, thermal conditions, and
urbanization, while minimizing redundancy and ensuring interpretability.

The cluster analysis was designed to directly address the leading
research question: which environmental gradients differentiate ponds
that dry out from those that remain stable or receive supplemental
water? In addition, it responds to the secondary question of whether
ponds form distinct environmental groupings or instead fall along a
continuous gradient of conditions.

By grouping ponds based on similarities in the selected predictor
variables, the analysis provides a way to identify whether discrete
environmental “types” of ponds exist (e.g., highly urban and exposed vs.
shaded and cooler), or whether variation in pond conditions and drying
rates is more gradual. This approach complements earlier analyses (e.g.,
PCA and redundancy analysis) by translating continuous environmental
gradients into potential group structures that can be more easily
interpreted and compared.

Cluster analysis also allows for evaluation of the hypothesis that ponds
with higher rates of water loss are characterized by warmer
temperatures, reduced shading, and greater exposure to urban conditions.
By comparing drying rates across clusters, it becomes possible to assess
whether these conditions form distinct groups or whether drying
behaviour varies continuously across environmental gradients.

Data preparation

```{r}
# Selecting already chosen predictor variables
selected_vars <- metrics_wide_2[, c(
  "canopy_50",
    "av_hrs_air_july_mintomax",
    "july_msf",
    "veg_percent_change",
    "depth_1"
)]

# Standardize variables
selected_scaled <- scale(selected_vars)
```

Running cluster analysis
```{r}
# create distance matrix
dist_matrix <- dist(selected_scaled , method = "euclidean")
```

```{r}
# Hierarchical clustering (Ward’s method)
hc <- hclust(dist_matrix, method = "ward.D2")
```

```{r}
library(dendextend)

# Convert to dendrogram
dend <- as.dendrogram(hc)

# FIX labels to use pond names
labels(dend) <- metrics_wide_2$pond_id[order.dendrogram(dend)]

# Create named drought class vector
drought_class <- metrics_wide_2$drought_class
names(drought_class) <- metrics_wide_2$pond_id

# Assign colors (your exact palette)
labels_colors(dend) <- ifelse(drought_class[labels(dend)] == "dryout", "#F8766D",
                        ifelse(drought_class[labels(dend)] == "refill", "#00BA38",
                               "#619CFF"))
# Plot
plot(dend,
     main = "Dendrogram of Urban Ponds",
     xlab = "",
     sub = "",
     ylab = "Height")

# Legend
legend("topright",
       legend = c("dryout", "refill", "stable"),
       fill = c("#F8766D", "#00BA38", "#619CFF"),
       border = NA,
       bty = "n")
```

#Interpretation

The hierarchical cluster analysis using Ward’s method revealed three
primary clusters of ponds based on similarities in environmental
conditions. At a cut height of approximately 5–6, the dendrogram
separates the ponds into three main groups. The first cluster consists
of a single pond (VAN), which joins the rest of the dendrogram at a high
dissimilarity level, indicating that it is environmentally distinct from
all other ponds. The second cluster includes BGL, RUP, AVL, SCT, BOO,
CHS, GVI, and QED, and represents a group of ponds with moderate
similarity across the selected environmental variables. The third
cluster contains BGS, MCC, TRT, MSP, CSN, and JCO, which are more
tightly grouped at lower heights, indicating greater similarity among
these ponds.

Across the dendrogram, drought classes (dryout, refill, stable) are
distributed across all clusters rather than forming distinct groupings.
Ponds with different drying outcomes are often grouped together,
indicating that similar environmental conditions can correspond to
different drying behaviours.

This lack of clear separation from the dendrogram support earlier
findings from PCA, regression, and discriminant analysis, showing that
ponds are structured along continuous environmental gradients rather
than forming distinct types. While the cluster analysis identifies
groupings in environmental space, these groupings represent positions
along gradients of shading, thermal dynamics, vegetation feedbacks, and
basin capacity, rather than fixed ecological states. Overall, while
three clusters can be identified, the dendrogram does not show clear
separation of ponds based on drought class, suggesting that
environmental variation is gradual rather than divided into distinct
categories.

```{r}
# Compute WSS for different k
wss <- numeric(10)

for (k in 1:10) {
  kmeans_result <- kmeans(selected_scaled, centers = k, nstart = 25)
  wss[k] <- kmeans_result$tot.withinss
}

# Plot elbow curve
plot(1:10, wss,
     type = "b",
     pch = 19,
     xlab = "Number of clusters (k)",
     ylab = "Total Within-Cluster Sum of Squares (WSS)",
     main = "Elbow Method for Optimal Clusters")
```

#Interpretation

The elbow method was used to evaluate the optimal number of clusters for
grouping ponds based on environmental conditions, with the aim of
assessing whether ponds form distinct patterns or fall along a
continuous gradient of conditions. The total within-cluster sum of
squares (WSS) was calculated for values of k ranging from 1 to 10. The
results show a sharp decrease in WSS from k = 1 to k = 4, indicating
that increasing the number of clusters substantially improves the
grouping of ponds within this range. Beyond k = 4, the rate of decrease
becomes more gradual, suggesting diminishing returns in reducing
within-cluster variation. This pattern indicates that the elbow occurs
around k = 4. However, the transition is relatively smooth rather than
sharply defined, suggesting that there is no single, clearly optimal
number of clusters based on WSS alone.

*Justification for Selecting k = 3 Clusters*

Although the elbow method suggests k = 4 as a potential optimal number
of clusters, this solution did not produce groupings that were
ecologically meaningful in the context of the study. Specifically, the
four-cluster solution resulted in over-partitioning of ponds into
smaller groups that did not clearly correspond to differences in
environmental gradients or drying behaviour. In contrast, the dendrogram
revealed a clear and interpretable separation into three primary
clusters, capturing the dominant structure of environmental similarity
among ponds. These three clusters align more closely with the study’s
objective of identifying environmental gradients that differentiate
drying behaviour, while maintaining interpretability in terms of
shading, thermal dynamics, vegetation feedbacks, and basin capacity.

Importantly, the three-cluster solution also supports the broader
research question of whether ponds form distinct groups or fall along a
continuum. The clusters identified at k = 3 show substantial overlap in
drought classes, indicating that ponds with different rates of water
loss are not cleanly separated into discrete categories. Instead, this
pattern reinforces the interpretation that pond drying dynamics are
structured along continuous environmental gradients. Therefore, k = 3
was selected as the most appropriate number of clusters, balancing
statistical evidence with ecological relevance and consistency with the
overall analytical framework.

```{r}
# Cut dendrogram into 3 clusters
metrics_wide_2$Cluster3 <- cutree(hc, k = 3)
```

*Comparing Clusters with Water loss (Z_loss_rate)*

```{r}
# Mean water loss per cluster
aggregate(metrics_wide_2$z_loss_rate,
          by = list(Cluster = metrics_wide_2$Cluster3),
          mean)

# Mean + SD
aggregate(metrics_wide_2$z_loss_rate,
          by = list(Cluster = metrics_wide_2$Cluster3),
          function(x) c(mean = mean(x), sd = sd(x)))

boxplot(z_loss_rate ~ factor(Cluster3,
                             levels = c(1, 2, 3),
                             labels = c("Cluster 1", "Cluster 2", "Cluster 3")),
        data = metrics_wide_2,
        col = c("#F8766D", "#00BA38", "#619CFF"),
        main = "Water Loss (z_loss_rate) Across Clusters",
        xlab = "Cluster",
        ylab = "z_loss_rate")
```

#Interpretation of Water Loss Across Clusters

0 = average drying
positive (+) = faster drying
negative (−) = slower drying

The average standardized water loss rate (z_loss_rate) varies across the three clusters, indicating differences in drying behaviour among groups. Cluster 1 has a negative mean value (-0.75), indicating that ponds in this cluster tend to experience lower-than-average water loss rates and therefore represent slower-drying systems. In contrast, Cluster 2 (mean = 0.26) and Cluster 3 (mean = 0.43) both exhibit positive mean values, indicating higher-than-average water loss rates. Among these, Cluster 3 shows the highest mean water loss, suggesting that ponds in this group are associated with more rapid drying conditions. However, the differences between Cluster 2 and Cluster 3 are relatively modest, indicating that while both clusters represent faster-drying systems, they may differ in degree rather than in fundamentally distinct drying behaviour.

The boxplot further illustrates the distribution of water loss within each cluster, highlighting both central tendency and variability. Cluster 1 shows consistently low values of z_loss_rate, with most observations falling below zero, indicating relatively uniform slow-drying conditions. Cluster 2 exhibits a wider spread of values, ranging from negative to strongly positive, suggesting greater variability in drying behaviour within this group. This indicates that ponds in Cluster 2 may experience a broader range of environmental conditions, leading to both moderate and high rates of water loss. Cluster 3 appears to have consistently positive values with minimal spread, indicating that ponds in this cluster tend to experience uniformly higher water loss rates. Overall, the boxplot shows that while clusters differ in average drying behaviour, there is still some overlap in values, particularly between Clusters 1 and 2, reinforcing that drying dynamics are not strictly separated into discrete categories.

*Integrated Interpretation of Cluster Analysis in Relation to PCA and MDA*

Together, the clustering results and associated water loss patterns indicate that ponds can be broadly grouped into three categories based on environmental similarity and drying behaviour: slower-drying systems (Cluster 1), intermediate and variable systems (Cluster 2), and more consistently faster-drying systems (Cluster 3). These groupings reflect general differences in environmental conditions and their influence on water loss rates, with Cluster 1 characterized by lower-than-average drying, Cluster 2 showing moderate and variable behaviour, and Cluster 3 associated with consistently higher rates of water loss. However, these clusters do not represent sharply defined ecological types. The observed overlap in water loss values across clusters, along with the gradual transitions between them, indicates that pond behaviour is not separated into discrete categories. Instead, ponds occupy positions along a continuous spectrum of environmental conditions and drying responses. The clustering therefore captures broad patterns in the data while highlighting the absence of strict boundaries between pond types.

These findings are strongly consistent with the results of the PCA and MDA analyses, which also demonstrate that pond dynamics are structured along continuous environmental gradients rather than discrete groupings. PCA identified shading and solar exposure (PC1) as the dominant gradient controlling variation in drying rates, with more shaded ponds drying more slowly and more exposed ponds drying more rapidly. In contrast, MDA showed only partial separation among drought classes (dryout, refill, stable), indicating that these categories do not occupy entirely distinct regions of environmental space. The cluster analysis complements these results by showing that, although ponds can be grouped based on overall environmental similarity, these groupings do not align cleanly with drying outcomes or drought classifications.

Instead, the clusters represent different regions along underlying environmental gradients defined by shading, thermal dynamics, vegetation feedbacks, and basin capacity. The lack of clear separation among clusters reinforces the interpretation that multiple interacting processes shape pond behaviour, and that similar environmental conditions can lead to different drying outcomes. Overall, this integrated analysis supports the conclusion that urban pond drying dynamics are best understood as a continuous, multivariate process, where environmental gradients influence drying behaviour in a gradual and overlapping manner rather than producing distinct ecological types.



#########

Pond-based Risk Axes

#########

To translate multivariate results into pond-specific management insights, PC1 was used as a composite index of exposure-related deficit. Because PC1 represents a gradient of increasing canopy cover and shading opposed by vegetation-driven exposure, lower PC1 scores indicate conditions associated with faster drying. Deficit scores were therefore calculated as the inverse of PC1, allowing ponds to be ranked according to their relative position along this gradient. To further diagnose the sources of deficit, variable-level contributions to PC1 were decomposed using PCA loadings and standardized values. This approach identifies which factors, such as low canopy cover or high vegetation change, contribute most strongly to each pond’s position along the exposure gradient. By linking these deficits directly to the primary axis controlling drying rate, this method provides a statistically grounded basis for prioritizing site-specific management interventions.

Standardize PC1 scores into a “deficit” metric.
```{r}
model_df$pc1_deficit_z <- scale(-model_df$PC1)
```

- Larger values = greater “deficit” relative to optimal conditions
- Directly tied to your regression result
- Multivariate by construction

Each variable contributes to PC1 with a loading. So a deficit in a variable should be interpreted relative to its loading, and its standardized value.

Method: Signed contribution to PC1 for each pond
```{r}
pca_result <- prcomp(
  metrics_wide_2[, c("canopy_50",
                     "av_hrs_air_july_mintomax",
                     "july_msf",
                     "veg_percent_change",
                     "depth_1")],
  scale. = TRUE
)

scaled_data <- scale(metrics_wide_2[, c("canopy_50",
                                        "av_hrs_air_july_mintomax",
                                        "july_msf",
                                        "veg_percent_change",
                                        "depth_1")])

loadings <- pca_result$rotation[, "PC1"]

loadings

```
Calculate contributions
```{r}
contributions <- sweep(scaled_data, 2, loadings, "*")
```

Calculate deficits
```{r}
deficits <- ifelse(contributions < 0, abs(contributions), 0)
deficits_df <- as.data.frame(deficits)
```

Add pond names
```{r}
deficits_df$pond <- metrics_wide_2$pond

#add overall PC1 deficit score
model_df$pc1_deficit <- max(model_df$PC1) - model_df$PC1
```

Visual
```{r}
#install.packages("reshape2")
library(reshape2)
library(ggplot2)

deficits_long <- melt(deficits_df, id.vars = "pond")

ggplot(deficits_long, aes(x = variable, y = value, fill = pond)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(y = "Deficit Contribution", x = "Variable") +
  theme_minimal()
```
Each bar represents:

How much that variable is pulling a pond toward the “high drying risk” side of PC1

Important points:

Values are only shown when they contribute negatively to PC1
Larger bars = stronger contribution to drying risk
Zero means that variable is not contributing to risk for that pond


## Interpretation of Pond-Based Risk Axes
*Structure of the PC1 Deficit Metric*

The PC1-based deficit metric provides a pond-specific measure of exposure-related risk, directly linked to the primary gradient controlling drying rate. Because PC1 is defined by positive contributions from canopy cover (0.646) and July mean shade fraction (0.546), and a negative contribution from vegetation percent change (-0.486), ponds with low PC1 scores are characterized by reduced shading and increased vegetation-driven exposure. Depth (0.139) and thermal timing (0.169) contribute only weakly to this axis.

By inverting PC1, higher deficit scores therefore represent ponds that are more exposed, less shaded, and more dynamically changing, conditions associated with faster drying in the regression analysis. Decomposing PC1 into variable-level contributions allows identification of which specific factors are driving this exposure for each pond.

*Dominant Drivers of Risk Across Ponds*

Across ponds, the largest deficit contributions are consistently associated with canopy cover, July mean shade fraction, and vegetation percent change, reflecting the dominant structure of PC1. In most cases, high risk is driven by a combination of low canopy cover and reduced realized shading, often accompanied by positive vegetation change, which reinforces exposure rather than mitigating it.

Canopy cover emerges as the most consistent contributor to deficit, with several ponds showing large negative contributions along this variable. July MSF shows a similar pattern, indicating that both structural and realized shading are aligned in driving exposure-related risk. This reinforces the earlier PCA result that these variables operate together as a shared shading gradient.

Vegetation percent change contributes strongly to deficit in a subset of ponds, particularly where large positive values coincide with low shading. Because this variable loads negatively on PC1, increases in vegetation are associated with movement toward the high-risk end of the gradient. This suggests that vegetation dynamics are not acting as a buffering mechanism, but instead are linked to exposed, rapidly changing conditions, potentially reflecting shallow water expansion or ecological responses to drying.

*Secondary and Minor Contributors*

In contrast, depth and thermal timing contribute relatively little to PC1-based deficit across most ponds. While some ponds show moderate contributions from these variables, their overall influence is small compared to shading-related factors. This aligns with the regression results, which showed that these variables do not independently control drying rate.

However, their presence in some ponds suggests that they may still influence system behavior in combination with other factors, particularly in shaping broader drought outcomes rather than directly controlling drying intensity.

*Pond-Specific Risk Profiles*

The deficit decomposition highlights that while ponds share a common exposure gradient, the sources of risk differ among ponds. Some ponds are primarily driven by low canopy cover, others by reduced realized shading, and others by strong vegetation change. In several cases, multiple factors act together, producing compounded exposure.

This reinforces the interpretation that ponds occupy different positions along a shared environmental continuum, but arrive there through different combinations of underlying conditions. As a result, two ponds with similar overall deficit scores may require different management interventions depending on which variables are driving their position along PC1.

*Management-Relevant Interpretation*

These results provide a direct link between the statistical analysis and actionable management priorities. Because PC1 is the dominant predictor of drying rate, variables contributing to PC1 deficit represent the most immediate targets for intervention.

The consistent importance of canopy cover and realized shading indicates that increasing shading is the most effective system-wide strategy for reducing drying risk. This may involve enhancing surrounding canopy, modifying vegetation structure, or otherwise increasing shade persistence during peak summer conditions.

At the same time, the role of vegetation change suggests that vegetation should be managed strategically rather than assumed to provide passive protection. In some ponds, increasing vegetation may actually coincide with greater exposure and faster drying, indicating a need for more targeted management of plant dynamics.

Although depth and thermal timing are not primary drivers of drying rate, they may still influence how ponds respond to exposure, particularly in terms of persistence and resilience. As such, they remain relevant for site-specific considerations, even if they are not the primary focus for reducing drying intensity.

*Final Interpretation*
The PC1-based deficit analysis confirms that drying risk is primarily driven by a shared exposure gradient, with shading as the dominant control. However, the variable-level breakdown reveals that different ponds reach high-risk conditions through different combinations of factors, highlighting the need for a management approach that combines system-wide shading strategies with pond-specific interventions targeting the dominant sources of exposure.


Filtering to only the dry-out ponds
```{r}
dryout_ponds <- model_df$pond[model_df$drought_class == "dryout"]

deficits_dryout <- deficits_df[deficits_df$pond %in% dryout_ponds, ]
```

Plot
```{r}
library(reshape2)

deficits_dryout_long <- melt(deficits_dryout, id.vars = "pond")
```

```{r}
library(ggplot2)

ggplot(deficits_dryout_long, aes(x = variable, y = value)) +
  geom_col(fill = "firebrick") +
  facet_wrap(~ pond) +
  labs(y = "Deficit Contribution", x = "Variable",
       title = "PC1 Deficits for Dryout Ponds") +
  theme_minimal() + theme(
    axis.text.x = element_text(angle = 60, hjust = 1, size = 8),
    strip.text = element_text(size = 10)
  )
```

# Pond-Specific Deficit Interpretation (Dryout Ponds)
These dryout ponds do not share a single pathway to high drying risk. Instead, they reach similar outcomes through different combinations of deficits along the same PC1 gradient. That is, dryout ponds are not uniformly driven by the same deficit. While reduced shading is the dominant system-wide mechanism, individual ponds exhibit distinct deficit profiles, including structural canopy limitations, reduced realized shading, compound exposure effects, and, in some cases, capacity constraints. This indicates that effective management should prioritize shading at the system level while tailoring interventions to the dominant deficit mechanisms within each pond.


*BOO*

BOO shows low overall deficit, with small contributions from canopy and thermal timing, and a moderate contribution from depth.

This is not a strongly exposed pond in the same way as others. Its position on the drying side of PC1 appears to be driven more by limited basin capacity than by extreme shading deficits. In other words, BOO may not be highly exposed, but it lacks sufficient storage to buffer water loss.

Interpretation:
Drying here is likely capacity-limited rather than exposure-driven. Increasing shading may help, but depth constraints are likely a key factor.

*CSN*

CSN shows a dominant deficit from july_msf, with a secondary contribution from vegetation change.

This is a clear signal of low realized shading, even if structural canopy may not be extremely low. The vegetation contribution reinforces that this pond is dynamically responding under exposed conditions.

Interpretation:
CSN is primarily a realized shade deficit system. Management should focus on improving functional shading during peak summer, not just canopy presence.

*JCO*

JCO shows moderate contributions from july_msf and depth, with minimal influence from other variables.

This suggests a mixed mechanism, where both reduced shading and limited storage contribute to drying. Neither factor is extreme on its own, but together they push the pond toward higher risk.

Interpretation:
JCO represents a combined exposure and capacity limitation, requiring a balanced approach to both shading and storage considerations.

*MSP*

MSP stands out with a very large canopy deficit, far exceeding other variables, along with smaller contributions from MSF, thermal timing, and depth.

This is the clearest example of a structural exposure-driven system. The dominant issue is lack of canopy cover, which likely drives increased solar input and downstream drying processes.

Interpretation:
MSP is strongly canopy-limited, making it highly responsive to interventions that increase structural shading. This is a high-priority candidate for canopy enhancement.

*QED*

QED shows distributed deficits across multiple variables, including MSF, vegetation change, thermal timing, and some canopy contribution.

No single variable dominates, but the combined effect is substantial. This indicates a compound exposure system, where multiple moderate deficits align rather than a single extreme driver.

Interpretation:
QED is a multi-factor exposure system, requiring integrated management rather than a single targeted intervention. Shading improvements will help, but vegetation dynamics and thermal conditions also need to be considered.



#### *Total deficit by pond*

Compare overall deficit between groups:
```{r}
aggregate(pc1_deficit ~ drought_class, data = model_df, mean)
```

And visualize:
```{r}
library(ggplot2)

ggplot(model_df, aes(x = drought_class, y = pc1_deficit, fill = drought_class)) +
  geom_boxplot() +
  geom_jitter(width = 0.1) +
  theme_minimal()
```
Refill ponds are not stable because of their environment; they are stable because of intervention. Some refill ponds have deficit values as high or higher than dryout ponds. Refill ponds occupy a similar and in some cases more exposed position along the PC1-derived deficit gradient compared to dryout ponds, indicating that they share underlying environmental conditions associated with rapid water loss. This suggests that their current persistence is likely maintained by artificial refilling rather than inherent resistance to drying. In this context, refill ponds can be interpreted as systems with latent drying risk, where environmental drivers of exposure, particularly reduced shading, are comparable to those observed in dryout ponds. This highlights the extent to which management intervention may be compensating for otherwise unfavorable conditions and underscores the importance of distinguishing between environmentally stable systems and those that are maintained through external inputs.

Comparing refill vs. dryout directly
```{r}
aggregate(pc1_deficit ~ drought_class, data = model_df, mean)
```

*Refill ponds are, on average, more exposed than dryout ponds.* Group-level comparison of PC1-based deficit scores revealed that refill ponds exhibited the highest average deficit (2.84), followed by dryout ponds (2.65), with stable ponds showing the lowest values (2.39). This pattern indicates that refill ponds occupy the most exposed positions along the primary environmental gradient associated with drying rate, characterized by reduced canopy cover, lower realized shading, and greater vegetation-driven exposure. Despite this, refill ponds do not consistently exhibit complete drying, suggesting that their persistence is maintained through artificial water inputs rather than inherent environmental buffering. In contrast, stable ponds are associated with lower deficit values, reflecting greater shading and reduced exposure. These results highlight that PC1 captures underlying drying pressure, while observed pond outcomes reflect the interaction between environmental conditions and management intervention.

*Approach 1: Predict “Expected Drying Pressure”*

```{r}
names(model_df)
```

Use the earlier regression model to generate predicted z_loss_rate for refill ponds specifically.
```{r}
model <- lm(z_loss_rate ~ PC1, data = model_df)

model_df$predicted_loss <- predict(model, newdata = model_df)
```

Then look at refill ponds:
```{r}
subset(model_df, drought_class == "refill")[, c("pond", "PC1", "pc1_deficit", "predicted_loss")]
```

Rank refill ponds by vulnerability
```{r}
refill_df <- subset(model_df, drought_class == "refill")

refill_df <- refill_df[order(refill_df$predicted_loss), ]

refill_df[, c("pond", "PC1", "pc1_deficit", "predicted_loss")]
```

```{r}
cor(model_df$PC1, model_df$predicted_loss)
```

Interpretation: Refill ponds span the full range of the exposure gradient, from conditions comparable to stable ponds (BGL) to conditions exceeding those of dryout ponds (BGS). Some refill ponds exhibit environmental conditions consistent with rapid drying, suggesting their persistence is dependent on continued water inputs.

Refill ponds exhibited substantial variation in predicted drying pressure, spanning a wide range of positions along the PC1 exposure gradient. Predicted drying values ranged from -0.59 in BGL (PC1 = 1.42; deficit = 1.21), indicating low exposure and relatively low drying pressure, to 0.67 in BGS (PC1 = -1.61; deficit = 4.24), representing the highest exposure and strongest predicted drying pressure among refill ponds. Intermediate systems such as RUP (predicted = -0.26; deficit = 2.00) and GVI (predicted = 0.17; deficit = 3.03) occupied transitional positions, while MCC (predicted = 0.46; deficit = 3.74) aligned closely with high-risk conditions. Notably, the group mean PC1 deficit for refill ponds (2.84) exceeded that of dryout ponds (2.65) and stable ponds (2.39), indicating that refill ponds, on average, occupy more exposed positions along the primary gradient controlling drying rate. Together, these results demonstrate that refill ponds are not environmentally uniform, but instead span a continuum from low-risk systems to highly exposed, intervention-dependent systems, with some ponds exhibiting predicted drying pressures comparable to or exceeding those observed in dryout ponds.


*Approach 2: Compare refill ponds to dryout range*
```{r}
subset_df <- subset(model_df, drought_class %in% c("dryout", "refill"))

ggplot(subset_df, aes(x = drought_class, y = pc1_deficit, fill = drought_class)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.1, size = 2) +
  theme_minimal()
```

Get dryout range:
```{r}
dryout_range <- range(model_df$pc1_deficit[model_df$drought_class == "dryout"])
dryout_range
```

Plot
```{r}
ggplot(model_df, aes(x = drought_class, y = pc1_deficit)) +
  geom_jitter(width = 0.1, size = 2) +
  geom_hline(yintercept = dryout_range[1], linetype = "dashed") +
  geom_hline(yintercept = dryout_range[2], linetype = "dashed") +
  theme_minimal()
```

Identify refill ponds overlapping dryout range:
```{r}
refill_df <- subset(model_df, drought_class == "refill")

refill_df$within_dryout_range <- refill_df$pc1_deficit >= dryout_range[1] & 
                                refill_df$pc1_deficit <= dryout_range[2]

refill_df
```

Density plot
```{r}
ggplot(model_df, aes(x = pc1_deficit, fill = drought_class)) +
  geom_density(alpha = 0.4) +
  theme_minimal()
```

Interpretation: 
This anchors refill ponds directly against the observed environmental envelope of dryout ponds. Comparison of refill ponds to the observed dryout range of PC1-based deficit values (1.99 to 3.00) indicates substantial overlap in environmental conditions associated with drying. Two refill ponds, RUP (1.996) and GVI (3.031), fall directly within or at the boundary of the dryout range, while MCC (3.744) and BGS (4.240) exceed the upper bound, indicating exposure levels greater than those observed in any dryout pond. Only BGL (1.211) falls clearly below the dryout range, suggesting comparatively lower vulnerability. This pattern demonstrates that 4 out of 5 refill ponds (80%) occupy environmental conditions consistent with or more extreme than those associated with dryout, despite not exhibiting complete drying. These results reinforce that refill ponds are not environmentally distinct from dryout systems, but instead frequently experience comparable or elevated exposure along the primary gradient controlling drying rate. The persistence of these ponds therefore likely reflects the influence of artificial water inputs rather than inherent resistance to drying, with several refill ponds representing highly exposed, intervention-dependent systems.



*Approach 3: Rank refill ponds by vulnerability*
```{r}
refill_df <- subset(model_df, drought_class == "refill")

refill_ranked <- refill_df[order(-refill_df$pc1_deficit), 
                           c("pond", "PC1", "pc1_deficit", "predicted_loss")]

refill_ranked
```
Since we already established dryout range = 1.99 to 3.00. Ties refill ponds to dryout range.
```{r}
dryout_range <- c(1.989573, 3.004020)

refill_ranked$vulnerability <- ifelse(
  refill_ranked$pc1_deficit > dryout_range[2], "Exceeds dryout",
  ifelse(refill_ranked$pc1_deficit >= dryout_range[1], "Within dryout", "Below dryout")
)

refill_ranked
```
Interpretation: 

Among refill ponds, there is a clear gradient in latent drying vulnerability once you remove artificial inputs. Three ponds fall well into the “would likely dry out” space: BGS (PC1 = –1.61, predicted loss = 0.67), MCC (–1.12, 0.46), and GVI (–0.40, 0.17). These are not marginal cases. Their predicted loss values are positive and relatively high, meaning their environmental conditions align closely with your observed dry-out ponds. BGS in particular sits at the extreme end, with the largest PC1 deficit (4.24), indicating a substantial mismatch between its current state and the stability threshold.

RUP (PC1 = 0.63, predicted loss = –0.26) sits near the boundary. It falls within the dry-out envelope but does not strongly exceed it, suggesting it may be conditionally stable and sensitive to moderate shifts in inputs or climate.

BGL (PC1 = 1.42, predicted loss = –0.59) is the only refill pond clearly positioned in a stable region. Its positive PC1 score and strongly negative predicted loss indicate environmental conditions that are inherently resistant to drying, even without supplementation.

The key point is this: refill ponds are not a homogeneous group. Your results show a hidden vulnerability structure, where some ponds are effectively being kept wet by management rather than environmental resilience. If inputs were removed, at least half of these systems would likely behave more like dry-out ponds, with BGS and MCC being the most at risk.


*Interpretation of the 3 methods*
Here, they align remarkably well, and that is what gives your result real weight.

Across all three approaches, refill ponds consistently separate into two distinct groups: those that are structurally vulnerable to drying and those that are inherently stable regardless of inputs. The regression-based predictions show that BGS (predicted loss = 0.67), MCC (0.46), and GVI (0.17) all fall within or exceed the dry-out range, indicating that their environmental conditions closely match those of observed dry-out ponds. The PC1 deficit analysis reinforces this, with these same ponds showing the largest deviations from the stability threshold (4.24, 3.74, and 3.03, respectively), confirming that they are not just statistically similar to dry-out ponds, but meaningfully displaced along the dominant environmental gradient driving water loss.

The ranking approach makes this pattern unavoidable. When ordered by vulnerability, the same three ponds consistently occupy the highest-risk positions, followed by RUP, which sits near the threshold (predicted loss = –0.26; deficit = 2.00) and appears conditionally stable. Only BGL (predicted loss = –0.59; deficit = 1.21) remains clearly separated as a low-risk system, with environmental characteristics that place it firmly outside the dry-out space.

What matters here is the convergence. These are not three independent answers producing noise. They are three different lenses on the same underlying gradient, and they all identify the same structure. That tells you the signal is strong.

The broader implication is hard to ignore. Refill ponds are not a coherent ecological category. Instead, they mask a continuum of drying vulnerability, where some ponds are only maintained through artificial inputs while others are fundamentally resilient. Your analysis shows that if those inputs were removed, a substantial portion of refill ponds would likely transition into the dry-out class, with BGS and MCC representing the most extreme cases of hidden vulnerability.

#########

ALL PONDS RANKING

########

```{r}
# Rank all ponds by predicted drying risk
ranked_df <- model_df[order(-model_df$predicted_loss), ]

# Add a rank column
ranked_df$dryout_rank <- 1:nrow(ranked_df)

# View results
ranked_df[, c("dryout_rank", "pond", "drought_class", "PC1", "predicted_loss")]
```

Add risk categories
```{r}
ranked_df$risk_category <- cut(
  ranked_df$predicted_loss,
  breaks = c(-Inf, -0.3, 0, 0.3, Inf),
  labels = c("Low", "Moderate", "High", "Extreme")
)

ranked_df[, c("dryout_rank", "pond", "drought_class", 
              "PC1", "pc1_deficit", "predicted_loss", "risk_category")]
```

Importantly: This is a relative risk ranking based on environmental similarity, not a deterministic forecast.

*Recap: How Risk Was Quantified*

Dryout risk was quantified using the primary environmental gradient identified through PCA (PC1), which represents a continuum from high shading and lower exposure to low shading and increased exposure. Because regression analysis showed that PC1 significantly predicts drying rate, this axis was used as the foundation for all risk assessments.

Risk was operationalized in two linked ways. First, PC1 deficit quantified how far each pond deviates from the most stable conditions along this gradient, with higher values indicating greater exposure and reduced shading. Second, predicted drying pressure was calculated using the regression relationship between PC1 and standardized water loss rate (z_loss_rate), allowing each pond to be assigned an expected drying intensity based on its environmental position.

Dryout ponds provide a critical reference for interpreting this gradient. Their PC1 deficit values define the observed environmental range associated with actual drying, allowing other ponds to be evaluated relative to known outcomes. Ponds that fall within or exceed this range can therefore be interpreted as experiencing comparable or greater underlying drying pressure, regardless of their current classification.

*Ordered Ranking of Ponds by Dryout Risk*

From highest to lowest vulnerability:

Extreme Risk
VAN (stable) – highest deficit (5.86), highest predicted loss (1.34)
BGS (refill) – deficit 4.24, predicted 0.67
MCC (refill) – deficit 3.74, predicted 0.46

High Risk
GVI (refill) – deficit 3.03, predicted 0.17
MSP (dryout) – deficit 3.00, predicted 0.16
QED (dryout) – deficit 2.95, predicted 0.13
CSN (dryout) – deficit 2.90, predicted 0.11

Moderate Risk
TRT (stable) – deficit 2.57
JCO (dryout) – deficit 2.42
RUP (refill) – deficit 2.00
BOO (dryout) – deficit 1.99
CHS (stable) – deficit 1.98

Low Risk
SCT (stable) – deficit 1.53
BGL (refill) – deficit 1.21
AVL (stable) – lowest deficit (0.00), lowest predicted loss (-1.09)

*Final Interpretation of Ranking System*

The ranking reveals that dryout risk is not confined to ponds that currently dry, but instead spans all drought classes along a continuous environmental gradient. Notably, the highest-risk pond (VAN) is classified as stable, while two refill ponds (BGS and MCC) rank above all observed dryout systems. This indicates that current pond states do not always reflect underlying environmental vulnerability, and that management interventions or short-term conditions may mask latent drying risk.

Dryout ponds cluster within the high-risk portion of the gradient, reinforcing that PC1 captures the dominant environmental conditions associated with observed drying. However, the presence of refill and stable ponds within this same range demonstrates that similar environmental conditions can lead to different outcomes depending on external inputs or system-specific characteristics.

Overall, these results confirm that pond drying operates along a shared exposure-driven gradient, where shading and energy balance control underlying drying pressure, while factors such as water inputs and system capacity influence whether that pressure results in complete drying. This highlights the importance of distinguishing between observed state and environmental vulnerability, as ponds that appear stable may still occupy high-risk positions and require proactive management to prevent future drying under changing conditions.



###########

#DISCUSSION
Synthesis of Techniques

#########

The combined analyses indicate that urban ponds do not form discrete functional types, but instead occupy positions along continuous environmental gradients shaped by interacting structural, thermal, and hydrological conditions. While dryout, stable, and refill classifications describe observed outcomes, these categories overlap substantially in environmental space, indicating that similar environmental configurations can produce different hydrologic responses depending on how multiple factors interact .

Principal Component Analysis (PCA) identified a dominant gradient associated with structural and realized shading (PC1), which emerged as the primary control on drying intensity. Regression of PCA scores shows that ponds with greater canopy cover and higher midsummer shade persistence consistently exhibit slower rates of water loss, while more exposed systems dry more rapidly. This establishes solar exposure as the central mechanism governing the rate at which ponds lose water across the study system.

In contrast, secondary gradients, particularly those associated with basin capacity, vegetation dynamics, and thermal timing (PC2), do not independently explain variation in drying rate but instead influence broader system behavior. Discriminant analysis indicates that these gradients are more strongly associated with differences among dryout, stable, and refill ponds, suggesting that while exposure determines how quickly water is lost, system-level characteristics influence how that loss translates into overall outcomes, such as persistence, seasonal stability, or complete dry out.

Cluster analysis further supports this interpretation by identifying broad groupings of slower-drying, intermediate, and faster-drying systems, while also demonstrating substantial overlap among hydrologic classes within these clusters. Ponds classified as dryout, stable, and refill are distributed across similar regions of environmental space rather than forming distinct groupings, reinforcing that drying dynamics are structured along continuous gradients rather than discrete categories. The clusters therefore reflect general regions of environmental similarity, rather than fixed ecological types.

Together, these results indicate that pond drying is governed by a shared underlying mechanism, solar exposure and shading, with additional environmental dimensions shaping variability in system response. This supports a gradient-based interpretation in which all ponds are subject to the same fundamental controls, but differ in how those controls interact with basin structure, vegetation dynamics, and thermal behavior to produce different outcomes. Within this framework, observed pond classes reflect positions along a continuum of drying vulnerability rather than fundamentally distinct system types.

*Management Implications*

From a management perspective, these findings support a tiered, process-based approach that distinguishes between dominant system-wide controls and secondary, site-specific modifiers. At the system level, efforts should prioritize reducing solar exposure, as structural and realized shading consistently emerge as the strongest predictors of reduced drying rates. Increasing canopy cover, preserving surrounding vegetation, or enhancing shading around ponds are therefore likely to be the most effective strategies for moderating water loss under current climatic conditions.

At the same time, pond-specific characteristics such as depth, vegetation dynamics, and thermal behavior influence how individual systems respond within this broader framework. Deeper ponds may be more resistant to complete dry out despite similar exposure conditions, while vegetation dynamics appear to play a more complex role, with increased vegetation change associated with more exposed and dynamically drying systems rather than consistently buffering against water loss. These factors help explain why ponds with similar positions along the primary exposure gradient may still differ in whether they dry out, remain stable, or require supplemental inputs.

Accordingly, effective management should integrate system-wide strategies targeting dominant controls on solar exposure with targeted, site-specific adjustments that account for basin capacity, vegetation behavior, and local thermal conditions. This approach recognizes that while drying rate is largely governed by a common mechanism, variation in outcomes reflects the interaction of multiple environmental factors operating within each system.

*Incorporate risk systems, merge everything together in more detail here*


***Describe next steps***

- PCA
- Regression
- Cluster
- MDA


###########

#CONCLUSION

###########

*short, 2-paragraph outline*





############

#REFERENCES

############






##################

#APPENDIX

#################

Original set of response variables: 

*Landscape*
- Impervious surface (%) within 50 m and 100 m buffers
- Canopy cover (%) within 50 m and 100 m buffers

*Microclimate*
- Air temperature (°C): monthly summaries
- Water temperature (°C): monthly summaries
- Derived thermal metrics (e.g., water–air temperature difference, mean monthly warming duration [in hours] from - minimum to maximum daily values)

*Shade Metrics (LiDAR-derived)*
- Mean Shade Duration (MSD)
- Shade Coverage Percentage (SCP)
- Moderate Shade Fraction (MSF)
- High Shade Fraction (HSF)
- Shade Evenness Index (SEI)

*Vegetation & Water Chemistry*
- Vegetation cover (%) (grid-based summaries, in addition to monthly and seasonal changes)
pH changes


*1. Removal of thermal response variables*

Summary: july_water_air was excluded because it behaved as a downstream thermal response rather than an independent environmental driver. It reflects the outcome of shading, exposure, and pond structure, so including it in the PCA would risk collapsing the ordination into a generalized hot versus cool gradient rather than revealing the underlying mechanisms structuring ponds.

- daily_loss_est
- total_depth_loss_cm
- total_volume_lost_m3
- mean_daily_volume_lost_m3
- z_loss_rate
- z_total_loss


*2. Selection of SEI over other shade metrics*

Summary: alternative shade metrics such as MSD, SCP, and MSF were tested, but these were strongly correlated with temperature-related variables and behaved more like composite or response-like indices. july_sei was retained because it showed more moderate relationships and captured the spatial distribution of shade across pond surfaces rather than simply overall shade magnitude or persistence. This made it more interpretable and less redundant within the PCA.

```{r, eval=FALSE}
shade_compare <- metrics_wide_2 %>%
  dplyr::select(
    canopy_50,
    july_water_air,
    av_hrs_water_july_mintomax,
    july_msd,
    july_scp,
    july_msf,
    july_sei
  )

round(cor(shade_compare, use = "pairwise.complete.obs"), 2)
```

*3. Vegetation variable refinement*

Summary: veg_percent_change was initially considered, but it showed instability and was less suitable for PCA because extreme values could disproportionately influence ordination structure. It was replaced with veg_mean_cover_july, which better represented vegetation state at peak summer conditions and provided a more stable, interpretable measure of biological structure.

```{r, eval=FALSE}
veg_compare <- metrics_wide_2 %>%
  dplyr::select(
    veg_percent_change,
    veg_mean_cover_july,
    slope_per_day
  )

summary(veg_compare)
```

*4. Transformation of growth dynamics*

Summary: slope_per_day was retained as a measure of vegetation change through time, but it showed a right-skewed distribution. To reduce leverage from a few high values, it was transformed using log1p(), producing slope_per_day_log. This allowed the variable to remain in the PCA without overly dominating component structure.

```{r, eval=FALSE}
metrics_wide_2 <- metrics_wide_2 %>%
  dplyr::mutate(
    slope_per_day_log = log1p(slope_per_day)
  )

summary(metrics_wide_2$slope_per_day_log)
boxplot(metrics_wide_2$slope_per_day_log)
```

*5. Final conceptual grouping of candidate variables*

Summary: The final exploratory PCA candidate set was organized to represent distinct ecological dimensions rather than a cluster of statistically convenient variables. This grouping helped ensure that the PCA captured multiple complementary gradients in pond structure and function.

Landscape structure: canopy_50, imp_100
Thermal mediation: july_sei, av_hrs_water_july_mintomax
Biological and chemical state: veg_mean_cover_july, slope_per_day_log, mean_season_ph

This step is important because it shows that variable selection was driven by ecological interpretation as well as statistical screening.
```{r, eval=FALSE}
pca_df <- metrics_wide_2 %>%
  dplyr::select(
    canopy_50,
    imp_100,
    av_hrs_water_july_mintomax,
    july_sei,
    veg_mean_cover_july,
    slope_per_day_log,
    mean_season_ph
  )

names(pca_df)
summary(pca_df)
```

Together, these steps show that variable selection was not based on simple trial and error, but on an iterative process balancing redundancy, ecological meaning, and interpretability. Downstream thermal response variables were removed to reduce circularity, SEI was selected as the most defensible realized shade metric, unstable vegetation change metrics were replaced with more robust summer state variables, and skewed growth measures were transformed before ordination. The resulting variable set was intentionally structured to represent landscape, thermal, and biological dimensions of pond systems rather than a single dominant process.

*6. Motivation: Estimating Pond Storage Capacity*

Summary: A derived variable was explored to approximate pond storage capacity, with the goal of capturing a pond’s ability to buffer against seasonal water loss. This was motivated by the idea that drying dynamics are not solely driven by environmental exposure, but also by inherent physical capacity. This is a proxy, not a direct measurement of volume or bathymetry, and assumes depth_1 is representative of basin structure.

Key idea: 
- Dry-out is a function of both inputs/outputs and storage
- PCA variables so far mostly represent flux drivers, not capacity

Code (representative, not exhaustive)
```{r, eval=FALSE}
metrics_wide_2 <- metrics_wide_2 %>%
  mutate(storage_proxy = depth_1 * area)
```

*7. Evaluation and Rejection of Storage Proxy*

Summary: The storage proxy was evaluated but ultimately excluded from PCA due to conceptual and statistical limitations.

Reasons:
- Strongly correlated with pond size → risks dominating PCA as a “size axis”
- Does not capture dynamic hydrology (inflow, seepage, evapotranspiration)
- Based on a single depth measure, which may not represent basin morphology

*8. Evaluation and Exclusion of Impervious Surface Metrics*

Impervious surface cover was initially considered as a proxy for surrounding urban intensity, with the expectation that higher imperviousness would amplify thermal exposure, alter runoff dynamics, and contribute to increased drying rates in urban ponds.

To evaluate its contribution, impervious cover was tested at two spatial scales:

imp_50 (local influence)
imp_100 (broader landscape context)

Two parallel PCA configurations were constructed to assess whether impervious surface variables contributed meaningful explanatory structure.


Code (representative)
```{r, eval=FALSE}
# Define variable sets
vars_no_imp <- c(
  "canopy_50",
  "av_hrs_air_july_mintomax",
  "july_msf",
  "veg_percent_change",
  "depth_1"
)

vars_with_imp <- c(vars_no_imp, "imp_100")

# Run PCA
pca_no_imp  <- prcomp(metrics_wide_2[vars_no_imp], scale. = TRUE)
pca_with_imp <- prcomp(metrics_wide_2[vars_with_imp], scale. = TRUE)

# Compare structure
summary(pca_no_imp)
summary(pca_with_imp)

round(pca_no_imp$rotation, 2)
round(pca_with_imp$rotation, 2)

# Compare scores
scores_no_imp  <- as.data.frame(pca_no_imp$x)
scores_with_imp <- as.data.frame(pca_with_imp$x)

cor(scores_no_imp$PC1, scores_with_imp$PC1)
cor(scores_no_imp$PC2, scores_with_imp$PC2)
```


*Final Synthesis*

All exploratory testing and iterative refinement converged on a final set of variables that balance ecological interpretability with statistical independence. Early stages of variable selection revealed that many candidate predictors were either redundant, overly correlated, or reflective of downstream responses rather than independent drivers. Through successive removal of thermal outcome variables, consolidation of shade metrics, and refinement of vegetation and growth indicators, the analysis shifted toward representing distinct ecological mechanisms rather than overlapping signals. Additional exploratory configurations, including the incorporation of storage proxies and alternative thermal and vegetation variables, further highlighted the importance of separating structural capacity, environmental forcing, and biological response.

The final selection of canopy_50, av_hrs_air_july_mintomax, july_msf, veg_percent_change, and depth_1 reflects this process, with each variable contributing uniquely to the representation of landscape structure, thermal dynamics, realized shading, seasonal biological change, and physical storage capacity. By emphasizing July-based metrics, the model isolates peak environmental forcing while avoiding late-season feedback effects, ensuring that the PCA captures underlying drivers of drying rather than cumulative outcomes. Together, this variable set provides a parsimonious yet mechanistically grounded framework for interpreting variation in pond drying dynamics across sites, with minimal redundancy and clear ecological meaning.
