---
title: "Final Project Proposal"
author: 'Group 3: Brandon Weigand, Sandra Inusah'
date: "2026-03-21"
output:
  word_document: default
  html_document: default
instructor: Dr. Suborna Ahmed
---

```{r echo=FALSE}
#install.packages("dplyr")
#install.packages("ggplot2")
#install.packages("tidyr")
#install.packages("lubridate")
#install.packages("slider")
#install.packages("stringr")
#install.packages("dendextend")
#library(dplyr)
#library(ggplot2)
#library(tidyr)
#library(lubridate)
#library(stringr)
#library(slider)
#library(dendextend)
#library(tidyverse)
```


## Title: Drivers of Urban Pond Dry-Out: Identifying Environmental Gradients and Patterns of Water Loss in Pond Habitats Across Vancouver

## General Description

In response to growing pressures on freshwater resources, the City of Vancouver recently implemented water conservation strategies through its 2017 and 2023 Water Priority Action Plans. These policies prioritize long-term water security by reducing potable water use across urban infrastructure, including fountains, spray parks, and ornamental water features. As part of this transition, several aesthetic ponds that were historically maintained through regular potable water inputs have been reclassified as seasonal systems and no longer receive supplemental water. Consequently, ponds that once retained stable water levels throughout the year now undergo substantial seasonal drawdown, with some transitioning from water-filled habitats in winter to dry basins by late summer.

This policy-driven shift creates a rare opportunity to examine urban pond dynamics under reduced human intervention, allowing processes of water loss and persistence to emerge more clearly than in artificially maintained systems. Despite their small size, urban freshwater ponds play a critical role in moderating microclimates and supporting biodiversity, particularly during the summer months when aquatic and semi-aquatic species experience peak resource demand. However, the mechanisms governing seasonal water loss in these systems remain poorly understood. In particular, it is unclear how interacting biophysical factors including temperature, nutrient availability, vegetation growth, canopy structure, and shading influence the rate and extent of water decline across ponds embedded within the same urban landscape.

## Leading Research Question

Which environmental gradients differentiate urban ponds that dry out from those that remain stable or receive supplemental water?

## Secondary Research Questions:

Which biophysical variables contribute most strongly to the environmental gradients associated with variation in rates of pond water loss?

How do environmental gradients and pond conditions shift over the summer season, and how do these shifts relate to differences in rates of water loss among ponds?

Do ponds with different rates of water loss form distinct clusters or patterns, or do they fall along a continuous gradient of environmental conditions?

## Hypothesis

Ponds that experience higher rates of water loss display distinct patterns characterized by higher temperatures, reduced shading, and greater water loss rates.

## Motivation

This research is motivated by the need to better understand how urban freshwater ecosystems respond to increasing climatic variability and urban pressures. As ponds provide critical habitat for aquatic and semi-aquatic species, their seasonal loss can have cascading ecological impacts. By identifying the key environmental factors and patterns associated with pond persistence or dry-out, this study aims to clarify the mechanisms of urban pond water loss and inform habitat management strategies for enhancing pond resilience in Vancouver and other cities facing similar water conservation pressures. In doing so, it provides an early diagnosis of urban pond vulnerability under a future in which reduced water inputs and stricter management policies become increasingly common.

## Data Overview

## 1. Study Area

This study was conducted across 15 urban freshwater ponds in the City of Vancouver, British Columbia, Canada, during the June–August 2025 summer period, when precipitation is typically low and evaporative demand is high. Vancouver has a temperate coastal climate, characterized by mild temperatures, dry summers, and wet winters, making summer months particularly important for understanding water balance and drought dynamics in shallow urban water bodies with uneven annual climate regimes.

The ponds include: Avalon Pond (AVL), Botanical Garden Large (BGL), Botanical Garden Small (BGS), Cariboo Park (BOO), Charles Park (CHS), Charleson Park (CSN), Granville Island (GVI), Jericho Beach Park (JCO), McCleery Golf Course (MCC), Memorial South Park (MSP), Queen Elizabeth Park Duck Pond (QED), Rupert Putt Course Pond (RUP), Sanctuary Pond (SCT), Trout Lake (TRT), and Vanier Park (VAN).

The ponds vary substantially in size, depth, and hydrological behavior, ranging from small, shallow systems that frequently dry out to larger, managed ponds that remain stable or are periodically refilled. Sites were intentionally selected to capture variation along key gradients, including pond area, canopy cover, and spatial distance across the urban matrix, while also representing contrasting hydrological conditions. Of the fifteen ponds, five experienced complete or near-complete dry-out during the 2024 summer season, whereas five others maintained consistent water levels due to continued inputs and exemption from water restriction policies, with the remaining ponds exhibiting intermediate behavior. This distribution enables direct comparison across a spectrum of water persistence conditions. All ponds were delineated using high-resolution aerial imagery and field validation, representing their maximum early-season water extent.

The selected ponds are distributed across a range of urban land use contexts, including public parks (e.g., Trout Lake, Queen Elizabeth Park), recreational areas (e.g., Jericho Beach, Vanier Park), institutional landscapes (e.g., UBC Botanical Garden), and highly built environments (e.g., Granville Island). This spatial diversity captures variation in surrounding land cover, including impervious surfaces, within-pond vegetation structure, and canopy density, which are hypothesized to influence microclimate and water loss processes.

The study area is embedded within a highly urbanized landscape, where anthropogenic factors such as heat retention from built surfaces, altered hydrology, and fragmented vegetation interact with natural processes. These conditions create strong spatial gradients in shading, temperature, and moisture availability.

## 2. Data Source

The study integrates field-based measurements, in-situ sensor data, and spatial datasets to examine patterns associated with excelerated rates of urban pond water loss.

Water level data were collected through repeated field visits, where pond depth was measured using meter sticks at fixed sampling locations (minimum of six visits per pond, roughly 3 weeks between visits). 

Vegetation cover was estimated using a grid-based visual survey approach during each site visit, where 2 observers independently assessed percentage cover within predefined 20x20 meter grid cells and averaged their estimates. Water quality variables (pH and phosphate) were measured in the field using high-sensitivity test strips and vial-based assays.

Microclimate data (air temperature, water temperature, and humidity) were recorded using iButton loggers (MAXIM Inc.) deployed at each pond. These sensors recorded measurements at 20-minute intervals throughout the study period. Loggers were installed at centroid locations within each pond. 

For shade data, ArcGIS Pro was used to model within-pond shading dynamics using a LiDAR-derived workflow. Digital elevation (DEM), surface (DSM), and canopy height (CHM) models were generated from high-resolution LiDAR data and used to represent terrain and vegetation structure. Solar position (azimuth and altitude) was calculated at hourly intervals on June 21, July 21, and August 21 between 07:00 and 18:00, and hillshade rasters were produced for each time step to simulate incident solar radiation. These outputs were reclassified into binary shade and sun exposure layers and aggregated to derive spatially explicit metrics of shade intensity, coverage, and distribution. Based on these outputs, areas within each pond were categorized into high shade, intermediate, and high exposure conditions to capture daily shade variability.

Spatial predictor variables were derived from geospatial datasets. Impervious surface and canopy cover were calculated within 50 m and 100 m buffers using a classified City of Vancouver land cover raster (4.8 m resolution) and processed in R using the landscapemetrics package.

All raw and cleaned data are available at: <https://docs.google.com/spreadsheets/d/1INgRR1rHrhaoGNrlkq3hGot1tx46EbdfggESUael5dQ/edit?gid=0#gid=0>.



## 3. Sample Size

The dataset was structured hierarchically across both spatial and temporal scales. Fifteen ponds served as the primary sampling units, each characterized by site-level attributes including pond area, hydrological condition, and surrounding land cover. Land cover metrics were obtained once prior to the field season from the 2022 Vancouver land cover dataset and treated as static landscape variables. Within each pond, paired air and water temperature loggers recorded microclimatic conditions at 20-minute intervals for a minimum of approximately 70 days, although deployment length and continuity varied somewhat among sites due to occasional logger failure. Additional field-based measurements, including vegetation extent, water recession, and water quality variables, were collected at roughly three-week intervals throughout the summer. Monthly shade metrics were also derived for each pond to represent seasonal changes in canopy influence and solar exposure.

## 4. Variables and Measurement Scales

## Predictor Variables
Predictor variables captured landscape context, microclimate, shading, and within-pond conditions. Landscape variables included impervious surface and canopy cover within 50 m and 100 m buffers. Microclimate variables were derived from high-frequency logger data and summarized as monthly air and water temperatures, and thermal metrics such as water–air temperature offsets and daily warming duration. Shade metrics were derived from LiDAR-based modeling and included mean shade intensity, shade coverage, temporal persistence of shade, high-shade extent, and shade evenness. Finally, field-based measurements captured within-pond conditions, including vegetation cover and water chemistry (e.g., pH and phosphate), collected at regular intervals throughout the summer.

a.  Landscape
1.  Impervious surface (%) within 50 m and 100 m buffers
2.  Canopy cover (%) within 50 m and 100 m buffers

b.  Microclimate
1.  Air temperature (°C): monthly summaries
2.  Water temperature (°C): monthly summaries
3.  Derived thermal metrics (e.g., water–air temperature difference, mean monthly warming duration [in hours] from minimum to maximum daily values)

c.  Shade Metrics (LiDAR-derived)
1.  Mean Shade Duration (MSD)
2.  Shade Coverage Percentage (SCP)
3.  Moderate Shade Fraction (MSF)
4.  High Shade Fraction (HSF)
5.  Shade Evenness Index (SEI)

d.  Vegetation & Water Chemistry
1.  Vegetation cover (%) (grid-based summaries, in addition to monthly and seasonal changes)
2.  pH changes

## Response Variables
Response variables quantified both the magnitude and rate of water loss, including total seasonal water loss (cm), daily water loss (cm/day), total and mean daily volume loss (m³), and standardized indices of drying intensity and cumulative loss (Z-scores).

1.  Total seasonal water loss (cm)
2.  Daily water loss rate (cm/day)
3.  Total volume lost (m³) and mean daily volume loss (m³/day)
4.  Standardized metrics (Z-scores): drying rate and total loss

## 5. Known Issues (Missingness, Outliers, and Limitations)

1.  Missing water and air temperature data occurred due to sensor failures and/or field logistics, resulting in gaps of approximately 1–3 weeks in some ponds; these periods were recorded but not interpolated. 
2.  Water input sources from precipitation and runoff are not yet represented in the dataset. 
2.  Temperature data were cleaned to remove outliers using jump filters, physical thresholds (e.g., values exceeding 40 °C), and a rolling median with MAD-based filtering. *See appendix for a description of the cleaning rules.* In total, about 0.5% of values were not interpolated. 
3.  Because sensors were deployed at different times, timestamps did not always align and were standardized by rounding forward to the nearest matching interval, using water temperature as the primary reference.  
4.  Some measurement assumptions may also introduce uncertainty, as water volume loss was estimated assuming constant pond area despite some ponds losing water surface area over time, and depth measurements were collected at discrete intervals that may not capture short-term fluctuations. 
5.  The land cover dataset used (~4.8 m resolution) may miss fine-scale variation, although classes were simplified to reduce classification noise. 
6.  Shade metrics are derived from the same set of raster "stacks" per month and are therefore highly correlated between each other, which limits their combined use in multivariate analyses.


## Method

1. PCA Analysis

To address the research questions, Principal Component Analysis (PCA) and cluster analysis will be used as complementary statistical approaches. Prior to both analyses, the environmental predictor variables, including temperature, shade metrics, and landscape cover percentages, will be standardized to ensure comparability across units and scales. PCA will then be applied to these variables to reduce dimensionality and summarize them into a smaller set of orthogonal components representing the dominant environmental gradients among ponds. PCA will assess all ponds simultaneously, treating each pond as a single observation within the dataset. 

The PCA approach is preferred in order to reduce the large set of correlated environmental variables into a smaller number of independent components that represent dominant environmental gradients among ponds. Many of the measured variables, including temperature, canopy cover, and the shade metrics, are expected to covary, making it difficult to interpret their individual effects in isolation. PCA would address this concern by combining related variables into composite axes that capture the main patterns of biophysical variations in the dataset.

Each pond will receive a score on these components in PCA, indicating its position along these gradients. For example, one component may represent a gradient from cool, shaded, high-canopy ponds to warm, exposed ponds with low canopy cover, while another may reflect differences in surrounding landscape cover, like increasing impervious surface or reduced vegetation. In other cases, the components may capture gradients related to variations in pH or nutrient concentrations, depending on how these variables covary across ponds. These components would be derived directly from the structure of the data and would be interpreted based on the contribution of the original environmental variables.

The interpretation of PCA results will depend on which variables load most strongly onto each component. For instance, if the first principal component is dominated by temperature and shade variables, this would indicate that thermal and radiation-related conditions are the primary drivers differentiating ponds. Alternatively, if a component is driven by canopy cover and impervious surface, this would suggest that broader landscape structure plays a stronger role. Or, if water pH loads strongly onto a component, this may indicate that internal pond conditions or nutrient dynamics are important in interpreting differences among sites. By examining how ponds are distributed along these components, the analysis will reveal which combinations of environmental conditions define the major gradients across the system. 

Water loss metrics, including daily drying rates and total seasonal loss, will then be compared to these component scores to assess how drying behavior aligns with these environmental gradients. For instance, if ponds with higher drying rates consistently have high scores along a component associated with warm temperatures and low shade, this would suggest that these conditions contribute to faster water loss. Similarly, if total seasonal loss is associated with a gradient of high impervious surface and reduced canopy cover, this would indicate a broader landscape influence on pond persistence.

Importantly, the variables describing water loss, including daily drying rates and total seasonal loss, will not be included in the PCA itself. Instead, PCA will be based solely on the predictor variables. After the PCA is completed, scores for each pond on the main principal component axes will be extracted. These scores will indicate where each pond falls along the identified dominant environmental gradients. Water loss variables will then be compared with these PCA scores using correlation analysis to determine whether ponds with faster or greater water loss are consistently associated with particular gradients. For example, correlations will test whether higher drying rates tend to occur in ponds scoring more strongly along axes associated with warmer temperatures, reduced canopy cover, or lower shade. 

2. Cluster Analysis 

Cluster analysis will be used alongside PCA to group ponds based on overall similarity in their environmental conditions. While PCA identifies the main environmental gradients across ponds, cluster analysis asks a different question: do ponds with similar conditions naturally group together into distinct types, or do they instead vary gradually along a continuum? In this way, PCA describes how ponds differ, whereas cluster analysis evaluates whether those differences form clear groupings.

To do this, ponds will be grouped based on how similar they are across all standardized environmental variables, including temperature, shade, and landscape characteristics. A hierarchical clustering approach will be used, which builds groups step by step by first linking the most similar ponds and then progressively combining them into larger clusters. The results will be visualized using a dendrogram, which shows how ponds are related to one another and where natural groupings may occur.

Once clusters are identified, they will be compared with observed rates of water loss and pond classifications. This will help determine whether ponds with similar drying behavior also share similar environmental conditions. For example, if ponds that dry quickly are consistently grouped together, this would suggest that certain combinations of conditions, such as higher temperatures or reduced shading, are associated with rapid water loss. In contrast, if ponds with different drying rates are spread across multiple clusters, this would suggest that water loss varies more gradually across environmental conditions rather than forming distinct pond types.

Cluster results will also be interpreted alongside the PCA outputs. This combined approach allows assessment of whether environmental variation among ponds is best understood as a set of distinct groups or as a continuous gradient. Rather than serving as a test of the PCA, cluster analysis provides an additional way to interpret the structure of the system and to better understand how patterns of water loss relate to underlying environmental conditions.

Cluster analysis is particularly relevant to the third research question, which asks whether ponds with different rates of water loss form distinct groups or fall along a continuous environmental gradient. While PCA identifies the dominant gradients structuring variation among ponds, it does not determine whether these gradients produce discrete categories of pond conditions. Cluster analysis addresses this by evaluating whether ponds with similar environmental characteristics, and potentially similar drying behavior, group together into identifiable clusters. This allows assessment of whether rapidly drying ponds represent a distinct pattern or simply occupy one end of a broader environmental spectrum. In this way, clustering provides a good link between the identified environmental gradients and the actual structure of pond drying patterns.

The interpretation of cluster analysis will depend on how clearly ponds group together and how those groupings relate to both environmental gradients and water loss patterns identified in the PCA. For example, if cluster analysis reveals distinct groups of ponds and these groups align with patterns observed in the PCA, such as one group occupying warmer, more exposed positions along a temperature and shading gradient, this would suggest that ponds can be categorized into identifiable types with shared environmental conditions and drying behavior. In this case, rapidly drying ponds may form a distinct cluster characterized by high temperatures, low canopy cover, and reduced shading. Alternatively, if cluster analysis does not reveal clear groupings, or if ponds with different drying rates are distributed across clusters, this would indicate that environmental variation and water loss are better understood as part of a continuous gradient rather than discrete categories. In this scenario, PCA results would still identify the dominant environmental drivers, but cluster analysis would suggest that ponds transition more gradually along these gradients, rather than forming distinct types. Comparing both approaches will therefore help to determine whether patterns of pond drying reflect clearly defined environmental groupings or more gradual shifts in conditions across pond sites. Perhaps most importantly, these two analyses might reveal patterns inherent to ponds within urban environments. 


# Hypothetical Plots & Interpretations

```{r}
library(tidyverse)
```


## PCA
```{r setup-mock-ponds, echo=FALSE, message=FALSE, warning=FALSE}
set.seed(123)

# ----------------------------------------
# Create pond identities and classes
# ----------------------------------------
ponds <- tibble(
  pond_id = paste0("Pond_", str_pad(1:15, width = 2, pad = "0")),
  pond_group = c(rep("Dry", 5), rep("Stable", 5), rep("Supplemented", 5))
)

months <- tibble(
  month = factor(c("June", "July", "August"), levels = c("June", "July", "August")),
  month_num = c(6, 7, 8),
  seasonal_heat = c(0, 1.2, 2.3),
  seasonal_loss = c(0, 0.8, 1.8)
)

# ----------------------------------------
# Simulate pond x month environmental data
# ----------------------------------------
mock_pond_data <- expand_grid(ponds, months) %>%
  mutate(
    # Landscape context
    impervious_100 = case_when(
      pond_group == "Dry" ~ rnorm(n(), mean = 58, sd = 8),
      pond_group == "Stable" ~ rnorm(n(), mean = 28, sd = 7),
      pond_group == "Supplemented" ~ rnorm(n(), mean = 45, sd = 10)
    ),
    
    canopy_100 = case_when(
      pond_group == "Dry" ~ rnorm(n(), mean = 28, sd = 8),
      pond_group == "Stable" ~ rnorm(n(), mean = 72, sd = 7),
      pond_group == "Supplemented" ~ rnorm(n(), mean = 42, sd = 10)
    ),
    
    # Shade metrics
    SCP = case_when(
      pond_group == "Dry" ~ rnorm(n(), mean = 42, sd = 10),
      pond_group == "Stable" ~ rnorm(n(), mean = 81, sd = 8),
      pond_group == "Supplemented" ~ rnorm(n(), mean = 58, sd = 10)
    ) - 3 * seasonal_heat + rnorm(n(), 0, 3),
    
    HSF = case_when(
      pond_group == "Dry" ~ rnorm(n(), mean = 0.18, sd = 0.06),
      pond_group == "Stable" ~ rnorm(n(), mean = 0.52, sd = 0.08),
      pond_group == "Supplemented" ~ rnorm(n(), mean = 0.30, sd = 0.08)
    ) - 0.03 * seasonal_heat + rnorm(n(), 0, 0.02),
    
    SEI = case_when(
      pond_group == "Dry" ~ rnorm(n(), mean = 0.9, sd = 0.18),
      pond_group == "Stable" ~ rnorm(n(), mean = 2.0, sd = 0.30),
      pond_group == "Supplemented" ~ rnorm(n(), mean = 1.4, sd = 0.25)
    ),
    
    # Microclimate
    mean_air_temp = case_when(
      pond_group == "Dry" ~ rnorm(n(), mean = 24.5, sd = 1.2),
      pond_group == "Stable" ~ rnorm(n(), mean = 20.8, sd = 1.1),
      pond_group == "Supplemented" ~ rnorm(n(), mean = 23.0, sd = 1.3)
    ) + seasonal_heat,
    
    mean_water_temp = case_when(
      pond_group == "Dry" ~ rnorm(n(), mean = 22.5, sd = 1.0),
      pond_group == "Stable" ~ rnorm(n(), mean = 18.6, sd = 0.9),
      pond_group == "Supplemented" ~ rnorm(n(), mean = 21.2, sd = 1.0)
    ) + 0.8 * seasonal_heat - 0.015 * SCP + rnorm(n(), 0, 0.4),
    
    warming_duration_hrs = case_when(
      pond_group == "Dry" ~ rnorm(n(), mean = 9.0, sd = 0.8),
      pond_group == "Stable" ~ rnorm(n(), mean = 6.7, sd = 0.7),
      pond_group == "Supplemented" ~ rnorm(n(), mean = 8.0, sd = 0.8)
    ) + 0.35 * seasonal_heat,
    
    thermal_offset = mean_water_temp - mean_air_temp,
    
    # Within pond conditions
    vegetation_cover = case_when(
      pond_group == "Dry" ~ rnorm(n(), mean = 22, sd = 8),
      pond_group == "Stable" ~ rnorm(n(), mean = 61, sd = 12),
      pond_group == "Supplemented" ~ rnorm(n(), mean = 38, sd = 10)
    ),
    
    pH = case_when(
      pond_group == "Dry" ~ rnorm(n(), mean = 8.0, sd = 0.25),
      pond_group == "Stable" ~ rnorm(n(), mean = 7.1, sd = 0.20),
      pond_group == "Supplemented" ~ rnorm(n(), mean = 7.5, sd = 0.25)
    ),
    
    # Response variables
    daily_water_loss_cm = case_when(
      pond_group == "Dry" ~ rnorm(n(), mean = 0.95, sd = 0.15),
      pond_group == "Stable" ~ rnorm(n(), mean = 0.28, sd = 0.09),
      pond_group == "Supplemented" ~ rnorm(n(), mean = 0.42, sd = 0.10)
    ) +
      0.035 * (mean_air_temp - 20) -
      0.004 * SCP +
      0.012 * impervious_100 +
      seasonal_loss / 10 +
      rnorm(n(), 0, 0.04),
    
    total_volume_loss_m3 = daily_water_loss_cm * rnorm(n(), mean = 38, sd = 8)
  ) %>%
  mutate(
    across(
      c(impervious_100, canopy_100, SCP, vegetation_cover),
      ~ pmin(pmax(., 0), 100)
    ),
    HSF = pmin(pmax(HSF, 0), 1),
    daily_water_loss_cm = pmax(daily_water_loss_cm, 0.02)
  )

```

```{r pca-plot, echo=FALSE, fig.cap="Mock PCA of pond environmental gradients", fig.width=7, fig.height=5}

mock_pca_input <- mock_pond_data %>%
  group_by(pond_id, pond_group) %>%
  summarise(
    impervious_100 = mean(impervious_100),
    canopy_100 = mean(canopy_100),
    mean_air_temp = mean(mean_air_temp),
    mean_water_temp = mean(mean_water_temp),
    thermal_offset = mean(thermal_offset),
    warming_duration_hrs = mean(warming_duration_hrs),
    SCP = mean(SCP),
    HSF = mean(HSF),
    SEI = mean(SEI),
    vegetation_cover = mean(vegetation_cover),
    pH = mean(pH),
    daily_water_loss_cm = mean(daily_water_loss_cm),
    .groups = "drop"
  )

pca_vars <- mock_pca_input %>%
  select(
    impervious_100, canopy_100, mean_air_temp, mean_water_temp,
    thermal_offset, warming_duration_hrs, SCP, HSF, SEI,
    vegetation_cover, pH, daily_water_loss_cm
  )

mock_pca <- prcomp(pca_vars, scale. = TRUE)

scores <- as_tibble(mock_pca$x[, 1:2]) %>%
  bind_cols(mock_pca_input %>% select(pond_id, pond_group))

loadings <- as_tibble(mock_pca$rotation[, 1:2], rownames = "variable") %>%
  mutate(
    PC1 = PC1 * 3,
    PC2 = PC2 * 3
  )

ggplot(scores, aes(x = PC1, y = PC2, color = pond_group)) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(aes(label = pond_id), size = 3, show.legend = FALSE) +
  geom_segment(
    data = loadings,
    aes(x = 0, y = 0, xend = PC1, yend = PC2),
    inherit.aes = FALSE,
    arrow = arrow(length = unit(0.18, "cm"))
  ) +
  geom_text(
    data = loadings,
    aes(x = PC1, y = PC2, label = variable),
    inherit.aes = FALSE,
    size = 3,
    vjust = -0.4
  ) +
  theme_minimal() +
  labs(
    title = "Mock PCA of pond environmental gradients",
    subtitle = "Points represent ponds; arrows represent variable loadings",
    x = "PC1",
    y = "PC2",
    color = "Pond group"
  )
```
## Interpretation: 
The mock PCA shows a clear environmental gradient along PC1, separating ponds with higher water loss from those that remain stable. “Dry” ponds cluster on the negative end of the axis and are associated with higher temperatures, longer warming durations, greater impervious surface, and increased water loss. In contrast, “stable” ponds group on the positive end and align with higher canopy cover, greater shading, and increased vegetation.

The second axis (PC2) appears to capture secondary variation related to thermal dynamics, such as thermal offset and water temperature, and primarily differentiates ponds within groups rather than between them.

Supplemented ponds occupy an intermediate position and partially overlap with both dry and stable ponds, suggesting that pond conditions may fall along a continuous environmental gradient rather than forming discrete clusters.

## Cluster Analysis 

```{r mock-dendrogram, echo=FALSE, fig.cap="Mock hierarchical clustering of ponds based on environmental variables", fig.width=8, fig.height=4}

# Use the same PCA input dataset
dist_mat <- dist(scale(pca_vars))

hc <- hclust(dist_mat, method = "ward.D2")

plot(
  hc,
  labels = mock_pca_input$pond_id,
  main = "Mock hierarchical clustering of ponds",
  xlab = "Pond",
  ylab = "Dissimilarity",
  sub = ""
)

# Draw 3 clusters (matching your hypothesis groups)
rect.hclust(hc, k = 3, border = 2:4)

```

## Interpretation: 
The mock dendrogram shows a clear grouping of ponds into three main clusters based on overall environmental similarity. Ponds with higher water loss group together, reflecting shared conditions such as elevated temperatures, reduced shading, and greater impervious influence. Stable ponds form a separate cluster characterized by cooler, more shaded, and vegetated environments. A third cluster captures intermediate conditions, corresponding to supplemented ponds.

Together, the PCA and cluster analysis suggest that while broad groupings of ponds can be identified, these groups are structured along underlying gradients driven primarily by temperature, shading, and urban landscape context, which in turn relate to differences in water loss.


## Assumptions and Checks

Several assumptions will be evaluated to ensure the methods are appropriate for the dataset and research question:

-    Variable scaling
All variables will be standardized (e.g., using z-scores) to ensure comparability and prevent variables with larger magnitudes from dominating the analysis.

-    Linearity (for PCA)
PCA assumes that relationships among the variables are approximately linear. This assumption is reasonable for expected relationships such as temperature, shading, and canopy effects, but will be evaluated using scatterplots and correlation matrices to ensure that major patterns are not strongly non-linear.

-    Multicollinearity / redundancy
Many variables in the dataset, particularly shade metrics and temperature summaries, are expected to be highly correlated. While PCA can accommodate correlated variables, excessive redundancy can complicate interpretation of component loadings. For example, multiple shade metrics such as Mean Shade Duration (MSD), Shade Coverage Percentage (SCP), and High Shade Fraction (HSF) may all capture similar aspects of shading and therefore show strong correlations with one another. Including all of these variables may cause them to load strongly onto the same component, making it difficult to interpret what that component represents beyond general “shade.” Similarly, monthly temperature variables (e.g., June, July, and August water temperature) may be highly correlated and reflect a common thermal signal. Correlation matrices will be used to identify strongly correlated variables (e.g., >0.8), and redundant variables may be reduced or grouped (e.g., selecting a representative month or metric) to retain key ecological signals while improving interpretability.

-    Sample size and variable selection
With 15 ponds, the sample size is relatively small compared to the number of variables. To address this, we will reduce the number of variables and focus on key predictors (e.g., selecting representative months such as July or key metrics) to ensure stable results.

-   Missing data
PCA does not handle missing values well. Observations with excessive missing data will be removed or carefully handled. Missing data are limited but will be checked prior to analysis.

-   Outliers
Both PCA and cluster analysis are sensitive to outliers. Although major outliers were removed during the data cleaning process, variables like water loss and pond size will be checked to ensure no extreme values influence results.

-   Distance metric and clustering method (CA)
Cluster analysis requires selecting an appropriate distance metric and linkage method. Euclidean distance and Ward’s method will likely be used, since they minimize within-cluster variance and produce more interpretable groupings. Alternative methods may be tested. 

-  Meaningful variable selection
Only variables that represent meaningful ecological processes will be included. Redundant or highly correlated variables will be excluded, and temporal variables may be simplified (e.g., focusing on peak summer conditions).

-  Interpretation and linkage to response variables
Because water loss variables are not included in the PCA, their relationship to environmental gradients will be evaluated separately through correlation and visualization. Because water loss variables are not included in the PCA, their relationship to environmental gradients will be evaluated separately through correlation and visualization. This avoids circularity, which would occur if water loss variables were included in the PCA and then later used to interpret the same components. In other words, the PCA would be partially built from the phenomenon we're trying to explain, making it unclear whether environmental variables are actually explaining variation in drying or simply reflecting it. By excluding water loss from the PCA and comparing it to the resulting components afterward, the analysis maintains a clear distinction between environmental predictors and drying responses, allowing for more meaningful interpretation of how environmental conditions relate to observed water loss patterns.


## Preliminary work

All data streams have been processed, cleaned, and integrated prior to analysis. All variables were then combined into a single, pond-level dataset, with each pond represented by a consistent set of environmental predictors and water loss metrics. This integrated dataset forms the basis for subsequent analyses.


## Interpretation plan

To address the research question—what environmental gradients best differentiate ponds that dry out from those that remain stable or are refilled—we will interpret outputs from both PCA and Cluster Analysis.

1. For PCA, we will report:

a.  Scree plots and proportion of variance explained to determine the number of meaningful components. Loadings (variable contributions) for each principal component to identify key environmental gradients (e.g., temperature, shade, canopy cover, land cover)
b.  PCA biplots showing ponds positioned in ordination space, with vectors representing environmental variables

Interpretation will focus on identifying the dominant environmental gradients represented by each component and understanding how ponds are distributed along these gradients. Water loss variables, including daily drying rates and total seasonal loss, will then be compared to PCA results by examining correlations with principal component scores and by visualizing drying rates within the ordination space. This will allow assessment of whether higher rates of water loss are associated with specific environmental gradients. Pond classifications (dry-out, stable, refill) will also be overlaid to support interpretation, but emphasis will be placed on continuous variation in drying behavior rather than categorical differences alone.


2. For Cluster Analysis, we will report:

a.  Dendrograms to visualize how ponds group based on environmental similarity
b.  Cluster groupings for each pond

Interpretation will focus on whether ponds form distinct environmental groupings or instead vary along a continuous gradient. The environmental characteristics of each cluster will be examined to identify shared conditions (e.g., warm and exposed versus cool and shaded ponds). Cluster membership will then be compared with water loss metrics to assess whether ponds with similar drying rates group together. If ponds with faster drying rates consistently fall within the same cluster, this would suggest that distinct combinations of environmental conditions are associated with rapid water loss. Alternatively, if drying rates vary within clusters, this would indicate that water loss occurs along a continuum rather than within clearly defined pond types.

Across both methods, results will be interpreted together to evaluate whether patterns of water loss are best explained by discrete pond types or by continuous environmental gradients. PCA will identify the dominant environmental drivers, while cluster analysis will assess whether these gradients correspond to distinct groupings of ponds. Particular attention will be given to variables related to temperature, shading, canopy structure, and surrounding land cover, as these are hypothesized to significantly influence rates of water loss.

# Expected Group Contributions 

Proposal writing: 
a.  Sandra: wrote base outline, created summaries, described variables, generated base research questions based on inputs. 
b.  Brandon: refined research questions, added datasets, expanded on all sections, created the appendix and tables.  

Data Preparation & Methods
a.  Sandra: Established the base methods, identified the appropriate analysis pipeline, summarized the variables.  
b.  Brandon: Created the data tables, generated workflows, refined analysis pipelines, created the R analysis space.   

Presentation: 
a.  Sandra: Amended and clarified where necessary, worked on scripts and talking points. 
b.  Brandon: Created the base presentation and flow, outlined the study parameters.  

Overall Contributions: 
a.  Sandra: 50%
b.  Brandon: 50%

# Appendix 

## Summary of temperature data cleaning rules
Temperature time series from all ponds were filtered to June through August and cleaned using a multi-step quality control procedure designed to remove implausible values while retaining natural variability. Cleaning was applied independently to each pond and logger combination, with observations ordered chronologically.

## 1. Temporal consistency and jump detection
To identify unrealistic spikes between consecutive measurements, a rate-of-change threshold was applied. Temperature differences between successive observations were evaluated relative to the elapsed time between measurements. Observations were flagged when the rate of change exceeded a scaled threshold equivalent to 5°C per 20-minute interval. This accounts for irregular sampling intervals and removes abrupt, non-physical jumps likely caused by sensor error or direct solar exposure.

## 2. Physical plausibility threshold
A fixed upper bound was applied to remove physically unrealistic air temperature values. Any observation exceeding 40°C was flagged and removed. This threshold reflects conditions that are highly unlikely in the study region and are indicative of sensor exposure or malfunction.

## 3. Local anomaly detection using rolling median and MAD
To detect subtler anomalies within each time series, a rolling window approach was used. For each observation, a trailing window of 71 measurements (corresponding to roughly 24-hours of time) was used to calculate:

- a rolling median representing local central tendency
- a median absolute deviation (MAD) representing local variability

Observations were flagged if they deviated more than 3 MAD from the rolling median. This method is robust to outliers and adapts to changing temperature conditions over time. A minimum of 10 observations was required within each window to compute reliable statistics.

## 4. Combined flagging rule
An observation was classified as invalid if it met any of the following conditions:

- exceeded the rate-of-change threshold
- exceeded the physical temperature cap
- exceeded the rolling MAD threshold

Flagged observations were removed by setting their values to missing, while retaining the original time series structure for subsequent analysis.

```{r}
ponds_raw <- read_csv("ponds_raw_clean.csv")
```

```{r}
# ----------------------------------------
# Parameters
# ----------------------------------------
w <- 71
k <- 3
min_obs <- 10
jump_threshold_20min <- 5
temp_upper_bound <- 40

# ----------------------------------------
# Helper: trailing rolling median/MAD envelope
# ----------------------------------------

calc_mad_envelope_trailing <- function(x, w = 71, min_obs = 10) {
  
  roll_median <- zoo::rollapply(
    x,
    width = w,
    FUN = median,
    fill = NA,
    align = "right",
    partial = min_obs,
    na.rm = TRUE
  )
  
  roll_mad <- zoo::rollapply(
    x,
    width = w,
    FUN = mad,
    fill = NA,
    align = "right",
    partial = min_obs,
    na.rm = TRUE
  )
  
  data.frame(
    roll_median = roll_median,
    roll_mad = roll_mad,
    upper_3mad = roll_median + 3 * roll_mad,
    lower_3mad = roll_median - 3 * roll_mad
  )
}
```

```{r}
# ----------------------------------------
# Air temperature for all ponds, June–August
# ----------------------------------------

allponds_air_summer <- ponds_raw %>%
  filter(
    month(datetime_parsed) %in% c(6, 7, 8)
  ) %>%
  select(
    pond_id,
    datetime = datetime_parsed,
    air_temp_c,
    air_logger_id
  ) %>%
  rename(
    value = air_temp_c,
    logger_id = air_logger_id
  ) %>%
  filter(!is.na(pond_id), !is.na(datetime), !is.na(logger_id), !is.na(value)) %>%
  arrange(pond_id, logger_id, datetime)


```

```{r}
# ----------------------------------------
# Apply air temperature cleaning rules across all ponds
# ----------------------------------------

allponds_air_summer_flagged <- allponds_air_summer %>%
  group_by(pond_id, logger_id) %>%
  arrange(datetime, .by_group = TRUE) %>%
  mutate(
    dt = as.numeric(difftime(datetime, lag(datetime), units = "mins")),
    
    # Jump rule
    flag_jump =
      !is.na(value) &
      !is.na(lag(value)) &
      !is.na(dt) &
      dt > 0 &
      abs(value - lag(value)) > (jump_threshold_20min / 20) * dt,
    
    # Physical cap
    flag_cap =
      !is.na(value) &
      value > temp_upper_bound
  ) %>%
  group_modify(~{
    df <- .x
    
    env <- calc_mad_envelope_trailing(
      x = df$value,
      w = w,
      min_obs = min_obs
    )
    
    df$roll_median <- env$roll_median
    df$roll_mad <- env$roll_mad
    df$upper_3mad <- env$upper_3mad
    df$lower_3mad <- env$lower_3mad
    
    df$flag_mad <- abs(df$value - df$roll_median) > k * df$roll_mad
    df$flag_mad[is.na(df$flag_mad)] <- FALSE
    
    df
  }) %>%
  ungroup() %>%
  mutate(
    flag_final = flag_jump | flag_cap | flag_mad,
    value_clean = if_else(flag_final, NA_real_, value),
    month_num = month(datetime),
    month = month(datetime, label = TRUE, abbr = FALSE)
  )
```

```{r}
# ----------------------------------------
# Monthly cleaned air temperature averages for all ponds
# ----------------------------------------

allponds_air_monthly_avg <- allponds_air_summer_flagged %>%
  group_by(pond_id, month_num, month) %>%
  summarise(
    mean_air_temp = mean(value_clean, na.rm = TRUE),
    min_air_temp = min(value_clean, na.rm = TRUE),
    max_air_temp = max(value_clean, na.rm = TRUE),
    n_obs = sum(!is.na(value_clean)),
    n_flagged = sum(flag_final, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(pond_id, month_num)

allponds_air_monthly_avg
```


```{r}
# ----------------------------------------
# Full table of flagged entries
# ----------------------------------------
allponds_air_flag_table <- allponds_air_summer_flagged %>%
  filter(flag_final) %>%
  select(
    pond_id,
    datetime,
    logger_id,
    value,
    roll_median,
    roll_mad,
    lower_3mad,
    upper_3mad,
    flag_jump,
    flag_cap,
    flag_mad,
    flag_final
  ) %>%
  arrange(pond_id, datetime)

allponds_air_flag_table
```

```{r}
ggplot(allponds_air_monthly_avg, aes(x = month, y = mean_air_temp, fill = month)) +
  geom_col() +
  facet_wrap(~ pond_id, scales = "free_y") +
  labs(
    title = "Monthly mean air temperature by pond",
    subtitle = "Cleaned values, June to August",
    x = "Month",
    y = "Mean air temperature (°C)"
  ) +
  theme_minimal(base_size = 11)

```

Now for water temperature across all sites for June, July, and August 

```{r}
# ----------------------------------------
# Water temperature for all ponds (June–August)
# ----------------------------------------

allponds_water_summer <- ponds_raw %>%
  filter(
    lubridate::month(datetime_parsed) %in% c(6, 7, 8)
  ) %>%
  select(
    pond_id,
    datetime = datetime_parsed,
    water_temp_c,
    water_logger_id
  ) %>%
  rename(
    value = water_temp_c,
    logger_id = water_logger_id
  ) %>%
  filter(!is.na(pond_id), !is.na(datetime), !is.na(logger_id), !is.na(value)) %>%
  arrange(pond_id, logger_id, datetime)

```

## Water Temperature
```{r}
# ----------------------------------------
# Water temperature for all ponds (June–August)
# ----------------------------------------

allponds_water_summer <- ponds_raw %>%
  filter(
    lubridate::month(datetime_parsed) %in% c(6, 7, 8)
  ) %>%
  select(
    pond_id,
    datetime = datetime_parsed,
    water_temp_c,
    water_logger_id
  ) %>%
  rename(
    value = water_temp_c,
    logger_id = water_logger_id
  ) %>%
  filter(!is.na(pond_id), !is.na(datetime), !is.na(logger_id), !is.na(value)) %>%
  arrange(pond_id, logger_id, datetime)

```

```{r}
# ----------------------------------------
# Apply cleaning rules to water temperature
# ----------------------------------------

allponds_water_summer_flagged <- allponds_water_summer %>%
  group_by(pond_id, logger_id) %>%
  arrange(datetime, .by_group = TRUE) %>%
  mutate(
    dt = as.numeric(difftime(datetime, lag(datetime), units = "mins")),
    
    flag_jump =
      !is.na(value) &
      !is.na(lag(value)) &
      !is.na(dt) &
      dt > 0 &
      abs(value - lag(value)) > (jump_threshold_20min / 20) * dt,
    
    flag_cap =
      !is.na(value) &
      value > temp_upper_bound
  ) %>%
  group_modify(~{
    
    df <- .x
    
    env <- calc_mad_envelope_trailing(
      x = df$value,
      w = w,
      min_obs = min_obs
    )
    
    df$roll_median <- env$roll_median
    df$roll_mad <- env$roll_mad
    
    df$flag_mad <- abs(df$value - df$roll_median) > k * df$roll_mad
    df$flag_mad[is.na(df$flag_mad)] <- FALSE
    
    df
    
  }) %>%
  ungroup() %>%
  mutate(
    flag_final = flag_jump | flag_cap | flag_mad,
    value_clean = if_else(flag_final, NA_real_, value),
    month_num = lubridate::month(datetime),
    month = lubridate::month(datetime, label = TRUE, abbr = FALSE)
  )
```

```{r}
# ----------------------------------------
# Monthly mean water temperature (cleaned)
# ----------------------------------------

allponds_water_monthly_avg <- allponds_water_summer_flagged %>%
  group_by(pond_id, month_num, month) %>%
  summarise(
    mean_water_temp = mean(value_clean, na.rm = TRUE),
    
    n_obs = n(),
    n_clean = sum(!is.na(value_clean)),
    n_flagged = sum(flag_final),
    
    .groups = "drop"
  ) %>%
  arrange(pond_id, month_num)

allponds_water_monthly_avg

```

## Plots

```{r}
ggplot(allponds_air_monthly_avg,
       aes(x = month_num, y = mean_air_temp, group = pond_id)) +
  geom_line(alpha = 0.6) +
  geom_point(size = 2) +
  facet_wrap(~ pond_id, scales = "free_y") +
  scale_x_continuous(
    breaks = c(6, 7, 8),
    labels = c("June", "July", "August")
  ) +
  labs(
    title = "Monthly mean air temperature by pond",
    subtitle = "Cleaned values for June to August",
    x = "Month",
    y = "Mean air temperature (°C)"
  ) +
  theme_minimal(base_size = 11)
```
```{r}
ggplot(allponds_water_monthly_avg,
       aes(x = month_num, y = mean_water_temp, group = pond_id)) +
  geom_line(alpha = 0.6) +
  geom_point(size = 2) +
  facet_wrap(~ pond_id, scales = "free_y") +
  scale_x_continuous(
    breaks = c(6, 7, 8),
    labels = c("June", "July", "August")
  ) +
  labs(
    title = "Monthly mean water temperature by pond",
    subtitle = "Cleaned values for June to August",
    x = "Month",
    y = "Mean water temperature (°C)"
  ) +
  theme_minimal(base_size = 11)
```
```{r}
air_plot_df <- allponds_air_monthly_avg %>%
  select(pond_id, month_num, month, mean_temp = mean_air_temp) %>%
  mutate(source = "Air")

water_plot_df <- allponds_water_monthly_avg %>%
  select(pond_id, month_num, month, mean_temp = mean_water_temp) %>%
  mutate(source = "Water")

temp_compare_df <- bind_rows(air_plot_df, water_plot_df)

ggplot(temp_compare_df,
       aes(x = month_num, y = mean_temp, color = source, group = source)) +
  geom_line() +
  geom_point(size = 2) +
  facet_wrap(~ pond_id, scales = "free_y") +
  scale_x_continuous(
    breaks = c(6, 7, 8),
    labels = c("June", "July", "August")
  ) +
  labs(
    title = "Monthly air and water temperatures by pond",
    subtitle = "Comparison of cleaned monthly means",
    x = "Month",
    y = "Temperature (°C)",
    color = "Sensor type"
  ) +
  theme_minimal(base_size = 11)
```
## Flagged entries
```{r}
air_flag_summary <- allponds_air_summer_flagged %>%
  group_by(pond_id, month_num, month) %>%
  summarise(
    total_obs = n(),
    flagged_obs = sum(flag_final, na.rm = TRUE),
    flagged_pct = 100 * flagged_obs / total_obs,
    .groups = "drop"
  )

ggplot(air_flag_summary,
       aes(x = month_num, y = flagged_pct, group = pond_id)) +
  geom_line(alpha = 0.7) +
  geom_point(size = 2) +
  facet_wrap(~ pond_id) +
  scale_x_continuous(
    breaks = c(6, 7, 8),
    labels = c("June", "July", "August")
  ) +
  labs(
    title = "Proportion of flagged air temperature records by pond",
    x = "Month",
    y = "Flagged observations (%)"
  ) +
  theme_minimal(base_size = 11)
```
```{r}
water_flag_summary <- allponds_water_summer_flagged %>%
  group_by(pond_id, month_num, month) %>%
  summarise(
    total_obs = n(),
    flagged_obs = sum(flag_final, na.rm = TRUE),
    flagged_pct = 100 * flagged_obs / total_obs,
    .groups = "drop"
  )

ggplot(water_flag_summary,
       aes(x = month_num, y = flagged_pct, group = pond_id)) +
  geom_line(alpha = 0.7) +
  geom_point(size = 2) +
  facet_wrap(~ pond_id) +
  scale_x_continuous(
    breaks = c(6, 7, 8),
    labels = c("June", "July", "August")
  ) +
  labs(
    title = "Proportion of flagged water temperature records by pond",
    x = "Month",
    y = "Flagged observations (%)"
  ) +
  theme_minimal(base_size = 11)
```
## Tables of removed flags with # of entries

```{r}
air_flag_table <- allponds_air_summer_flagged %>%
  group_by(pond_id) %>%
  summarise(
    total_obs = n(),
    flagged_obs = sum(flag_final, na.rm = TRUE),
    retained_obs = total_obs - flagged_obs,
    pct_flagged = 100 * flagged_obs / total_obs,
    pct_retained = 100 * retained_obs / total_obs,
    .groups = "drop"
  ) %>%
  arrange(desc(pct_flagged))

air_flag_table
```

```{r}
water_flag_table <- allponds_water_summer_flagged %>%
  group_by(pond_id) %>%
  summarise(
    total_obs = n(),
    flagged_obs = sum(flag_final, na.rm = TRUE),
    retained_obs = total_obs - flagged_obs,
    pct_flagged = 100 * flagged_obs / total_obs,
    pct_retained = 100 * retained_obs / total_obs,
    .groups = "drop"
  ) %>%
  arrange(desc(pct_flagged))

water_flag_table
```

```{r}
air_flag_monthly <- allponds_air_summer_flagged %>%
  group_by(pond_id, month) %>%
  summarise(
    total_obs = n(),
    flagged_obs = sum(flag_final, na.rm = TRUE),
    retained_obs = total_obs - flagged_obs,
    pct_flagged = 100 * flagged_obs / total_obs,
    .groups = "drop"
  )

air_flag_monthly
```

```{r}
water_flag_monthly <- allponds_water_summer_flagged %>%
  group_by(pond_id, month) %>%
  summarise(
    total_obs = n(),
    flagged_obs = sum(flag_final, na.rm = TRUE),
    retained_obs = total_obs - flagged_obs,
    pct_flagged = 100 * flagged_obs / total_obs,
    .groups = "drop"
  )

water_flag_monthly
```

```{r}
air_flag_table <- air_flag_table %>%
  mutate(sensor = "Air")

water_flag_table <- water_flag_table %>%
  mutate(sensor = "Water")

combined_flag_table <- bind_rows(air_flag_table, water_flag_table)

combined_flag_table
```

## Heatmaps

```{r}
ggplot(allponds_air_monthly_avg,
       aes(x = month, y = pond_id, fill = mean_air_temp)) +
  geom_tile(color = "white") +
  labs(
    title = "Heatmap of monthly mean air temperature",
    x = "Month",
    y = "Pond",
    fill = "Mean air\nTemp (°C)"
  ) +
  theme_minimal(base_size = 11)
```

```{r}
ggplot(allponds_water_monthly_avg,
       aes(x = month, y = pond_id, fill = mean_water_temp)) +
  geom_tile(color = "white") +
  labs(
    title = "Heatmap of monthly mean water temperature",
    x = "Month",
    y = "Pond",
    fill = "Mean water\nTemp (°C)"
  ) +
  theme_minimal(base_size = 11)
```

