---
title: "PondAnalysis, PCA_MDA_Cluster"
author: "Brandon Weigand"
date: "2026-04-03"
output:
  html_document: default
  pdf_document: default
  word_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
metrics_wide_2 <- read.csv("metrics_wide_final.csv")
colnames(metrics_wide_2)
```

######## 

CLUSTER ANALYSIS

######## 

*Clustering Ponds Based on Environmental Conditions/Gradient*

To further investigate the environmental gradients influencing pond
drying, hierarchical cluster analysis using Ward’s method was conducted using the selected predictor
variables. This analysis builds on the variable selection process, which
identified a reduced set of key environmental factors representing
shading, thermal conditions, and urbanization, while minimizing
redundancy and ensuring interpretability.

The cluster analysis was designed to directly address the leading
research question: which environmental gradients differentiate ponds
that dry out from those that remain stable or receive supplemental
water? In addition, it responds to the secondary question of whether
ponds form distinct environmental groupings or instead fall along a
continuous gradient of conditions.

By grouping ponds based on similarities in the selected predictor
variables, the analysis provides a way to identify whether discrete
environmental “types” of ponds exist (e.g., highly urban and exposed vs.
shaded and cooler), or whether variation in pond conditions and drying
rates is more gradual. This approach complements earlier analyses (e.g.,
PCA and redundancy analysis) by translating continuous environmental
gradients into potential group structures that can be more easily
interpreted and compared.

Cluster analysis also allows for evaluation of the hypothesis that ponds
with higher rates of water loss are characterized by warmer
temperatures, reduced shading, and greater exposure to urban conditions.
By comparing drying rates across clusters, it becomes possible to assess
whether these conditions form distinct groups or whether drying
behaviour varies continuously across environmental gradients.

Data preparation

```{r}
# Selecting already chosen predictor variables
selected_vars <- metrics_wide_2[, c(
  "canopy_50",
    "av_hrs_air_july_mintomax",
    "july_msf",
    "veg_percent_change",
    "depth_1"
)]

# Standardize variables
selected_scaled <- scale(selected_vars)
```

Running cluster analysis

```{r}
# create distance matrix
dist_matrix <- dist(selected_scaled , method = "euclidean")
```

```{r}
# Hierarchical clustering (Ward’s method)
hc <- hclust(dist_matrix, method = "ward.D2")
```

```{r}
library(dendextend)

# Convert to dendrogram
dend <- as.dendrogram(hc)

# FIX labels to use pond names
labels(dend) <- metrics_wide_2$pond_id[order.dendrogram(dend)]

# Create named drought class vector
drought_class <- metrics_wide_2$drought_class
names(drought_class) <- metrics_wide_2$pond_id

# Assign colors (your exact palette)
labels_colors(dend) <- ifelse(drought_class[labels(dend)] == "dryout", "#F8766D",
                        ifelse(drought_class[labels(dend)] == "refill", "#00BA38",
                               "#619CFF"))
# Plot
plot(dend,
     main = "Dendrogram of Urban Ponds",
     xlab = "",
     sub = "",
     ylab = "Height")

# Legend
legend("topright",
       legend = c("dryout", "refill", "stable"),
       fill = c("#F8766D", "#00BA38", "#619CFF"),
       border = NA,
       bty = "n")
```

#Interpretation

The hierarchical cluster analysis using Ward’s method revealed three primary clusters of ponds based on similarities in environmental conditions. At a cut height of approximately 5–6, the dendrogram separates the ponds into three main groups. The first cluster consists of a single pond (VAN), which joins the rest of the dendrogram at a high dissimilarity level, indicating that it is environmentally distinct from all other ponds. The second cluster includes BGL, RUP, AVL, SCT, BOO, CHS, GVI, and QED, and represents a group of ponds with moderate similarity across the selected environmental variables. The third cluster contains BGS, MCC, TRT, MSP, CSN, and JCO, which are more tightly grouped at lower heights, indicating greater similarity among these ponds.

Across the dendrogram, drought classes (dryout, refill, stable) are distributed across all clusters rather than forming distinct groupings. Ponds with different drying outcomes are often grouped together, indicating that similar environmental conditions can correspond to different drying behaviours. 

This lack of clear separation from the dendrogram support earlier findings from PCA, regression, and discriminant analysis, showing that ponds are structured along continuous environmental gradients rather than forming distinct types. While the cluster analysis identifies groupings in environmental space, these groupings represent positions along gradients of shading, thermal dynamics, vegetation feedbacks, and basin capacity, rather than fixed ecological states. Overall, while three clusters can be identified, the dendrogram does not show clear separation of ponds based on drought class, suggesting that environmental variation is gradual rather than divided into distinct categories.

```{r}
# Compute WSS for different k
wss <- numeric(10)

for (k in 1:10) {
  kmeans_result <- kmeans(selected_scaled, centers = k, nstart = 25)
  wss[k] <- kmeans_result$tot.withinss
}

# Plot elbow curve
plot(1:10, wss,
     type = "b",
     pch = 19,
     xlab = "Number of clusters (k)",
     ylab = "Total Within-Cluster Sum of Squares (WSS)",
     main = "Elbow Method for Optimal Clusters")
```

#Interpretation

The elbow method was used to evaluate the optimal number of clusters for grouping ponds based on environmental conditions, with the aim of assessing whether ponds form distinct patterns or fall along a continuous gradient of conditions. The total within-cluster sum of squares (WSS) was calculated for values of k ranging from 1 to 10. The results show a sharp decrease in WSS from k = 1 to k = 4, indicating that increasing the number of clusters substantially improves the grouping of ponds within this range. Beyond k = 4, the rate of decrease becomes more gradual, suggesting diminishing returns in reducing within-cluster variation. This pattern indicates that the elbow occurs around k = 4. However, the transition is relatively smooth rather than sharply defined, suggesting that there is no single, clearly optimal number of clusters based on WSS alone.


*Justification for Selecting k = 3 Clusters*

Although the elbow method suggests k = 4 as a potential optimal number of clusters, this solution did not produce groupings that were ecologically meaningful in the context of the study. Specifically, the four-cluster solution resulted in over-partitioning of ponds into smaller groups that did not clearly correspond to differences in environmental gradients or drying behaviour. In contrast, the dendrogram revealed a clear and interpretable separation into three primary clusters, capturing the dominant structure of environmental similarity among ponds. These three clusters align more closely with the study’s objective of identifying environmental gradients that differentiate drying behaviour, while maintaining interpretability in terms of shading, thermal dynamics, vegetation feedbacks, and basin capacity.

Importantly, the three-cluster solution also supports the broader research question of whether ponds form distinct groups or fall along a continuum. The clusters identified at k = 3 show substantial overlap in drought classes, indicating that ponds with different rates of water loss are not cleanly separated into discrete categories. Instead, this pattern reinforces the interpretation that pond drying dynamics are structured along continuous environmental gradients. Therefore, k = 3 was selected as the most appropriate number of clusters, balancing statistical evidence with ecological relevance and consistency with the overall analytical framework.

```{r}
# Cut dendrogram into 3 clusters
metrics_wide_2$Cluster3 <- cutree(hc, k = 3)
```

*Comparing Clusters with Water loss (Z_loss_rate)*

```{r}
# Mean water loss per cluster
aggregate(metrics_wide_2$z_loss_rate,
          by = list(Cluster = metrics_wide_2$Cluster3),
          mean)

# Mean + SD
aggregate(metrics_wide_2$z_loss_rate,
          by = list(Cluster = metrics_wide_2$Cluster3),
          function(x) c(mean = mean(x), sd = sd(x)))

boxplot(z_loss_rate ~ factor(Cluster3,
                             levels = c(1, 2, 3),
                             labels = c("Cluster 1", "Cluster 2", "Cluster 3")),
        data = metrics_wide_2,
        col = c("#F8766D", "#00BA38", "#619CFF"),
        main = "Water Loss (z_loss_rate) Across Clusters",
        xlab = "Cluster",
        ylab = "z_loss_rate")
```

#Interpretation

#Final intrepation for cluster analysis

#How this fit with the PCA and MDA results -

#How this answers the research question and hypothesis.

######### 

Pond-based Risk Axes

######### 

To translate multivariate results into pond-specific management
insights, PC1 was used as a composite index of exposure-related deficit.
Because PC1 represents a gradient of increasing canopy cover and shading
opposed by vegetation-driven exposure, lower PC1 scores indicate
conditions associated with faster drying. Deficit scores were therefore
calculated as the inverse of PC1, allowing ponds to be ranked according
to their relative position along this gradient. To further diagnose the
sources of deficit, variable-level contributions to PC1 were decomposed
using PCA loadings and standardized values. This approach identifies
which factors, such as low canopy cover or high vegetation change,
contribute most strongly to each pond’s position along the exposure
gradient. By linking these deficits directly to the primary axis
controlling drying rate, this method provides a statistically grounded
basis for prioritizing site-specific management interventions.

Standardize PC1 scores into a “deficit” metric.

```{r}
model_df$pc1_deficit_z <- scale(-model_df$PC1)
```

-   Larger values = greater “deficit” relative to optimal conditions
-   Directly tied to your regression result
-   Multivariate by construction

Each variable contributes to PC1 with a loading. So a deficit in a
variable should be interpreted relative to its loading, and its
standardized value.

Method: Signed contribution to PC1 for each pond

```{r}
pca_result <- prcomp(
  metrics_wide_2[, c("canopy_50",
                     "av_hrs_air_july_mintomax",
                     "july_msf",
                     "veg_percent_change",
                     "depth_1")],
  scale. = TRUE
)

scaled_data <- scale(metrics_wide_2[, c("canopy_50",
                                        "av_hrs_air_july_mintomax",
                                        "july_msf",
                                        "veg_percent_change",
                                        "depth_1")])

loadings <- pca_result$rotation[, "PC1"]

loadings

```

Calculate contributions

```{r}
contributions <- sweep(scaled_data, 2, loadings, "*")
```

Calculate deficits

```{r}
deficits <- ifelse(contributions < 0, abs(contributions), 0)
deficits_df <- as.data.frame(deficits)
```

Add pond names

```{r}
deficits_df$pond <- metrics_wide_2$pond

#add overall PC1 deficit score
model_df$pc1_deficit <- max(model_df$PC1) - model_df$PC1
```

Visual

```{r}
#install.packages("reshape2")
library(reshape2)
library(ggplot2)

deficits_long <- melt(deficits_df, id.vars = "pond")

ggplot(deficits_long, aes(x = variable, y = value, fill = pond)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(y = "Deficit Contribution", x = "Variable") +
  theme_minimal()
```

Each bar represents:

How much that variable is pulling a pond toward the “high drying risk”
side of PC1

Important points:

Values are only shown when they contribute negatively to PC1 Larger bars
= stronger contribution to drying risk Zero means that variable is not
contributing to risk for that pond

## Interpretation of Pond-Based Risk Axes

*Structure of the PC1 Deficit Metric*

The PC1-based deficit metric provides a pond-specific measure of
exposure-related risk, directly linked to the primary gradient
controlling drying rate. Because PC1 is defined by positive
contributions from canopy cover (0.646) and July mean shade fraction
(0.546), and a negative contribution from vegetation percent change
(-0.486), ponds with low PC1 scores are characterized by reduced shading
and increased vegetation-driven exposure. Depth (0.139) and thermal
timing (0.169) contribute only weakly to this axis.

By inverting PC1, higher deficit scores therefore represent ponds that
are more exposed, less shaded, and more dynamically changing, conditions
associated with faster drying in the regression analysis. Decomposing
PC1 into variable-level contributions allows identification of which
specific factors are driving this exposure for each pond.

*Dominant Drivers of Risk Across Ponds*

Across ponds, the largest deficit contributions are consistently
associated with canopy cover, July mean shade fraction, and vegetation
percent change, reflecting the dominant structure of PC1. In most cases,
high risk is driven by a combination of low canopy cover and reduced
realized shading, often accompanied by positive vegetation change, which
reinforces exposure rather than mitigating it.

Canopy cover emerges as the most consistent contributor to deficit, with
several ponds showing large negative contributions along this variable.
July MSF shows a similar pattern, indicating that both structural and
realized shading are aligned in driving exposure-related risk. This
reinforces the earlier PCA result that these variables operate together
as a shared shading gradient.

Vegetation percent change contributes strongly to deficit in a subset of
ponds, particularly where large positive values coincide with low
shading. Because this variable loads negatively on PC1, increases in
vegetation are associated with movement toward the high-risk end of the
gradient. This suggests that vegetation dynamics are not acting as a
buffering mechanism, but instead are linked to exposed, rapidly changing
conditions, potentially reflecting shallow water expansion or ecological
responses to drying.

*Secondary and Minor Contributors*

In contrast, depth and thermal timing contribute relatively little to
PC1-based deficit across most ponds. While some ponds show moderate
contributions from these variables, their overall influence is small
compared to shading-related factors. This aligns with the regression
results, which showed that these variables do not independently control
drying rate.

However, their presence in some ponds suggests that they may still
influence system behavior in combination with other factors,
particularly in shaping broader drought outcomes rather than directly
controlling drying intensity.

*Pond-Specific Risk Profiles*

The deficit decomposition highlights that while ponds share a common
exposure gradient, the sources of risk differ among ponds. Some ponds
are primarily driven by low canopy cover, others by reduced realized
shading, and others by strong vegetation change. In several cases,
multiple factors act together, producing compounded exposure.

This reinforces the interpretation that ponds occupy different positions
along a shared environmental continuum, but arrive there through
different combinations of underlying conditions. As a result, two ponds
with similar overall deficit scores may require different management
interventions depending on which variables are driving their position
along PC1.

*Management-Relevant Interpretation*

These results provide a direct link between the statistical analysis and
actionable management priorities. Because PC1 is the dominant predictor
of drying rate, variables contributing to PC1 deficit represent the most
immediate targets for intervention.

The consistent importance of canopy cover and realized shading indicates
that increasing shading is the most effective system-wide strategy for
reducing drying risk. This may involve enhancing surrounding canopy,
modifying vegetation structure, or otherwise increasing shade
persistence during peak summer conditions.

At the same time, the role of vegetation change suggests that vegetation
should be managed strategically rather than assumed to provide passive
protection. In some ponds, increasing vegetation may actually coincide
with greater exposure and faster drying, indicating a need for more
targeted management of plant dynamics.

Although depth and thermal timing are not primary drivers of drying
rate, they may still influence how ponds respond to exposure,
particularly in terms of persistence and resilience. As such, they
remain relevant for site-specific considerations, even if they are not
the primary focus for reducing drying intensity.

*Final Interpretation* The PC1-based deficit analysis confirms that
drying risk is primarily driven by a shared exposure gradient, with
shading as the dominant control. However, the variable-level breakdown
reveals that different ponds reach high-risk conditions through
different combinations of factors, highlighting the need for a
management approach that combines system-wide shading strategies with
pond-specific interventions targeting the dominant sources of exposure.

Filtering to only the dry-out ponds

```{r}
dryout_ponds <- model_df$pond[model_df$drought_class == "dryout"]

deficits_dryout <- deficits_df[deficits_df$pond %in% dryout_ponds, ]
```

Plot

```{r}
library(reshape2)

deficits_dryout_long <- melt(deficits_dryout, id.vars = "pond")
```

```{r}
library(ggplot2)

ggplot(deficits_dryout_long, aes(x = variable, y = value)) +
  geom_col(fill = "firebrick") +
  facet_wrap(~ pond) +
  labs(y = "Deficit Contribution", x = "Variable",
       title = "PC1 Deficits for Dryout Ponds") +
  theme_minimal() + theme(
    axis.text.x = element_text(angle = 60, hjust = 1, size = 8),
    strip.text = element_text(size = 10)
  )
```

# Pond-Specific Deficit Interpretation (Dryout Ponds)

These dryout ponds do not share a single pathway to high drying risk.
Instead, they reach similar outcomes through different combinations of
deficits along the same PC1 gradient. That is, dryout ponds are not
uniformly driven by the same deficit. While reduced shading is the
dominant system-wide mechanism, individual ponds exhibit distinct
deficit profiles, including structural canopy limitations, reduced
realized shading, compound exposure effects, and, in some cases,
capacity constraints. This indicates that effective management should
prioritize shading at the system level while tailoring interventions to
the dominant deficit mechanisms within each pond.

*BOO*

BOO shows low overall deficit, with small contributions from canopy and
thermal timing, and a moderate contribution from depth.

This is not a strongly exposed pond in the same way as others. Its
position on the drying side of PC1 appears to be driven more by limited
basin capacity than by extreme shading deficits. In other words, BOO may
not be highly exposed, but it lacks sufficient storage to buffer water
loss.

Interpretation: Drying here is likely capacity-limited rather than
exposure-driven. Increasing shading may help, but depth constraints are
likely a key factor.

*CSN*

CSN shows a dominant deficit from july_msf, with a secondary
contribution from vegetation change.

This is a clear signal of low realized shading, even if structural
canopy may not be extremely low. The vegetation contribution reinforces
that this pond is dynamically responding under exposed conditions.

Interpretation: CSN is primarily a realized shade deficit system.
Management should focus on improving functional shading during peak
summer, not just canopy presence.

*JCO*

JCO shows moderate contributions from july_msf and depth, with minimal
influence from other variables.

This suggests a mixed mechanism, where both reduced shading and limited
storage contribute to drying. Neither factor is extreme on its own, but
together they push the pond toward higher risk.

Interpretation: JCO represents a combined exposure and capacity
limitation, requiring a balanced approach to both shading and storage
considerations.

*MSP*

MSP stands out with a very large canopy deficit, far exceeding other
variables, along with smaller contributions from MSF, thermal timing,
and depth.

This is the clearest example of a structural exposure-driven system. The
dominant issue is lack of canopy cover, which likely drives increased
solar input and downstream drying processes.

Interpretation: MSP is strongly canopy-limited, making it highly
responsive to interventions that increase structural shading. This is a
high-priority candidate for canopy enhancement.

*QED*

QED shows distributed deficits across multiple variables, including MSF,
vegetation change, thermal timing, and some canopy contribution.

No single variable dominates, but the combined effect is substantial.
This indicates a compound exposure system, where multiple moderate
deficits align rather than a single extreme driver.

Interpretation: QED is a multi-factor exposure system, requiring
integrated management rather than a single targeted intervention.
Shading improvements will help, but vegetation dynamics and thermal
conditions also need to be considered.

#### *Total deficit by pond*

Compare overall deficit between groups:

```{r}
aggregate(pc1_deficit ~ drought_class, data = model_df, mean)
```

And visualize:

```{r}
library(ggplot2)

ggplot(model_df, aes(x = drought_class, y = pc1_deficit, fill = drought_class)) +
  geom_boxplot() +
  geom_jitter(width = 0.1) +
  theme_minimal()
```

Refill ponds are not stable because of their environment; they are
stable because of intervention. Some refill ponds have deficit values as
high or higher than dryout ponds. Refill ponds occupy a similar and in
some cases more exposed position along the PC1-derived deficit gradient
compared to dryout ponds, indicating that they share underlying
environmental conditions associated with rapid water loss. This suggests
that their current persistence is likely maintained by artificial
refilling rather than inherent resistance to drying. In this context,
refill ponds can be interpreted as systems with latent drying risk,
where environmental drivers of exposure, particularly reduced shading,
are comparable to those observed in dryout ponds. This highlights the
extent to which management intervention may be compensating for
otherwise unfavorable conditions and underscores the importance of
distinguishing between environmentally stable systems and those that are
maintained through external inputs.

Comparing refill vs. dryout directly

```{r}
aggregate(pc1_deficit ~ drought_class, data = model_df, mean)
```

*Refill ponds are, on average, more exposed than dryout ponds.*
Group-level comparison of PC1-based deficit scores revealed that refill
ponds exhibited the highest average deficit (2.84), followed by dryout
ponds (2.65), with stable ponds showing the lowest values (2.39). This
pattern indicates that refill ponds occupy the most exposed positions
along the primary environmental gradient associated with drying rate,
characterized by reduced canopy cover, lower realized shading, and
greater vegetation-driven exposure. Despite this, refill ponds do not
consistently exhibit complete drying, suggesting that their persistence
is maintained through artificial water inputs rather than inherent
environmental buffering. In contrast, stable ponds are associated with
lower deficit values, reflecting greater shading and reduced exposure.
These results highlight that PC1 captures underlying drying pressure,
while observed pond outcomes reflect the interaction between
environmental conditions and management intervention.

*Approach 1: Predict “Expected Drying Pressure”*

```{r}
names(model_df)
```

Use the earlier regression model to generate predicted z_loss_rate for
refill ponds specifically.

```{r}
model <- lm(z_loss_rate ~ PC1, data = model_df)

model_df$predicted_loss <- predict(model, newdata = model_df)
```

Then look at refill ponds:

```{r}
subset(model_df, drought_class == "refill")[, c("pond", "PC1", "pc1_deficit", "predicted_loss")]
```

Rank refill ponds by vulnerability

```{r}
refill_df <- subset(model_df, drought_class == "refill")

refill_df <- refill_df[order(refill_df$predicted_loss), ]

refill_df[, c("pond", "PC1", "pc1_deficit", "predicted_loss")]
```

```{r}
cor(model_df$PC1, model_df$predicted_loss)
```

Interpretation: Refill ponds span the full range of the exposure
gradient, from conditions comparable to stable ponds (BGL) to conditions
exceeding those of dryout ponds (BGS). Some refill ponds exhibit
environmental conditions consistent with rapid drying, suggesting their
persistence is dependent on continued water inputs.

Refill ponds exhibited substantial variation in predicted drying
pressure, spanning a wide range of positions along the PC1 exposure
gradient. Predicted drying values ranged from -0.59 in BGL (PC1 = 1.42;
deficit = 1.21), indicating low exposure and relatively low drying
pressure, to 0.67 in BGS (PC1 = -1.61; deficit = 4.24), representing the
highest exposure and strongest predicted drying pressure among refill
ponds. Intermediate systems such as RUP (predicted = -0.26; deficit =
2.00) and GVI (predicted = 0.17; deficit = 3.03) occupied transitional
positions, while MCC (predicted = 0.46; deficit = 3.74) aligned closely
with high-risk conditions. Notably, the group mean PC1 deficit for
refill ponds (2.84) exceeded that of dryout ponds (2.65) and stable
ponds (2.39), indicating that refill ponds, on average, occupy more
exposed positions along the primary gradient controlling drying rate.
Together, these results demonstrate that refill ponds are not
environmentally uniform, but instead span a continuum from low-risk
systems to highly exposed, intervention-dependent systems, with some
ponds exhibiting predicted drying pressures comparable to or exceeding
those observed in dryout ponds.

*Approach 2: Compare refill ponds to dryout range*

```{r}
subset_df <- subset(model_df, drought_class %in% c("dryout", "refill"))

ggplot(subset_df, aes(x = drought_class, y = pc1_deficit, fill = drought_class)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.1, size = 2) +
  theme_minimal()
```

Get dryout range:

```{r}
dryout_range <- range(model_df$pc1_deficit[model_df$drought_class == "dryout"])
dryout_range
```

Plot

```{r}
ggplot(model_df, aes(x = drought_class, y = pc1_deficit)) +
  geom_jitter(width = 0.1, size = 2) +
  geom_hline(yintercept = dryout_range[1], linetype = "dashed") +
  geom_hline(yintercept = dryout_range[2], linetype = "dashed") +
  theme_minimal()
```

Identify refill ponds overlapping dryout range:

```{r}
refill_df <- subset(model_df, drought_class == "refill")

refill_df$within_dryout_range <- refill_df$pc1_deficit >= dryout_range[1] & 
                                refill_df$pc1_deficit <= dryout_range[2]

refill_df
```

Density plot

```{r}
ggplot(model_df, aes(x = pc1_deficit, fill = drought_class)) +
  geom_density(alpha = 0.4) +
  theme_minimal()
```

Interpretation: This anchors refill ponds directly against the observed
environmental envelope of dryout ponds. Comparison of refill ponds to
the observed dryout range of PC1-based deficit values (1.99 to 3.00)
indicates substantial overlap in environmental conditions associated
with drying. Two refill ponds, RUP (1.996) and GVI (3.031), fall
directly within or at the boundary of the dryout range, while MCC
(3.744) and BGS (4.240) exceed the upper bound, indicating exposure
levels greater than those observed in any dryout pond. Only BGL (1.211)
falls clearly below the dryout range, suggesting comparatively lower
vulnerability. This pattern demonstrates that 4 out of 5 refill ponds
(80%) occupy environmental conditions consistent with or more extreme
than those associated with dryout, despite not exhibiting complete
drying. These results reinforce that refill ponds are not
environmentally distinct from dryout systems, but instead frequently
experience comparable or elevated exposure along the primary gradient
controlling drying rate. The persistence of these ponds therefore likely
reflects the influence of artificial water inputs rather than inherent
resistance to drying, with several refill ponds representing highly
exposed, intervention-dependent systems.

*Approach 3: Rank refill ponds by vulnerability*

```{r}
refill_df <- subset(model_df, drought_class == "refill")

refill_ranked <- refill_df[order(-refill_df$pc1_deficit), 
                           c("pond", "PC1", "pc1_deficit", "predicted_loss")]

refill_ranked
```

Since we already established dryout range = 1.99 to 3.00. Ties refill
ponds to dryout range.

```{r}
dryout_range <- c(1.989573, 3.004020)

refill_ranked$vulnerability <- ifelse(
  refill_ranked$pc1_deficit > dryout_range[2], "Exceeds dryout",
  ifelse(refill_ranked$pc1_deficit >= dryout_range[1], "Within dryout", "Below dryout")
)

refill_ranked
```

Interpretation:

Among refill ponds, there is a clear gradient in latent drying
vulnerability once you remove artificial inputs. Three ponds fall well
into the “would likely dry out” space: BGS (PC1 = –1.61, predicted loss
= 0.67), MCC (–1.12, 0.46), and GVI (–0.40, 0.17). These are not
marginal cases. Their predicted loss values are positive and relatively
high, meaning their environmental conditions align closely with your
observed dry-out ponds. BGS in particular sits at the extreme end, with
the largest PC1 deficit (4.24), indicating a substantial mismatch
between its current state and the stability threshold.

RUP (PC1 = 0.63, predicted loss = –0.26) sits near the boundary. It
falls within the dry-out envelope but does not strongly exceed it,
suggesting it may be conditionally stable and sensitive to moderate
shifts in inputs or climate.

BGL (PC1 = 1.42, predicted loss = –0.59) is the only refill pond clearly
positioned in a stable region. Its positive PC1 score and strongly
negative predicted loss indicate environmental conditions that are
inherently resistant to drying, even without supplementation.

The key point is this: refill ponds are not a homogeneous group. Your
results show a hidden vulnerability structure, where some ponds are
effectively being kept wet by management rather than environmental
resilience. If inputs were removed, at least half of these systems would
likely behave more like dry-out ponds, with BGS and MCC being the most
at risk.

*Interpretation of the 3 methods* Here, they align remarkably well, and
that is what gives your result real weight.

Across all three approaches, refill ponds consistently separate into two
distinct groups: those that are structurally vulnerable to drying and
those that are inherently stable regardless of inputs. The
regression-based predictions show that BGS (predicted loss = 0.67), MCC
(0.46), and GVI (0.17) all fall within or exceed the dry-out range,
indicating that their environmental conditions closely match those of
observed dry-out ponds. The PC1 deficit analysis reinforces this, with
these same ponds showing the largest deviations from the stability
threshold (4.24, 3.74, and 3.03, respectively), confirming that they are
not just statistically similar to dry-out ponds, but meaningfully
displaced along the dominant environmental gradient driving water loss.

The ranking approach makes this pattern unavoidable. When ordered by
vulnerability, the same three ponds consistently occupy the highest-risk
positions, followed by RUP, which sits near the threshold (predicted
loss = –0.26; deficit = 2.00) and appears conditionally stable. Only BGL
(predicted loss = –0.59; deficit = 1.21) remains clearly separated as a
low-risk system, with environmental characteristics that place it firmly
outside the dry-out space.

What matters here is the convergence. These are not three independent
answers producing noise. They are three different lenses on the same
underlying gradient, and they all identify the same structure. That
tells you the signal is strong.

The broader implication is hard to ignore. Refill ponds are not a
coherent ecological category. Instead, they mask a continuum of drying
vulnerability, where some ponds are only maintained through artificial
inputs while others are fundamentally resilient. Your analysis shows
that if those inputs were removed, a substantial portion of refill ponds
would likely transition into the dry-out class, with BGS and MCC
representing the most extreme cases of hidden vulnerability.

######### 

ALL PONDS RANKING

######## 

```{r}
# Rank all ponds by predicted drying risk
ranked_df <- model_df[order(-model_df$predicted_loss), ]

# Add a rank column
ranked_df$dryout_rank <- 1:nrow(ranked_df)

# View results
ranked_df[, c("dryout_rank", "pond", "drought_class", "PC1", "predicted_loss")]
```

Add risk categories

```{r}
ranked_df$risk_category <- cut(
  ranked_df$predicted_loss,
  breaks = c(-Inf, -0.3, 0, 0.3, Inf),
  labels = c("Low", "Moderate", "High", "Extreme")
)

ranked_df[, c("dryout_rank", "pond", "drought_class", 
              "PC1", "pc1_deficit", "predicted_loss", "risk_category")]
```

Importantly: This is a relative risk ranking based on environmental
similarity, not a deterministic forecast.

*Recap: How Risk Was Quantified*

Dryout risk was quantified using the primary environmental gradient
identified through PCA (PC1), which represents a continuum from high
shading and lower exposure to low shading and increased exposure.
Because regression analysis showed that PC1 significantly predicts
drying rate, this axis was used as the foundation for all risk
assessments.

Risk was operationalized in two linked ways. First, PC1 deficit
quantified how far each pond deviates from the most stable conditions
along this gradient, with higher values indicating greater exposure and
reduced shading. Second, predicted drying pressure was calculated using
the regression relationship between PC1 and standardized water loss rate
(z_loss_rate), allowing each pond to be assigned an expected drying
intensity based on its environmental position.

Dryout ponds provide a critical reference for interpreting this
gradient. Their PC1 deficit values define the observed environmental
range associated with actual drying, allowing other ponds to be
evaluated relative to known outcomes. Ponds that fall within or exceed
this range can therefore be interpreted as experiencing comparable or
greater underlying drying pressure, regardless of their current
classification.

*Ordered Ranking of Ponds by Dryout Risk*

From highest to lowest vulnerability:

Extreme Risk VAN (stable) – highest deficit (5.86), highest predicted
loss (1.34) BGS (refill) – deficit 4.24, predicted 0.67 MCC (refill) –
deficit 3.74, predicted 0.46

High Risk GVI (refill) – deficit 3.03, predicted 0.17 MSP (dryout) –
deficit 3.00, predicted 0.16 QED (dryout) – deficit 2.95, predicted 0.13
CSN (dryout) – deficit 2.90, predicted 0.11

Moderate Risk TRT (stable) – deficit 2.57 JCO (dryout) – deficit 2.42
RUP (refill) – deficit 2.00 BOO (dryout) – deficit 1.99 CHS (stable) –
deficit 1.98

Low Risk SCT (stable) – deficit 1.53 BGL (refill) – deficit 1.21 AVL
(stable) – lowest deficit (0.00), lowest predicted loss (-1.09)

*Final Interpretation*

The ranking reveals that dryout risk is not confined to ponds that
currently dry, but instead spans all drought classes along a continuous
environmental gradient. Notably, the highest-risk pond (VAN) is
classified as stable, while two refill ponds (BGS and MCC) rank above
all observed dryout systems. This indicates that current pond states do
not always reflect underlying environmental vulnerability, and that
management interventions or short-term conditions may mask latent drying
risk.

Dryout ponds cluster within the high-risk portion of the gradient,
reinforcing that PC1 captures the dominant environmental conditions
associated with observed drying. However, the presence of refill and
stable ponds within this same range demonstrates that similar
environmental conditions can lead to different outcomes depending on
external inputs or system-specific characteristics.

Overall, these results confirm that pond drying operates along a shared
exposure-driven gradient, where shading and energy balance control
underlying drying pressure, while factors such as water inputs and
system capacity influence whether that pressure results in complete
drying. This highlights the importance of distinguishing between
observed state and environmental vulnerability, as ponds that appear
stable may still occupy high-risk positions and require proactive
management to prevent future drying under changing conditions.

################## 

APPENDIX

################# 

*1. Removal of thermal response variables*

Summary: july_water_air was excluded because it behaved as a downstream
thermal response rather than an independent environmental driver. It
reflects the outcome of shading, exposure, and pond structure, so
including it in the PCA would risk collapsing the ordination into a
generalized hot versus cool gradient rather than revealing the
underlying mechanisms structuring ponds.

-   daily_loss_est
-   total_depth_loss_cm
-   total_volume_lost_m3
-   mean_daily_volume_lost_m3
-   z_loss_rate
-   z_total_loss

*2. Selection of SEI over other shade metrics*

Summary: alternative shade metrics such as MSD, SCP, and MSF were
tested, but these were strongly correlated with temperature-related
variables and behaved more like composite or response-like indices.
july_sei was retained because it showed more moderate relationships and
captured the spatial distribution of shade across pond surfaces rather
than simply overall shade magnitude or persistence. This made it more
interpretable and less redundant within the PCA.

```{r, eval=FALSE}
shade_compare <- metrics_wide_2 %>%
  dplyr::select(
    canopy_50,
    july_water_air,
    av_hrs_water_july_mintomax,
    july_msd,
    july_scp,
    july_msf,
    july_sei
  )

round(cor(shade_compare, use = "pairwise.complete.obs"), 2)
```

*3. Vegetation variable refinement*

Summary: veg_percent_change was initially considered, but it showed
instability and was less suitable for PCA because extreme values could
disproportionately influence ordination structure. It was replaced with
veg_mean_cover_july, which better represented vegetation state at peak
summer conditions and provided a more stable, interpretable measure of
biological structure.

```{r, eval=FALSE}
veg_compare <- metrics_wide_2 %>%
  dplyr::select(
    veg_percent_change,
    veg_mean_cover_july,
    slope_per_day
  )

summary(veg_compare)
```

*4. Transformation of growth dynamics*

Summary: slope_per_day was retained as a measure of vegetation change
through time, but it showed a right-skewed distribution. To reduce
leverage from a few high values, it was transformed using log1p(),
producing slope_per_day_log. This allowed the variable to remain in the
PCA without overly dominating component structure.

```{r, eval=FALSE}
metrics_wide_2 <- metrics_wide_2 %>%
  dplyr::mutate(
    slope_per_day_log = log1p(slope_per_day)
  )

summary(metrics_wide_2$slope_per_day_log)
boxplot(metrics_wide_2$slope_per_day_log)
```

*5. Final conceptual grouping of candidate variables*

Summary: The final exploratory PCA candidate set was organized to
represent distinct ecological dimensions rather than a cluster of
statistically convenient variables. This grouping helped ensure that the
PCA captured multiple complementary gradients in pond structure and
function.

Landscape structure: canopy_50, imp_100 Thermal mediation: july_sei,
av_hrs_water_july_mintomax Biological and chemical state:
veg_mean_cover_july, slope_per_day_log, mean_season_ph

This step is important because it shows that variable selection was
driven by ecological interpretation as well as statistical screening.

```{r, eval=FALSE}
pca_df <- metrics_wide_2 %>%
  dplyr::select(
    canopy_50,
    imp_100,
    av_hrs_water_july_mintomax,
    july_sei,
    veg_mean_cover_july,
    slope_per_day_log,
    mean_season_ph
  )

names(pca_df)
summary(pca_df)
```

Together, these steps show that variable selection was not based on
simple trial and error, but on an iterative process balancing
redundancy, ecological meaning, and interpretability. Downstream thermal
response variables were removed to reduce circularity, SEI was selected
as the most defensible realized shade metric, unstable vegetation change
metrics were replaced with more robust summer state variables, and
skewed growth measures were transformed before ordination. The resulting
variable set was intentionally structured to represent landscape,
thermal, and biological dimensions of pond systems rather than a single
dominant process.

*6. Motivation: Estimating Pond Storage Capacity*

Summary: A derived variable was explored to approximate pond storage
capacity, with the goal of capturing a pond’s ability to buffer against
seasonal water loss. This was motivated by the idea that drying dynamics
are not solely driven by environmental exposure, but also by inherent
physical capacity. This is a proxy, not a direct measurement of volume
or bathymetry, and assumes depth_1 is representative of basin structure.

Key idea: - Dry-out is a function of both inputs/outputs and storage -
PCA variables so far mostly represent flux drivers, not capacity

Code (representative, not exhaustive)

```{r, eval=FALSE}
metrics_wide_2 <- metrics_wide_2 %>%
  mutate(storage_proxy = depth_1 * area)
```

*7. Evaluation and Rejection of Storage Proxy*

Summary: The storage proxy was evaluated but ultimately excluded from
PCA due to conceptual and statistical limitations.

Reasons: - Strongly correlated with pond size → risks dominating PCA as
a “size axis” - Does not capture dynamic hydrology (inflow, seepage,
evapotranspiration) - Based on a single depth measure, which may not
represent basin morphology

*8. Evaluation and Exclusion of Impervious Surface Metrics*

Impervious surface cover was initially considered as a proxy for
surrounding urban intensity, with the expectation that higher
imperviousness would amplify thermal exposure, alter runoff dynamics,
and contribute to increased drying rates in urban ponds.

To evaluate its contribution, impervious cover was tested at two spatial
scales:

imp_50 (local influence) imp_100 (broader landscape context)

Two parallel PCA configurations were constructed to assess whether
impervious surface variables contributed meaningful explanatory
structure.

Code (representative)

```{r, eval=FALSE}
# Define variable sets
vars_no_imp <- c(
  "canopy_50",
  "av_hrs_air_july_mintomax",
  "july_msf",
  "veg_percent_change",
  "depth_1"
)

vars_with_imp <- c(vars_no_imp, "imp_100")

# Run PCA
pca_no_imp  <- prcomp(metrics_wide_2[vars_no_imp], scale. = TRUE)
pca_with_imp <- prcomp(metrics_wide_2[vars_with_imp], scale. = TRUE)

# Compare structure
summary(pca_no_imp)
summary(pca_with_imp)

round(pca_no_imp$rotation, 2)
round(pca_with_imp$rotation, 2)

# Compare scores
scores_no_imp  <- as.data.frame(pca_no_imp$x)
scores_with_imp <- as.data.frame(pca_with_imp$x)

cor(scores_no_imp$PC1, scores_with_imp$PC1)
cor(scores_no_imp$PC2, scores_with_imp$PC2)
```

*Final Synthesis*

All exploratory testing and iterative refinement converged on a final
set of variables that balance ecological interpretability with
statistical independence. Early stages of variable selection revealed
that many candidate predictors were either redundant, overly correlated,
or reflective of downstream responses rather than independent drivers.
Through successive removal of thermal outcome variables, consolidation
of shade metrics, and refinement of vegetation and growth indicators,
the analysis shifted toward representing distinct ecological mechanisms
rather than overlapping signals. Additional exploratory configurations,
including the incorporation of storage proxies and alternative thermal
and vegetation variables, further highlighted the importance of
separating structural capacity, environmental forcing, and biological
response.

The final selection of canopy_50, av_hrs_air_july_mintomax, july_msf,
veg_percent_change, and depth_1 reflects this process, with each
variable contributing uniquely to the representation of landscape
structure, thermal dynamics, realized shading, seasonal biological
change, and physical storage capacity. By emphasizing July-based
metrics, the model isolates peak environmental forcing while avoiding
late-season feedback effects, ensuring that the PCA captures underlying
drivers of drying rather than cumulative outcomes. Together, this
variable set provides a parsimonious yet mechanistically grounded
framework for interpreting variation in pond drying dynamics across
sites, with minimal redundancy and clear ecological meaning.
